### 2025-05-05 Chapters/Chapter2.tex
first commit

\begin{enumerate}
    \item[\textbf{[1]}] The proposed architecture, Efficient Sleep Sequence Network (ESSN), has overcome the limitations of existing automatic sleep stage algorithms. This model addresses two main challenges. First, the model is quite complex, and often low-end systems are unable to process it; therefore, this model is designed to work on lightweight systems. The second challenge is the misclassification of the N1 stage, where models often confuse wake and REM stages. To address this, it introduces the N1 structure loss function. The ESSN model has achieved impressive metrics: 88.0\% accuracy, 81.2\% macro F1, and 0.831 Cohen’s kappa. These results were obtained on the SHHS dataset. Additionally, it has reduced computational requirements, with only 0.27M parameters and 0.35G floating-point operations, and it claims to be faster than models like L-SeqSleepNet.

    \item[\textbf{[2]}] The Multi-Domain View Self-Supervised Learning Framework (MV-TTFC) introduces a new approach to classify sleep stages by leveraging self-supervised learning (SSL) on unlabeled EEG data. By incorporating multi-view representation technology, this model enhances information exchange across different views. It also introduces the multisynchrosqueezing transform, which improves the quality of the time-frequency view. Ultimately, it captures the latent features within EEG signals. It was evaluated on two datasets (SleepEDF-78 and SHHS), and MV-TTFC achieved state-of-the-art performance with accuracies of 78.64\% and 81.45\%, and macro F1-scores of 70.39\% and 70.47\%, respectively.

    \item[\textbf{[3]}] The proposed CNN-Transformer-ConvLSTM-CRF hybrid model presents a new integration method between local and global feature extraction to enhance the classification ability of sleep stages. The model can identify relationships among EEG features by applying a multi-scale convolutional neural network combined with a Transformer for encoding features of the EEG signal and a spatio-temporal encoder via ConvLSTM. Additionally, the adaptive feature calibration module improves the extracted features, and there is efficient learning of the transition relationships between the stages of sleep by the CRF module. Based on evaluations on three datasets, this hybrid model outperforms existing state-of-the-art methods, demonstrating its efficacy in sleep stage classification.
\end{enumerate}

### 2025-05-05 Chapters/SAMDC.tex
first commit

\section{Data Pre-Processing}
In the preliminary stage of preparing MRI images for the segmentation of brain tumors, the first action involves resizing the images to a uniform dimension of (128, 128, 128, 1). Subsequently, a sequence of filtering techniques is employed to enhance the clarity and detail of these resized MRI images. These filtering methods are classified into two primary domains: frequency domain filters and spatial domain filters. Frequency domain filters comprise the Homomorphic Filter, Gabor Filter, and Gaussian High Pass Filter, each tailored to manipulate the frequency characteristics of the images. On the other hand, spatial domain filters encompass techniques like Sharpen Filters and Median Filters, which operate directly on the spatial distribution of pixel intensities within the images. This multi-step preprocessing workflow aims to optimize the quality and suitability of the MRI data for subsequent tumor segmentation tasks.

\begin{table}[h]
  \centering
  \small
  \begin{tabularx}{\linewidth}{|X|X|X|X|X|}
    \hline
    &\textbf{Filter Name} & \textbf{Pratts Merit} & \textbf{Entropy} & \textbf{SSI} \\
    \hline
    \hline
    \textbf{Frequency Domain }&Homomor-phic Filter & 8.692 & 18.000 & 0.145\\
    \cline{2-5}
    \textbf{}&Gabor Filter & 9.041 & 16.469 & 0.145\\
    \cline{2-5}
    \textbf{}&Gaussian High Pass Filter & 11.091 & 18.000 & 0.301 \\
    \hline
    \hline
    \textbf{Spatial Domain Filters}&Sharpen Filter (Kernal Size 3) & 19.512 & 12.043 & 0.814\\
    \cline{2-5}
    \textbf{}&Sharpen Filter (Kernal Size 5) & 1.099 & 13.208 & 0.072\\
    \cline{2-5}
    &Median Filter (Kernal Size 5) & 28.683 & 11.324 & 0.873\\
    \cline{2-5}
    \textbf{}&Median Filter (Kernal Size 3) & \textbf{32.838} & \textbf{11.478} & \textbf{0.955}\\
    \hline
  \end{tabularx}
  \caption{Filter Efficacy: A Performance Insight into Preprocessed images using Quality Metrics e.g. Pratts Merit, Entropy and Similarity stucture index(SSI).}
  \label{tab:Filter Efficacy}
\end{table}

The table \ref{tab:Filter Efficacy} offers a thorough assessment of diverse image preprocessing filters, evaluating their efficacy across three fundamental parameters: Pratt's Merit, Entropy, and SSI (Structural Similarity Index). Specifically, the Homomorphic Filter demonstrates a Pratt's Merit value of 8.692, an Entropy of 18, and an SSI of 0.145. Similarly, the Gabor Filter closely follows with scores of 9.041, 16.469, and 0.145 for Pratt's Merit, Entropy, and SSI respectively. Additionally, the Gaussian High Pass Filter yields results of 11.091 for Pratt's Merit, 18.000 for Entropy, and 0.301 for SSI. In the spatial domain, the analysis encompasses Sharpen Filters with kernel sizes of 3 and 5, along with Median Filters of kernel sizes 3 and 5. The Sharpen Filter with a kernel size of 3 exhibits a Pratt's Merit of 19.512, Entropy of 12.043, and SSI of 0.814. Conversely, the Sharpen Filter with a kernel size of 5 performs relatively lower, recording values of 1.099, 13.208, and 0.072 for Pratt's Merit, Entropy, and SSI respectively. The Median Filter with a kernel size of 3 emerges as a standout performer, showcasing a noteworthy Pratt's Merit of 32.838, minimal Entropy of 11.478, and high SSI of 0.955. Similarly, the Median Filter with a kernel size of 5 delivers competitive outcomes with scores of 28.683, 11.324, and 0.873 for Pratt's Merit, Entropy, and SSI. Based on this analysis, the adoption of spatial filters, such as the Median Filter with a kernel size of 3, is preferred due to their resilience against noise. This attribute holds significant importance in the realm of processing medical images such as MRI scans, where noise mitigation is crucial for precise diagnosis and analysis. This efficacy is visually apparent in Figure \ref{Image: Preprocess compare filters}, where the employment of spatial filters noticeably enhances image quality by mitigating noise and enhancing image contrast.

\begin{figure*}[!h]
  \centering
   \includegraphics[width=0.9\linewidth]{Images_4th/compare_filters.pdf}
    \caption{Comparison of Various Image Preprocessing Filters: Illustration depicting the comparative effectiveness of diverse preprocessing filters, encompassing both spatial and frequency-based variants.}
  \label{Image: Preprocess compare filters}
\end{figure*}

\section{Class-Enhanced Mahalanobis Distance Contrastive (CEMDC) Loss}\label{sec:samdcapproach}
Our method strategically addresses the issue of imbalanced class distribution within datasets by employing a multifaceted approach. In addition to utilizing feature-based representations to enhance segmentation accuracy, we focus on effectively managing the inherent imbalance in class distribution through the incorporation of the Tversky loss (\(L_{Tversky}\)). This loss function proves particularly useful in scenarios where class sizes exhibit significant disparities. Moreover, we enhance the quality of feature representations by integrating the Mahalanobis distance-contrastive loss (\(L_{MDC}\)). By leveraging the Mahalanobis distance, we extract valuable contrastive information from the spread distribution of the data, thereby facilitating more discriminative and efficient segmentation outcomes. To strike a balance between mitigating class imbalance (as indicated by \(L_{Tversky}\)) and improving discriminative feature learning (as indicated by \(L_{MDC}\)), we introduce a hyperparameter \(\Delta\). This parameter dynamically adjusts inversely with respect to the Tversky loss value, ensuring adaptability and refined optimization throughout the process. Specifically, as the Tversky loss value increases, the values of \(\Delta\) decrease, and vice versa. This adaptable nature facilitates the formulation of our Class-Enhanced Mahalanobis Distance Contrast Loss (\(L_{CEMDC}\)), as expressed in Equation \ref{EQ: L_CEMDC}. Experimental validation conducted on diverse brain tumor datasets (refer to Table \ref{tab:CEMDC_Results}) demonstrates the significant improvement in generalization achieved through our integrated approach. The equilibrium achieved through this integration is succinctly captured in Equation \ref{EQ: L_CEMDC}.

\begin{equation} \label{EQ: L_CEMDC}
L_{\text{CEMDC}} = \Delta \times L_{\text{Tversky}} +  L_{\text{MDC}}
\end{equation}

\subsection{Tversky Loss}
The Tversky loss, as delineated by Equation \ref{EQ: LTversky}, emerges as a fundamental cornerstone guiding the navigation through the intricacies of addressing imbalanced classes within segmentation tasks. Its essence lies in furnishing a comprehensive measure of the discrepancy between the ground truth annotations and the predictions generated by our segmentation model. This specialized loss function is meticulously tailored to confront and mitigate the inherent complexities stemming from uneven distributions across different classes.

\begin{equation}
\label{EQ: LTversky}
L_{\text{Tversky}} = 1 - \frac{{TP + \epsilon}}{{TP + \alpha \times FP + \beta \times FN + \epsilon}}
\end{equation}

At its core, the Tversky loss embodies a strategic framework that endeavors to uphold equity in the treatment of true positives while delicately navigating the intricate balance between false positives and false negatives. To facilitate smoother optimization and convergence during training, the inclusion of the parameter \(\epsilon\) serves to introduce a smoothing mechanism, thus alleviating potential numerical instabilities. The variables \(TP\), \(FP\), and \(FN\) encapsulate the instances of true positives, false positives, and false negatives, respectively, offering a comprehensive lens through which to gauge the performance of our segmentation model.
The dynamic interplay orchestrated by the adjustable parameters \(\alpha\) and \(\beta\) provides a nuanced means to fine-tune the model's sensitivity to false positives and negatives, thereby enhancing its adaptability across diverse datasets and segmentation scenarios. This meticulous calibration and steadfast commitment to harmonizing class imbalances underscore the robustness and reliability of our segmentation strategy, ensuring not only heightened precision but also steadfastness in the outcomes generated by our model.

\subsection{Mahalanobis Distance-Contrastive (MDC) Loss}

The Mahalanobis Distance-Contrastive (MDC) Loss technique is proposed with the aim of refining segmentation accuracy and enhancing the quality of feature representations. This methodology incorporates the Mahalanobis distance, a metric used to measure the dissimilarity between a specific point and a given distribution, in order to capture the inherent structure of the feature representation space. Additionally, a contrastive loss component is integrated, which serves to emphasize the learning of features by accentuating the differences between various classes. This strategic approach enables the model to effectively utilize the spread distribution of data, thereby resulting in more distinct feature representations and precise segmentation outcomes.

Let's define \(x_c\) and \(x_p\) as multidimensional feature vectors representing the current and previous sample batch histories, respectively. This allows us to analyze sequential batch representations thoroughly, ensuring consistency in how features are portrayed and improving generalization capabilities across datasets. To make these vectors comparable, they are scaled along each dimension by their L2 norms, ensuring they are adjusted to a unit length. This normalization process effectively places the vectors on a unit hypersphere, enabling us to make directional relationship comparisons more easily. This normalization is crucial for calculating the Mahalanobis distance-based representation (\(D_M\)) (see Equation \ref{EQ: MD}) for both the previous and current sample batch feature vectors.

The Mahalanobis distance serves as a measure of dissimilarity between a given sample batch and a distribution, taking into account both the mean (\(\mu\)) and covariance (\(\Sigma\)) of the distribution. By incorporating the inverse of the covariance matrix (\(\Sigma^{-1}\)), the distance metric is normalized to account for the spread and orientation of the feature distribution. This adjustment compensates for variations in scales and correlations among features, ensuring that each dimension contributes uniformly to the overall distance computation. Essentially, utilizing the inverse covariance matrix enhances the discriminative capability of the Mahalanobis distance metric by accommodating the distributional characteristics of the feature space.

\begin{equation} \label{EQ: MD}
D_M(x) = (x - \mu)^T \Sigma^{-1} (x - \mu) 
\end{equation}

To thoroughly examine differences within the feature representation space, we analyze both the typical characteristics and the interrelations among features, which are described respectively by the mean and covariance. This comprehensive assessment allows for a nuanced understanding of distinctions within the specific feature context, highlighting the Mahalanobis distance matrix as a crucial tool for identifying subtle variations in feature distributions. In the calculation of the contrastive loss, we employ the Mahalanobis Distance (\(D_M\)), where \(x\) denotes feature vectors for both the previous (\(x_p\)) and current (\(x_c\)) batches. The Mahalanobis Distance-Contrastive (MDC) Loss, as depicted in Equation \ref{EQ:MDC}, can be articulated as follows:

\begin{equation} \label{EQ:MDC}
\begin{aligned}
L_{\text{MDC}} &= -\frac{1}{N} \sum_{i=1}^{N} \log\left(\exp\left(-\frac{D_M(x_p)}{\tau}\right) + \exp\left(-\frac{D_M(x_c)}{\tau}\right)\right) \\
&\quad \times \text{lr}
\end{aligned}
\end{equation}

\noindent Here, \(N\) represents the batch size, while \(D_M(x_p)\) and \(D_M(x_c)\) refer to the Mahalanobis distances for the previous and current batches, respectively. The temperature hyperparameter \(\tau\) regulates the steepness of the exponential function, thus affecting the computation of the contrastive loss. By considering both past and present batch representations, this loss function encourages the model to learn more discerning features while maintaining consistency and stability across batches. Additionally, the learning rate scaling factor \(\text{lr}\) adjusts the overall magnitude of the loss to effectively guide the optimization process. This formulation effectively leverages the discriminative power of the Mahalanobis Distance within the domain of contrastive learning, with the objective of refining the quality of feature representations and enhancing the model's learning process.

\section{Model Intergration with CEMDC Loss}
In the context of our model integration, we have opted for the utilization of a cutting-edge 3D U-Net architecture, which is widely acknowledged for its efficacy in handling tasks related to the segmentation of medical images. This particular architectural design is particularly adept at capturing spatial relationships and complex structures inherent in volumetric data, thus making it an optimal choice for our segmentation objectives. Moreover, to ensure a comprehensive understanding of our integrated methodology, we have devised a meticulously crafted flowchart that elucidates the architectural framework, as illustrated in Figure \ref{Image: CEMDC Architecture}.

\begin{figure}[!h]
  \centering
   \includegraphics[width=0.95\linewidth]{Images_4th/CEMDC_LOSS_Architecture.pdf}
    \caption{A visual representation outlining the structure of our combined approach, which integrates the Class-Enhanced Mahalanobis Distance Contrast (CEMDC) model with the 3D-UNet model.}
  \label{Image: CEMDC Architecture}
\end{figure}

This flowchart serves to delineate the sequential progression of steps entailed within our methodology, thereby facilitating a deeper comprehension of the inner workings of the model. In addition to the visual depiction, we have also furnished a detailed algorithm encapsulating the entire architectural process, as delineated in Algorithm \ref{Algo: CEMDFrame}. This algorithm meticulously outlines the procedural steps involved in our approach, ranging from model training to inference generation. By furnishing both a visual representation and a comprehensive algorithmic description, our aim is to offer a holistic perspective of our model integration process, thereby enabling researchers and practitioners to grasp the intricacies of the methodology and effectively replicate the results.

\begin{algorithm}[h]
\caption{CEMDC Loss Calculation During Model Training}
\label{Algo: CEMDFrame}
\begin{algorithmic}[1]
\State \textbf{Input:} Number of epochs $E$, temperature $\tau$, learning rate $lr$, hyper-parameter $\Delta$, smoothness $\epsilon$ 
\For{epoch $i = 1, 2, \dots, E$}
    \For{each batch $b = \{x, y\}$ of $\mathcal{D}$}
        \State $L_{Tversky} \leftarrow TverskyLoss(y,\hat{y})$
        \State $\Delta \leftarrow \frac{1}{L_{Tversky}}$
        \State $x_c \leftarrow R_{w^t}(x)$
        \State $x_p \leftarrow R_{w^{t}}(x-1)$
        \State $D_M(x_c) \leftarrow (x_c - \mu)^T \Sigma^{-1} (x_c - \mu) $
        \State $D_M(x_p) \leftarrow (x_p - \mu)^T \Sigma^{-1} (x_p - \mu) $
        \State $X_p \leftarrow \exp\left(-\frac{D_M(x_p)}{\tau}\right) $
        \State $X_c \leftarrow \exp\left(-\frac{D_M(x_c)}{\tau}\right)$
        \State $L_{\text{MDC}} \leftarrow -\frac{1}{N} \sum_{i=1}^{N} \log\left( X_p + X_c\right) \times \text{lr}$
        \State $L_{\text{CEMDC}} \leftarrow \Delta \times L_{\text{Tversky}} + L_{\text{MDC}}$

        \State$w^t \leftarrow w^t - \eta \nabla L_{CEMDC}$
    \EndFor
\EndFor
\State \Return $w^t$
\end{algorithmic}
\end{algorithm}

### 2025-05-05 Chapters/SFAL.tex
first commit

We introduce a novel federated framework, as depicted in Figure \ref{Fig: fe4dFramework}, which incorporates a straightforward yet potent strategy utilizing an adaptive scale-aware weighted aggregation (ASWA) technique. This framework not only integrates local client training but also implements our suggested spatial-feature alignment loss (SFAL). Particularly in scenarios involving non-IID brain tumor data, a discernible shift occurs during local training, resulting in the global model consistently outperforming the local model in representation. Our primary aim revolves around minimizing the disparity between the representation acquired by the local model and that attained by the global model. Concurrently, we aim to maximize the difference between the representation learned by the local model and the one obtained by its previous iteration, drawing inspiration from contrastive learning, a widely utilized technique in visual representation development. The entire workflow of our proposed method is visually illustrated in Figure \ref{Fig: fe4dFramework}. In subsequent sections, we will delve deeper into the intricacies of the network architecture, provide detailed explanations of the local learning objective, and offer insights into the enhanced aggregation process.

\begin{figure}[!h]
  \centering
   \includegraphics[width=\linewidth]{Images_4th/ColorFlowchat.pdf}
    \caption{Overview of the new federated framework, presenting two strategies: (i) SFAL (Modified Dice + Enhanced Contrastive), aiming to achieve a balance between spatial precision and feature similarity in images. This involves combining modified dice loss and our improved contrastive loss at the model level during local training of individual clients. And (ii) ASWA, ensuring a balanced influence in the aggregation process at the global model from the scales of local client data.}
  \label{Fig: fe4dFramework}
\end{figure}

\section{Architectural Details}
Our segmentation strategy, depicted visually in Figure \ref{Fig: fe4dFramework}, employs the 3D U-Net architecture \cite{cai2022novel,11} as our preferred model for segmentation. The efficacy of the 3D U-Net stems from its adeptness at leveraging 3D contextual cues, facilitating the segmentation of volumetric data with notable precision. To encapsulate the essence of the inputs, we extract representation vectors from the final convolution layer of the 3D U-Net encoder. For the sake of clarity, we designate the model weight as $w$, and the mapped representation vector of input $x$ as $U_w$. Local loss function consists of two fundamental elements. The first component, our adapted dice loss ($L_{Dice}$), serves as a foundational aspect for our local training. The second component, our proposed enhanced contrastive loss function ($L_{EC}$), is formulated to enhance the discriminative capability of our segmentation model.

In the context of client engagement in the local training process, let's consider a specific client $c_i$. Initially, the client obtains the global model ($w^t$) from the server. Subsequently, during the local training phase, client $c_i$ updates its model to $w^t_i$. For each input $x$, we derive its representation from three sources: the global model $w^t$ (represented as $z_g = U_{w^t}(x)$), the local model from the previous round $w^{(t-1)}_i$ (represented as $z_p = U_{w^{(t-1)}_i}(x)$), and the current local model being updated $w^{t}_i$ (represented as $z_c = U_{w^{t}_i}(x)$). Given the comprehensive representation power of the global model across all data, our objective is to minimize the distance between $z_c$ and $z_g$, while simultaneously maximizing the distance between $z_c$ and $z_p$. This approach is designed to align the local model's representation with the global model, thereby capturing the overarching contextual information. At the same time, it aims to differentiate the local model's representation from that of the previous round to ensure progressive learning throughout the federated training process.

\section{Local Spatial-Feature Alignment Loss (SFAL) Function}
In the methodology we propose, we introduce a specialized loss function referred to as the spatial-feature alignment loss (SFAL). This distinctive loss function is designed to merge two crucial elements: the modified dice loss and an upgraded contrastive loss. The primary objective is to comprehensively encapsulate both spatial and feature-related information, thereby ensuring a robust segmentation process. The incorporation of the modified dice loss plays a pivotal role in achieving precise spatial localization, emphasizing the significance of accurate segmentation, which contributes to the overall spatial coherence of the segmented regions. Simultaneously, the inclusion of our enhanced contrastive loss aims to enhance the quality of feature representation, making the segmentation process more discriminative and facilitating a deeper understanding of the underlying features. To offer a mechanism for flexible adjustment, we introduce a hyperparameter denoted as \(\omega\) within the SFAL. This hyperparameter functions as a fine-tuning tool, enabling a balanced trade-off between prioritizing precise segmentation guided by the modified dice loss and enhancing discriminative feature learning influenced by the enhanced contrastive loss. This adaptability allows for a refined optimization process, where our local spatial-feature alignment loss function (\(L_{SFAL}\)) optimally harnesses the strengths of both loss components. The significance of this adaptability is underscored by the improved generalization observed in the outcomes of our experiments. The equilibrium achieved through the combination of these components is precisely articulated in Equation \ref{Eq: SFAL}, providing a mathematical representation of the balanced integration of spatial and feature considerations within our segmentation framework. This refined optimization process highlights the effectiveness of our SFAL function in capturing the intricacies of spatial and feature relationships, ultimately leading to superior segmentation outcomes.

\begin{equation} \label{Eq: SFAL}
    L_{SFAL} = \omega \times L_{Dice} + (1 - \omega) \times L_{EC}
\end{equation}

Here, the parameter $\omega$ serves as a hyperparameter governing the contribution of the dice loss to the SFAL loss, while the term \((1 - \omega)\) regulates the influence of the enhanced contrastive loss. It is essential to note that the sum of $\omega$ and \((1-\omega)\) equals 1.

\subsection{Modified Dice Loss}
The modified dice loss function, which has been tailored to suit our specific requirements, serves as a metric for assessing the degree of overlap between two binary vectors on a pixel-by-pixel basis. This assessment enables us to discern the disparities between the binary masks generated by our model and the ground truth across all dimensions of an image. Mathematically, the modified dice loss can be denoted as follows:

\begin{equation} \label{Eq: Dice Loss}
\text{Loss}_{\text{Dice}} = 1 - \frac{2 \sum_{i,j,k} y_{ijk} \cdot \hat{y}_{ijk} + \epsilon}{\sum_{i,j,k} y_{ijk} + \sum_{i,j,k} \hat{y}_{ijk} + \epsilon}
\end{equation}

Here, \(y_{ijk}\) represents the binary mask derived from the ground truth, while \(\hat{y}_{ijk}\) indicates the corresponding prediction generated by our model. The numerator of this expression computes the double intersection of the masks, incorporating a smoothing term \(\epsilon\) to ensure stability. The denominator calculates the sum of the individual areas for both the ground truth and predicted masks, again integrating the smoothing term \(\epsilon\) to prevent division by zero. This formulation guarantees that the modified dice loss effectively captures the distinctions between the predicted and actual binary masks, thereby serving as a robust evaluation metric for our segmentation model.

\subsection{Enhanced Contrastive Loss Function}

The enhanced contrastive loss function has been meticulously crafted with the specific objective of reducing the disparity between the representation acquired through the local model and that attained from the global model. In addition, it concurrently strives to widen the discrepancy between the representation learned by the local model and its previous iteration. To delve deeper into the complexities of this loss function, it is imperative to examine the multidimensional representation vectors originating from various sources, namely the global model (\(z_g\)), the current local model (\(z_c\)), and the preceding local model (\(z_p\)). These vectors are characterized by five distinct indices: \(n, i, j, k, l\), each representing different dimensions within the embedding space. Specifically, \(n\) denotes the batch size, while \(i, j, k\) delineate the spatial dimensions of the image, and \(l\) signifies the feature index. To ensure uniformity and facilitate unit length adjustment, a normalization procedure is employed on these vectors, whereby they are scaled by the reciprocal of their L2 norm across spatial dimensions. This normalization process is mathematically articulated through the following equations:

\begin{equation}
    z_{g_{ijk}} = \frac{z_{g_{ijk}}}{\sqrt{\sum_{i,j,k} (z_{g_{ijk}})^2}},
\end{equation}

\begin{equation}
    z_{c_{ijk}} = \frac{z_{c_{ijk}}}{\sqrt{\sum_{i,j,k} (z_{c_{ijk}})^2}},
\end{equation}

\begin{equation}
    z_{p_{ijk}} = \frac{z_{p_{ijk}}}{\sqrt{\sum_{i,j,k} (z_{p_{ijk}})^2}}.
\end{equation}

The process of normalization serves the crucial purpose of ensuring that the vectors are positioned precisely on the unit hypersphere, thereby simplifying the examination of directional relationships, which holds paramount importance in the realm of similarity calculations. Through this normalization procedure, a uniform scale is established, thereby providing a consistent foundation for gauging similarity. In our pursuit of attaining a thorough comprehension of contextual comparisons, we undertake a meticulous examination of the relationships inherent in feature representations. Our initial focus revolves around assessing the degree of similarity between the representation of the current local model (\(z_c\)) and that of the server model (\(z_g\)), a relationship articulated by:

\begin{equation} \label{eq: sim_zc_zg}
     \text{sim}(z_c, z_g) = \sum_{i,j,k} z_{c_{ijk}} \times z_{g_{ijk}}.
\end{equation}

This comparative analysis furnishes invaluable insights into the extent to which the feature vectors of the current local model correspond with those emanating from the global server model. The normalization to the unit hypersphere guarantees that these representations are standardized, thereby facilitating a meaningful assessment of their alignment. Furthermore, the evaluation of similarity between the representation of the current local model (\(z_c\)) and that of the preceding local model (\(z_p\)) is expressed by:

\begin{equation} \label{eq: sim_zc_zp}
     \text{sim}(z_c, z_p) = \sum_{i,j,k} z_{c_{ijk}} \times z_{p_{ijk}}.
\end{equation}

This examination offers detailed insights into the evolution and consistency of the feature vectors within the local model across successive time points. By employing a technique that involves multiplying individual components normalized across various spatial dimensions \(i\), \(j\), and \(k\), we delve into a granular analysis of the directional alignment and sensitivity to magnitude within these representations. Such an analysis holds particular significance in medical image segmentation tasks, where the interrelationships among these vectors significantly impact the precision with which the model delineates and separates distinct regions within an image. Instead of resorting to the conventional method of calculating dot products, we have chosen to utilize element-wise multiplication between the vectors representing the current local model (\(z_c\)) and both the server model (\(z_g\)) and the previous local model (\(z_p\))). This choice is motivated by the element-wise multiplication's ability to highlight specific elements where both feature vectors exhibit non-zero values. Essentially, this operation functions akin to a filter, amplifying elements where patterns align (thus emphasizing corresponding features) and attenuating elements where patterns diverge. The construction of the enhanced contrastive loss function (\(L_{EC}\)) is tailored to capitalize on the similarities between representations, with a specific emphasis on \(z_c\), \(z_p\), and \(z_g\). By quantifying the alignment and disparities among these representations, \(L_{EC}\) encourages the model to adapt to the overarching server model while preserving consistency with its prior states. The introduction of a temperature parameter (\(\tau\)) within the exponentials modulates the steepness of the probability distributions. A higher temperature facilitates a softer distribution, fostering a more nuanced exploration of model representations, whereas a lower temperature accentuates differences. Taking the logarithm of the ratio of exponentials and subsequently applying the negative function, as commonly observed in contrastive losses, effectively penalizes dissimilarities, prompting the model to prioritize similarities during its learning process. Furthermore, normalizing the loss by dividing it by \(N\), representing the total number of batch samples, ensures an average loss per sample, thus facilitating a coherent evaluation of model performance. The overarching equation for the enhanced contrastive loss function (\(L_{EC}\)) is thus expressed as:

\begin{equation} \label{eq: Enhanced Contrastive loss}
    L_{EC} = -\frac{1}{N} \sum_{i=1}^{N} \left[ \log\left( \frac{e^{\frac{\text{sim}(z_c, z_g)}{\tau}}}{e^{\frac{\text{sim}(z_c, z_g)}{\tau}} + e^{\frac{\text{sim}(z_c, z_p)}{\tau}}} \right) \right] \times \text{lr}.
\end{equation}

This formulation encapsulates the essence of the enhanced contrastive loss function, elucidating its pivotal role in steering the model's learning trajectory by accentuating adaptation to global representations while preserving consistency with historical states. The incorporation of the temperature parameter \(\tau\) and the learning rate \(\text{lr}\) endows this process with flexibility and control, thereby further augmenting the model's adaptability and convergence.

\section{Adaptive Scale-aware Weighted Aggregation (ASWA)}
In our proposed framework for federated learning, the adaptive scale-aware weighted aggregation (ASWA) method plays a crucial role in updating the server model \(w^{t+1}\) across all participating clients. This particular approach, which is executed centrally at the server, takes careful account of the varying sizes of local datasets (\(n_i\)) present on each client. The primary objective is to facilitate a nuanced and flexible integration of information from diverse sources. The determination of weights assigned to each client's contribution (\(w_i^t\)) is carried out meticulously by considering the ratio of its dataset size (\(n_i\)) to the total dataset size across all clients (\(N_{total}\)). This process can be succinctly represented mathematically as:

\begin{equation} \label{Eq: WEight update Equation}
    w^{t+1} = \sum_{i=1}^N \frac{n_i}{N_{total}} w_i^t.
\end{equation}

The essence of this adaptive scaling mechanism lies in its recognition that clients with larger datasets should exert a more significant influence on the global model update. By adopting this approach, a balanced assimilation of information is fostered, acknowledging the inherent diversity within local datasets. ASWA thus enhances the federated learning paradigm by dynamically adjusting the impact of individual clients based on the scale of their data holdings. This ensures the development of a global model that is not only robust but also representative, capturing the intricacies inherent in various local datasets.

### 2025-05-05 Chapters/appendix.tex
first commit

\section*{Code}
\begin{verbatim}
import nibabel as nib
import cv2
import gc
import os
import numpy as np
from glob import glob
import tensorflow as tf
from tqdm.auto import tqdm
import keras.backend as K
import matplotlib.pyplot as plt
from Codes.Dataloader import *
from Codes.Evalution_matrix import *
from Codes.own_method import *
from sklearn.preprocessing import MinMaxScaler
from sklearn.model_selection import train_test_split
import pandas as pd

class UNET3D(tf.keras.Model):
    def __init__(self, classes, dropout_rate=0.5, l2_reg=0.01):
        super(UNET3D, self).__init__()
        self.classes = classes
        self.dropout_rate = dropout_rate
        self.l2_reg = l2_reg

    def conv3D(self, x, filters, filter_size, activation='relu'):
        out = tf.keras.layers.Conv3D(
            filters,
            (filter_size, filter_size, filter_size),
            padding="same",
            kernel_regularizer=tf.keras.regularizers.l2(self.l2_reg)
        )(x)
        out = tf.keras.layers.BatchNormalization()(out)
        out = tf.keras.layers.Activation('relu')(out)
        out = tf.keras.layers.Dropout(self.dropout_rate)(out)
        return out

    def upsampling3D(self, x, filters, filter_size, stride=2, 
    activation='relu'):
        up_out = tf.keras.layers.Conv3DTranspose(
            filters, (filter_size, filter_size, filter_size),
            strides=(stride, stride, stride), padding='same'
        )(x)
        out = tf.keras.layers.BatchNormalization()(up_out)
        return out

    def concatenate(self, x1, x2):
        concat = tf.keras.layers.concatenate([x1, x2])
        return concat

    def max_pool3D(self, x, filter_size, stride):
        out = tf.keras.layers.MaxPooling3D(
            (filter_size, filter_size, filter_size), strides=stride
        )(x)
        return out

    def downconv(self, x, filters):
        s1 = self.conv3D(x, filters, 3)
        s2 = self.conv3D(s1, filters, 3)
        return s1

    def upconv(self, x, filters, skip_connection):
        e1 = self.upsampling3D(x, filters, 2)
        concat = self.concatenate(e1, skip_connection)
        conv1 = self.conv3D(concat, filters, 3)
        conv2 = self.conv3D(conv1, filters, 3)
        return conv2

    def call(self):
        inputs = tf.keras.layers.Input(shape=(64, 64, 64, 1))
        d1 = self.downconv(inputs, 32)
        m1 = self.max_pool3D(d1, filter_size=2, stride=2)
        d2 = self.downconv(m1, 64)
        m2 = self.max_pool3D(d2, filter_size=2, stride=2)
        d3 = self.downconv(m2, 128)
        m3 = self.max_pool3D(d3, filter_size=2, stride=2)
        d4 = self.downconv(m3, 256)
        m4 = self.max_pool3D(d4, filter_size=2, stride=2)

        bridge = self.conv3D(m4, 1024, 3, 1)
        bridge = self.conv3D(bridge, 1024, 3, 1)

        u1 = self.upconv(bridge, 256, d4)
        u2 = self.upconv(u1, 128, d3)
        u3 = self.upconv(u2, 64, d2)
        u4 = self.upconv(u3, 32, d1)

        logits = tf.keras.layers.Conv3D(self.classes, (1, 1, 1), 
        padding="same")(u4)
        logits = tf.nn.sigmoid(logits)

        model = tf.keras.Model(inputs=[inputs], outputs=[bridge, logits])
        return model

unet1 = UNET3D(classes=1, dropout_rate=0.2, l2_reg=0.01)

imgepath= "/kaggle/input/gbm-64/GBM/image"
maskpath= "/kaggle/input/gbm-64/GBM/mask"

images_file = os.listdir(imgepath)
masks_file = os.listdir(maskpath)

npy_images_path = [os.path.join(imgepath, img_file) for 
img_file in images_file[:]]
npy_masks_path = [os.path.join(maskpath,mas_file) for 
mas_file in masks_file[:]]
npy_images_path.sort()
npy_masks_path.sort()

metrics = [iou,dice_coef,precision,sensitivity,specificity]

LR = 0.00001
batch_size = 5
epochs = 151

data ={}
data["original_data"] = mini_batches_(npy_images_path[:100],
npy_masks_path[:100],batch_size)

data["model"] = unet1.call()
data["optimizer"] = tf.keras.optimizers.Adam(LR)

test_data_gen = mini_batches_(npy_images_path[100:110], 
npy_masks_path[100:110],batch_size)

plot_dice_losses=[]
plot_total_loss=[]
plot_dice_co=[]

past_model = None

for epoch in range(epochs):
    print(epoch+1,'/',epochs)
    model = data['model']
    optimizer = data["optimizer"]
    # optimizer = optimizer
    model.set_weights(data["model"].get_weights())

    dice = []
    pre = []
    batch_loss = []
    se = []
    spe = []
    io = []
    dice_losses=[]
    contrastive_losses=[]

    for batch_idx, (images, masks) in enumerate(tqdm(data['original_data'])):
        pro2, _ = data['model'](images)
        with tf.GradientTape() as tape:
            pro1 , logits = model(images)
            loss_combined = tversky_loss(y_true=masks, y_pred=logits)

            loss = loss_combined
            batch_loss.append(loss)

            dice_loss_value = dice_loss2(masks, logits)
            contrastive_loss_value = contrastive_loss(prev=pro2, pres=pro1)

            dice_losses.append(dice_loss_value)
            contrastive_losses.append(contrastive_loss_value)

            # metrics
            dice.append(dice_coef(masks, logits))
            pre.append(precision(masks, logits))
            se.append(sensitivity(masks, logits))
            spe.append(specificity(masks, logits))
            io.append(iou(masks, logits))

        grads = tape.gradient(loss, model.trainable_variables)
        gradients, _ = tf.clip_by_global_norm(grads, clip_norm=1.0)
        optimizer.apply_gradients(zip(gradients, model.trainable_variables))

    # local_weight_list.append(model.get_weights())
    batch_loss = np.array(batch_loss).mean()
    dice = np.array(dice).mean()
    pre = np.array(pre).mean()
    se = np.array(se).mean()
    spe = np.array(spe).mean()
    io = np.array(io).mean()

    data['past_model'] = model
    data['model'] = model

    mean_dice_loss = np.mean(dice_losses)
    mean_contrastive_loss = np.mean(contrastive_losses)
    print("\nWeighted Combined Loss: {} | Dice Coeff: {}  |
    \n\n Precision: {} Sensitivity: {} | 
    \n\n Specificity: {} , IOU: {} |"
    .format(batch_loss, dice, pre, se, spe, io))

    plot_dice_losses.append(mean_dice_loss)
    plot_total_loss.append(batch_loss)
    plot_dice_co.append(dice)

    if epoch%10 == 0 and epoch != 0:
        dice_test = []
        pre_test = []
        batch_loss_test=[]
        se_test=[]
        spe_test = []
        io_test = []
        model_test = data["model"]
        for batch_idx, (images, masks) in enumerate(tqdm(test_data_gen)):
            predictions, logits = model(images)

            loss = dice_loss2(masks,logits)
            batch_loss_test.append(loss)

            dice_test.append(dice_coef(masks, logits))
            pre_test.append(precision(masks, logits))
            se_test.append(sensitivity(masks, logits))
            spe_test.append(specificity(masks, logits))
            io_test.append(iou(masks, logits))

        batch_loss_test = np.array(batch_loss_test).mean()
        dice_test = np.array(dice_test).mean()
        pre_test =np.array(pre_test).mean()
        se_test =  np.array(se_test).mean()
        spe_test =  np.array(spe_test).mean()
        io_test =  np.array(io_test).mean()
        print("----------->Test<-----------")
        print("Epoch: {} , Loss: {} , Dice Coeff: {}
        \n\n, Precision: {} Sensitivity: {} 
        \n\n Specificity: {} , IOU {}".format(epoch+1,batch_loss_test,dice_test,
        pre_test,se_test,spe_test,io_test))
        print("----------->Trining Conti<-----------")

plt.figure(figsize=(10, 5))
plt.plot(plot_dice_losses, label='Dice Loss', color='blue')
plt.title('Dice Loss Over Epochs')
plt.xlabel('Epochs')
plt.ylabel('Dice Loss')
plt.legend()
plt.show()

# Plot Weighted Combined Loss
plt.figure(figsize=(10, 5))
plt.plot(plot_total_loss, label='Weighted Combined Loss', color='red')
plt.title('Weighted Combined Loss Over Epochs')
plt.xlabel('Epochs')
plt.ylabel('Weighted Combined Loss')
plt.legend()
plt.show()

\end{verbatim}

### 2025-05-05 Chapters/chapter 3.tex
first commit

\section{Methodology}

In this study, we propose a hybrid model combining Graph Convolutional Networks (GCNs) and Transformer encoders for sleep stage classification from EEG signals. Our approach captures both spatial relationships between EEG channels using GCNs and temporal dependencies using Transformer encoders. Below, we describe the methodology in detail.

biosignals as input and filter the four channels. The filtered signals are passed through a graph convolution layer, followed by a pooling layer to reduce their size. Next, we adjust the batch size and feed the output into a transformer encoder. Finally, we obtain the softmax classification into five stages.

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{img/Architechture.png}
    \caption{Proposed Architecture of GCN-Transformer Hybrid Approach}
    \label{fig:enter-label}
\end{figure}

\subsection{Input Data and Preprocessing}

We use EEG data represented as graph-structured inputs. Each sample contains 4 EEG channels recorded at 3000 frequency points. The dataset dimensions are:
\begin{itemize}
    \item \textbf{X\_all}: (11076, 4, 3000) - Samples, Channels, Frequency Points
    \item \textbf{Y\_all}: (11076,) - Labels for sleep stages
    \item Converted to PyTorch tensors:
    \begin{itemize}
        \item \textbf{X\_tensor}: torch.Size([11076, 4, 3000])
        \item \textbf{Y\_tensor}: torch.Size([11076])
    \end{itemize}
\end{itemize}

\subsubsection{Sleep Stage Mapping}

The sleep stages are mapped as follows:

\begin{table}[h]
\centering
\caption{Mapping of Original Sleep Stages to Labels}
\[
\begin{array}{|c|c|}
\hline
\textbf{Original Stage} & \textbf{Mapped Label} \\
\hline
\text{Sleep stage W} & 0 \\
\text{Sleep stage 1} & 1 \\
\text{Sleep stage 2} & 2 \\
\text{Sleep stage 3} & 3 \\
\text{Sleep stage 4} & 3 \\
\text{Sleep stage R} & 4 \\
\hline
\end{array}
\]
\label{tab:sleep_stage_mapping}
\end{table}

\subsubsection{Channel Selection}

The following EEG channels are selected for analysis: EEG Fpz-Cz, EEG Pz-Oz, EMG submental, and EOG horizontal.

\subsubsection{Preprocessing}

\textbf{Epoch Segmentation:}  
The EEG signals are segmented into 30-second epochs. Each epoch contains 3000 samples per channel, resulting in the final shape of the data as:

\[
\text{Final Data Shape: } [X, 4, 3000]
\]

\textbf{Band-Pass Filtering:}  
A band-pass filter with a frequency range of 0.3 - 30 Hz is applied to the signals. This removes frequencies outside this range, filtering out high-frequency noise and low-frequency drift.

\subsection{Graph Construction}

We represent each sample as a graph, where nodes correspond to EEG channels and edges represent spatial relationships. The graph is defined as:
\begin{itemize}
    \item \textbf{x}: Node features (shape: [3000, 4]) - EEG signals per channel
    \item \textbf{edge\_index}: Graph connectivity (shape: [2, 12])
    \item \textbf{y}: Sleep stage label for each sample (shape: [1])
\end{itemize}

\subsubsection{Methodology: Graph Dataset Creation}

The graph's adjacency matrix represents the spatial relationship between EEG channels. The edge weights are derived from the following graph adjacency matrix:

\begin{table}[h]
\centering
\caption{Graph Weight Representation}
\[
\begin{array}{|c|c|c|c|c|}
\hline
\textbf{} & \textbf{Fpz-Cz} & \textbf{Pz-Oz} & \textbf{EMG} & \textbf{EOG} \\
\hline
\textbf{Fpz-Cz} & 0 & 0.9 & 0.6 & 0.6 \\
\textbf{Pz-Oz}  & 0.9 & 0 & 0.6 & 0.6 \\
\textbf{EMG}    & 0.6 & 0.6 & 0 & 0.5 \\
\textbf{EOG}    & 0.6 & 0.6 & 0.5 & 0 \\
\hline
\end{array}
\]
\label{tab:correlation_matrix}
\end{table}

This matrix is used to define the weights of the edges between the EEG channels in the graph. The value at position \((i, j)\) indicates the weight of the edge between node \(i\) and node \(j\).

\begin{table}[ht]
\centering
\caption{Dataset shape Information}
\begin{tabular}{|c|c|}
\hline
\textbf{Component} & \textbf{Details} \\
\hline
\textbf{Total Samples} & 11,076 \\
\hline
\textbf{Example Sample Format} & Data(x=[3000, 4], edge index=[2, 12], edge attr=[12], y=[1]) \\
\hline
\textbf{x (Node features)} & EEG signals (shape: [3000, 4]) \\
\hline
\textbf{edge index} & Connectivity between EEG channels (shape: [2, 12]) \\
\hline
\textbf{edge attr} & Weights of edges between EEG channels (shape: [12]) \\
\hline
\textbf{y (Label)} & Sleep stage label for the sample (shape: [1]) \\
\hline
\end{tabular}

\end{table}

\subsubsection{Graph Representation of EEG Channels}
The EEG channels are represented as nodes in the graph, with edges indicating spatial relationships based on the adjacency matrix. The resulting graph structure captures both the spatial dependencies between the channels and the temporal dynamics of the EEG signals.
\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\linewidth]{img/Graph Weightage.png}
    \caption{Graph Weight Visulization}
    \label{fig:enter-label}
\end{figure}

\subsection{GCN Layers for Spatial Feature Extraction}

The GCN module captures spatial dependencies among EEG channels. We employ five Graph Convolutional Network (GCN) layers to learn the connectivity patterns between EEG channels, thereby enhancing feature extraction by leveraging the graph structure of EEG data. The architecture of each GCN layer is described as follows:

\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth]{img/Graph Convolution Neural Network.png}
    \caption{Graph Convolution Neural Network}
    \label{fig:enter-label}
\end{figure}
\begin{table}[ht]
\centering
\caption{GCN Layer Architecture}

\begin{tabular}{|c|c|}
\hline
\textbf{Layer} & \textbf{Operations} \\
\hline
\textbf{GCN Layer 1} & GCNConv(4, 256) $\rightarrow$ BatchNorm(256) $\rightarrow$ ReLU $\rightarrow$ Dropout(0.1) \\
\hline
\textbf{GCN Layer 2 to 5} & GCNConv(256, 256) $\rightarrow$ BatchNorm(256) $\rightarrow$ ReLU $\rightarrow$ Dropout(0.1) \\
\hline
\end{tabular}
\end{table}

The GCN layers work by taking the input features, which are of shape (4, 256) after the first layer, then passing them through a \textbf{Batch Normalization} layer with 256 output channels. The output is then activated using the \textbf{ReLU} activation function, formulated as:

\[
\text{ReLU}(x) = \max(0, x)
\]

This is followed by \textbf{Dropout}(0.1) for regularization, helping to avoid overfitting by randomly setting 10\% of the input units to zero.

After passing through all five GCN layers, the final output tensor has a shape of \( (3000, 256) \), representing 3000 frequency points with a 256-dimensional feature representation for each frequency point across all EEG channels.

To obtain a graph-level representation, we apply \textbf{global mean pooling} across nodes, which reduces the output to a \( \text{(Batch\_size}=11076, 256) \) tensor. This means that each sample is represented by a single 256-dimensional feature vector. The hidden dimension in this final layer is 256, which captures the most relevant features of the input data.

\subsection{Transformer Encoder for Temporal Dependencies}

\begin{table}[!h]
\centering
\caption{Transformer Encoder Overview}

\begin{tabular}{|c|c|}
\hline
\textbf{Component} & \textbf{Details} \\
\hline
\textbf{Preprocessing} & Expand graph embedding to (Batch\_size, 1, 256) \\
\hline
\textbf{Transformer Encoder} & 2 Transformer Encoder Layers \\
\hline
\textbf{d\_model (Hidden dimension)} & 256 \\
\hline
\textbf{nhead (Number of heads)} & 4 \\
\hline
\textbf{Dropout} & 0.1 \\
\hline
\textbf{batch first} & True \\
\hline
\textbf{Postprocessing} & Squeeze output to (Batch\_size, 256) \\
\hline
\textbf{Fully Connected Layer} & Linear(256 $\rightarrow$ 5) \\
\hline
\textbf{Output} & Logits for 5-class classification \\
\hline
\end{tabular}
\end{table}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\linewidth]{img/transformer.png}
    \caption{Enter Caption}
    \label{fig:enter-label}
\end{figure}

The Transformer architecture follows the GCN layer pooling shapes. After pooling, the embeddings are passed through the multi-head attention mechanism, followed by Add \& Norm, which normalizes the output. This is then followed by a feed-forward layer, dropout with a 0.1 probability, and another Add \& Norm. The input embeddings are connected in parallel with the Add \& Norm. Positional encoding is applied during input to account for the temporal order of the sequence. The batch size is set to 1, with a sequence length of 1 and a hidden dimension of 256. The input passes through two Transformer Encoder layers, followed by a fully connected linear layer with a transformation from 256 to 5, which outputs the logits for 5-class classification. The final output is obtained using the softmax function, represented by:

\[
\text{Softmax}(z_i) = \frac{e^{z_i}}{\sum_j e^{z_j}}
\]

\subsection{Loss Function and Optimization}
We employ Focal Loss to address class imbalance, with parameters:
\begin{itemize}
    \item Gamma ($\gamma$): 3
    \item Label smoothing: 0.1
    \item Class weights computed from the dataset
\end{itemize}

The optimizer and learning rate scheduler are configured as follows:
\begin{itemize}
    \item \textbf{Optimizer}: AdamW (learning rate=0.0003, weight decay=1e-4)
    \item \textbf{Scheduler}: Cosine Annealing Learning Rate Scheduler
\end{itemize}

\subsubsection{Why Focal Loss Instead of Standard Cross-Entropy?}
\textbf{Motivation for Focal Loss:}  
Standard Cross-Entropy treats all samples equally, leading to bias towards majority classes. In imbalanced datasets, minority class predictions get suppressed. Focal Loss dynamically adjusts the loss contribution based on prediction confidence. It reduces the importance of well-classified samples and focuses more on hard-to-classify ones.

\textbf{Key Features of Focal Loss:}
\begin{itemize}
    \item Introduces a focusing parameter $\gamma$ to adjust class weighting.
    \item Includes class weighting factor $\alpha$ to handle imbalance.
    \item Works well for highly imbalanced datasets in classification tasks.
\end{itemize}

\subsubsection{Focal Loss Formulation}
The Focal Loss is formulated as follows:

\[
FL(p_t) = -\alpha (1 - p_t)^{\gamma} \log(p_t)
\]
where:
\begin{itemize}
    \item $p_t$ is the predicted probability for the target class.
    \item $\alpha$ is the weighting factor for class imbalance.
    \item $\gamma$ is the focusing parameter (higher values focus more on hard examples).
\end{itemize}

\subsubsection{Implementation Details}
\textbf{Label Smoothing:} Label smoothing is applied to prevent the log(0) issue by adding a small constant $\epsilon$:

\[
y_{\text{smooth}} = y(1 - \epsilon) + \frac{\epsilon}{C}
\]
where $C$ is the number of classes.

The PyTorch-based computation for Focal Loss with label smoothing becomes:

\[
L = \alpha (1 - p)^{\gamma} (-y_{\text{smooth}} \log p)
\]

\subsubsection{Why Use a Learning Rate Scheduler?}
\textbf{Importance of Learning Rate Scheduling:} The learning rate is crucial for training deep models efficiently. A high learning rate can lead to divergence, while a low one may cause slow convergence. Adaptive learning rate schedules help balance stability and speed.

\textbf{Why CosineAnnealingLR?} Cosine Annealing smoothly reduces the learning rate following a cosine decay. It starts with a large step size for exploration and gradually fine-tunes. This helps avoid sharp drops in the learning rate, improving generalization.

\subsubsection{Cosine Annealing Learning Rate Decay}
The learning rate decay is governed by the formula:

\[
\eta_t = \eta_{\min} + \frac{1}{2} (\eta_{\max} - \eta_{\min}) \left( 1 + \cos\left( \frac{T_{\text{cur}}}{T_{\text{max}}} \pi \right) \right)
\]
where:
\begin{itemize}
    \item $\eta_t$ is the learning rate at epoch $t$.
    \item $\eta_{\max}$ and $\eta_{\min}$ are the max/min learning rates.
    \item $T_{\text{cur}}$ is the current epoch.
    \item $T_{\text{max}}$ is the total number of epochs.
\end{itemize}

\textbf{Key Benefits:}
\begin{itemize}
    \item Encourages large updates early in training.
    \item Smoothly transitions into finer updates as training progresses.
    \item Helps the model avoid getting stuck in poor local minima.
\end{itemize}

\subsubsection{Training Methodology: Overview}
\textbf{SleepTrainer Class: Key Features:}
\begin{itemize}
    \item Handles model training, validation, and optimization.
    \item Uses Focal Loss to address class imbalance.
    \item Applies CosineAnnealingLR scheduler for smooth learning rate decay.
\end{itemize}

\textbf{Training Process:}
\begin{itemize}
    \item Compute class weights for imbalanced data.
    \item Iterate through training batches, compute loss, and update weights.
    \item Validate model performance on a separate validation set.
    \item Adjust learning rate dynamically using a scheduler.
\end{itemize}

\subsubsection{Training Methodology: Hyperparameters}
\textbf{Key Hyperparameters:}
\begin{itemize}
    \item Batch Size: 32
    \item Learning Rate: 0.0003
    \item Weight Decay: 1e-4
    \item Epochs: 20
    \item Optimizer: AdamW
    \item Learning Rate Scheduler: CosineAnnealingLR
\end{itemize}
Gradually reduces learning rate over time for smooth convergence, helping prevent sudden drops in performance.

\subsubsection{Training Methodology: Handling Class Imbalance}
\textbf{Why Compute Class Weights?}  
EEG sleep data is imbalanced, with some sleep stages appearing more frequently. Without weighting, the model may favor majority classes. Weights ensure rare classes contribute more to the loss.

\textbf{Class Weight Computation:}
The class weight for class $c$ is computed as:

\[
w_c = \left( \frac{\text{Total Samples}}{\text{Class Count} + 1} \right)^{0.5}
\]
where:
\begin{itemize}
    \item $w_c$ is the computed weight for class $c$.
    \item Smaller classes receive higher weights.
\end{itemize}

These weights are applied to the Focal Loss during training.

### 2025-05-05 Chapters/chapter4.tex
first commit

\subsection{Testing Data Distribution Analysis}

\textbf{Why Ensure Balanced Testing Data?}
\begin{itemize}
    \item Prevents bias toward majority classes.
    \item Ensures the model’s performance is fairly evaluated.
    \item Helps achieve reliable generalization across all sleep stages.
\end{itemize}

The figure below shows the normalized class distribution during testing. Each class maintains an equalized density, avoiding class imbalance. This confirms that the model’s evaluation is not biased toward any specific sleep stage.

\textbf{Sampling Density Plot Showing Balanced Class Distribution:}
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/sample distribution plot pdf.png}
    \caption{Balanced class distribution across the test set.}
\end{figure}

\subsection{Model Performance: Training vs Testing}

Figures below demonstrate the accuracy and loss curves during training and testing, providing a clear visual comparison of model performance:

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/accuracy plot.png}
    \caption{Accuracy Curve: Training vs Testing.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/loss plot.png}
    \caption{Loss Curve: Training vs Testing.}
\end{figure}

\subsection{Model Evaluation: Confusion Matrix}

Performance metrics are calculated to ensure a comprehensive evaluation of the model's performance across all classes. We use precision, recall, and F1-score to assess the model:

\[
\text{Precision} = \frac{TP}{TP + FP}
\]
\[
\text{Recall} = \frac{TP}{TP + FN}
\]
\[
\text{F1-Score} = \frac{2 \times \text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}
\]

These metrics ensure a balanced evaluation of model performance across all sleep stages.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/confusion matrix samples.png}
    \caption{Confusion Matrix showing model performance across sleep stages.}
\end{figure}

\subsection{Gradient Analysis: Training Progression}

Understanding model training dynamics:

\textbf{Early Training (Epochs 0-5):} High loss with accuracy starting to improve.  
\textbf{Mid Training (Epochs 5-15):} Loss steadily decreases with stable gradient flow.  
\textbf{Late Training (Epochs 15-20):} Accuracy plateaus with no severe overfitting.

\textbf{Conclusion:} The training process remains stable, with no vanishing or exploding gradients.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/3d gradient.png}
    \caption{Gradient 3D Surface: Training vs Validation Metrics.}
\end{figure}

\subsection{Performance Metrics: Precision, Recall, F1-Score}

Evaluating model performance across all classes using key metrics:

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/precision_plot.png}
    \caption{Precision Scores per Class.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/recall_plot.png}
    \caption{Recall Scores per Class.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/f1_score_plot.png}
    \caption{F1 Scores per Class.}
\end{figure}

\subsection{Feature Importance Analysis with LIME}

Understanding the contribution of different channels to model predictions, we used LIME (Local Interpretable Model-agnostic Explanations) to analyze feature importance. The analysis showed that the \textbf{EMG submental} and \textbf{EEG Pz-Oz} channels contribute the most to predictions, while the \textbf{EOG horizontal} channel has minimal importance, indicating lower relevance for classification.

This insight helps optimize feature selection and improve model efficiency.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/feature importance chanels analysis.png}
    \caption{Feature Importance Analysis for 4 Channels.}
\end{figure}

### 2025-05-05 Chapters/conclusion.tex
first commit

\section{Conclusion}
Our proposed SleepGCN-Transformer model achieves 93.12\% training accuracy and
93.04\% validation accuracy, demonstrating its effectiveness in sleep stage classification.
The integration of Graph Convolution Networks (GCN) captures spatial dependencies
across EEG, EOG, and EMG channels, while the Transformer extracts temporal patterns. The
use of Focal Loss enhances class balancing, improving performance on underrepresented sleep
stages. Feature importance analysis highlights EMG and EEG Pz-Oz as key predictors. This
robust approach lays the foundation for future work in Explainable AI, enabling medical
professionals to interpret AI-driven sleep diagnostics effectively.

### 2025-05-05 Chapters/introduction.tex
first commit

Sleep is an essential component of human health, contributing significantly to cognitive function, memory consolidation, and emotional regulation. Adequate and quality sleep ensures optimal brain performance and supports overall physical and mental well-being. Disruptions in sleep patterns can lead to various health issues, including impaired concentration, mood disorders, and chronic conditions such as cardiovascular diseases.

The classification of sleep stages is fundamental in understanding sleep architecture and diagnosing sleep disorders. Experts categorize sleep into five primary stages—N1, N2, N3, REM, and Wake—based on distinct patterns observed in brain activity. Each sleep stage is characterized by specific brainwave frequencies, which can be detected using electroencephalography (EEG) signals. Accurate classification of these stages is crucial for analyzing sleep quality and identifying abnormalities such as insomnia, sleep apnea, and narcolepsy.

Table~\ref{tab:sleep_stages} provides an overview of the frequency ranges associated with each sleep stage.

\begin{table}[h]
\centering
\caption{Frequency ranges of different sleep stages.}
\begin{tabular}{ll}
\toprule
\textbf{Sleep Stage} & \textbf{Frequency Range (Hz)} \\
\midrule
Wake (Beta waves)       & 12--30 \\
N1 (Light Sleep)        & 4--8  \\
N2 (Moderate Sleep)     & 4--6  \\
N3 (Deep Sleep, Delta)  & 0.5--4  \\
REM (Theta waves)       & 4--6  \\
\bottomrule
\end{tabular}
\label{tab:sleep_stages}
\end{table}

To classify sleep stages effectively, bio-signals from multiple physiological channels are analyzed. The primary sources of information for sleep stage classification are EEG (electroencephalogram), EMG (electromyogram), and EOG (electrooculogram) signals. These channels capture brain activity, muscle movement, and eye movemets, respectively, offering a comprehensive view of sleep patterns.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\linewidth]{img/Sleep EDF Data Visulization.png}
    \caption{Sleep EDF Data Visulization}
    \label{fig:enter-label}
\end{figure}

Despite the availability of multiple channels in the SleepEDF dataset, not all contribute equally to classification performance. To reduce model complexity without compromising accuracy, we selectively utilize four channels from the nine standard channels provided in the dataset. The selected channels are:

\begin{itemize}
    \item \textbf{EEG Fpz-Cz}: Captures frontal-to-central brain activity.  
    \item \textbf{EEG Pz-Oz}: Measures parietal-to-occipital brain activity.  
    \item \textbf{EMG submental}: Records muscle tone from the chin area.  
    \item \textbf{EOG horizontal}: Tracks horizontal eye movements.  
\end{itemize}

In this study, we adopt a graph-based modeling approach to capture spatial relationships between the selected channels. Additionally, we employ a Transformer encoder to effectively model temporal dependencies across sleep cycles. To facilitate a better understanding of the input data, we also present visualizations of bio-signals from all channels.

This approach aims to leverage the strengths of both spatial and temporal modeling techniques, resulting in a robust and efficient sleep stage classification framework.

### 2025-05-05 Title/abstract.tex
first commit

\begin{center}
\section*{\textcolor{internationalkleinblue}{\textbf{{Abstract}}}}
\end{center} 
This project presents \textit{SleepGCN-Transformer}, a hybrid model combining Graph Convolutional Networks (GCN) and Transformer encoders for sleep stage classification. Using the SleepEDF dataset, it incorporates four physiological signals: EEG (Fpz-Cz, Pz-Oz), EMG (submental), and EOG (horizontal). Preprocessing includes 30-second epochs, band-pass filtering (0.3--30 Hz), and graph-based EEG channel representation.

The GCN module captures spatial relationships across EEG channels, while the Transformer encoder models temporal dependencies from graph-level embeddings. Focal Loss addresses class imbalances, and a CosineAnnealingLR scheduler optimizes learning rate decay. Training with the AdamW optimizer for 20 epochs achieves 93.12 \% training and 93.04\% validation accuracy.

Model performance, assessed using precision, recall, and F1-score, demonstrates high efficacy. LIME-based feature importance analysis highlights EMG and EEG Pz-Oz channels as key contributors. This hybrid model exhibits state-of-the-art performance, with future directions focused on enhancing explainability and clinical integration.

### 2025-05-05 Title/acknowledgement.tex
first commit

%\begin{center}
%\textbf{\large Acknowledgment}
%\end{center} 
\section*{\textcolor{internationalkleinblue}{\textbf{Acknowledgment}}}
 While my name appears as the sole contributor to the completion of this summer internship, it is important to recognize that the guidance and support of many individuals played a significant role in its success. This internship is the result of a collective effort, shaped by the contributions of numerous people, to whom I am deeply grateful.

 I extend my heartfelt appreciation to my internal guide, Dr. Santosh Kumar Satapathy
Author, for his consistent guidance, encouragement, and insightful suggestions throughout this internship. His belief in my abilities provided the confidence and motivation necessary to complete this project. Dr. Santosh Kumar Satapathy
Author’s support, along with his generous provision of time and access to resources, was instrumental in achieving the objectives of this internship.

 I am also profoundly thankful for the unwavering support of my parents. Their continuous inspiration, moral backing, and blessings have been the cornerstone of my journey. Their understanding and encouragement have been invaluable, and I am truly fortunate to have had their steadfast support throughout this endeavor.
\begin{flushright}
\textbf{Tanmay Rathod}\\
\end{flushright}

### 2025-05-05 Title/approval.tex
first commit

\begin{center}
    \textbf{\large Approval Sheet}
\end{center}

This thesis entitled \textbf{\enquote{SleepGCN-Transformer: A Hybrid Graph
Convolutional and Transformer Network for Sleep
Stage Classification}} by \textbf{Tanmay Rathod} is recommended for the degree of \textbf{M.Tech} in \textbf{Artificial Intelligence}.
\vspace{0.8cm}

\hspace{9cm}  
\begin{flushright}   
    \makebox[1.8in]{\hrulefill}\\
    Guide : Dr Santosh Satapathy, \\
    Assistant Professor \\
    Dept. of ICT \\

\end{flushright}
\vspace{0.5cm}

\hspace{9cm}  
\begin{flushright}   
    \makebox[1.8in]{\hrulefill}\\
    Dr Nitin Singh Rajput, \\
    Assistant Professor \\
    Dept. of ICT
\end{flushright}
\vspace{0.5cm}

% \hspace{9cm}\\
\begin{flushright}
    \makebox[1.8in]{\hrulefill}\\
    Dr. Paawan Sharma,\\
    Head Of Department \\ 
    Dept. of ICT
\end{flushright}

\vfill
\begin{flushleft}
    Date: \makebox[1.8in]{\hrulefill}\\
    Place: \makebox[1.8in]{\hrulefill}\\    
\end{flushleft}

### 2025-05-05 Title/conclusion.tex
first commit

--- 2025-05-05 Chapters/conclusion.tex
@@
-\section{Conclusion}
 Our proposed SleepGCN-Transformer model achieves 93.12\% training accuracy and

### 2025-05-05 Title/declaration.tex
first commit

\begin{center}
	\textbf{\large Student Declaration}
\end{center}

I, \textcolor{internationalkleinblue}{\textbf{Tanmay Rathod}}, hereby declare that this written submission represents my ideas in my own words, and where others’ ideas or words have been included, I have adequately cited and referenced the original sources. I also declare that I have adhered to all principles of academic honesty and integrity and have not misrepresented or fabricated, or falsified any idea/data/fact/source in my submission. I understand that any violation of the above will be cause for disciplinary action by the Pandit Deendayal Energy University and can also evoke penal action from the sources which have thus not been properly cited or from whom proper permission has not been taken when needed.
\vspace{0.8cm}
\vspace{0.8cm}
\begin{flushright}
    \makebox[1.8in]{\hrulefill}\\
    Tanmay Rathod\\
    Roll No: 23MAI007\\
\end{flushright}
\vfill
\begin{flushleft}
	Date: \makebox[1.8in]{\hrulefill}
\end{flushleft}

### 2025-05-05 Title/publications.tex
first commit

%\fancyhead[L]{\fontsize{8}{8}\selectfont\itshape\nouppercase{List of Publication}}
%\thispagestyle{empty}

\section*{\textcolor{internationalkleinblue}{\textbf{Publications on Related Topic}}}
\begin{enumerate}
	\item \textbf{C Chodvadiya}, P Sharma, \textquotedblleft Class-Enhanced Mahalanobis Distance Contrastive (CEMDC) Loss for Brain Tumor Segmentation, \textquotedblright \,\textcolor{internationalkleinblue}{\textbf{Under Review with Professor}}

    \item P Divya, \textbf{C Chodvadiya}, B Raman, YW Chen, CK Mohan, \textquotedblleft FSF3A: Federated Spatial Feature Alignment and Adaptive Aggregation for Heterogeneous Brain Tumor Segmentation, \textquotedblright \,\textcolor{internationalkleinblue}{\textbf{Under Review}} in 27TH EUROPEAN CONFERENCE ON ARTIFICIAL INTELLIGENCE.

\end{enumerate}
\newpage

\section*{\textcolor{internationalkleinblue}{\textbf{Other Publications During M.Tech}}}
\begin{enumerate}
   	\item \textbf{C Chodvadiya}, KG Singh, R Vyas*, \textquotedblleft Detection of Palmprint from Contactless Smartphone-Based Video Hand Dataset, \textquotedblright  NCVPRIPG-CHBGR, 2023.

    \item \textbf{C Chodvadiya*}, V Solanki, KG Singh, \textquotedblleft Intelligent Virtual Worlds: A Survey of the Role of AI in the Metaverse, \textquotedblright  In 2024 3rd IEEE International Conference for Innovation in Technology (INOCON).

    \item KG Singh, \textbf{C Chodvadiya}, C Bhatt, P Shah, A Bruno, \textquotedblleft Seeing in the Dark: A Different Approach to Night Vision Face Detection with Thermal IR Images, \textquotedblright  In CEURWORKSHOP PROCEEDINGS. CEUR-WS, 2023.\\ \textbf{DOI : }\textcolor{internationalkleinblue}{\textbf{\url{https://ceur-ws.org/Vol-3563/paper_1.pdf}}}

    \item KG Singh, \textbf{C Chodvadiya}, S Vyas, H Patel, S Mishra, B Tripathi, \textquotedblleft Prediction of Solar Power Generation and Maintenance Activities for1MWPower Plant, \textquotedblright In 2023 IEEE IAS Global Conference on Emerging Technologies (GlobConET), pp. 1-6. IEEE, 2023. \\ \textbf{DOI : }\textcolor{internationalkleinblue}{\textbf{\url{10.1109/GlobConET56651.2023.10150076}}}

    \item KG Singh, \textbf{C Chodvadiya}, S Mishra, \textquotedblleft Optimizing Solar Power Forecasting Patterns using Multi-factor Cumulative Casualty matrix, \textquotedblright \,\textcolor{internationalkleinblue}{\textbf{Under Review with Professor}}

\end{enumerate}

### 2025-05-05 Title/title.tex
first commit

\begin{center}
    {{\Large \textbf{SleepGCN-Transformer: A Hybrid Graph Convolutional and
Transformer Network for Sleep Stage Classification
}}}\\
    \vspace{0.6cm}
    {\large A Master's Report}\\
    {\large Submitted in partial fulfillment of the requirement for the award of the degree of}\\
    \vspace{0.8cm}
    {\large Master of Technology in Artificial Intelligence}\\
    {\large by}\\
    \textbf{{\Large  Tanmay Rathod}}\\
    {\large 23MAI007}\\
    \vspace{0.7cm}
    {\large Under the guidance of}\\
    {\large  \textbf{Dr. Santosh Kumar Satapathy}  {\large }}\\
    {\large Department of Information and Communication Technology}\\
    \vspace{0.6cm}
        \vfill
    {\centering \includegraphics[width=0.21\textwidth]{img/pdpu_logo.png}}\\
    {\large School of Technology}\\
    {\large Pandit Deendayal Energy University}\\
    {\large Gandhinagar – 382426. Gujarat - India}\\
    {\large February, 2025}
\end{center}

### 2025-05-05 Chapters/chapter4.tex
changes

@@
 	\item Prevents bias toward majority classes.
-    \item Ensures the model’s performance is fairly evaluated.
-    \item Helps achieve reliable generalization across all sleep stages.
+	\item Ensures fair evaluation across all sleep stages.
+	\item Enhances generalization reliability.
 \end{itemize}
 
-The figure below shows the normalized class distribution during testing.
-Each class maintains an equalized density, avoiding class imbalance.
-This confirms that the model’s evaluation is not biased toward any specific sleep stage.
+The figure below shows the normalized class distribution during testing, confirming balance and the absence of class imbalance.
 
-\textbf{Sampling Density Plot Showing Balanced Class Distribution:}
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/sample distribution plot pdf.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/sample_distribution_plot_pdf.png}
 	\caption{Balanced class distribution across the test set.}
@@
 
-\subsection{Model Performance: Training vs Testing}
+\subsection{Model Performance: Accuracy and Loss Trends}
 
-Figures below demonstrate the accuracy and loss curves during training and testing, providing a clear visual comparison of model performance:
+Figures below illustrate the model's training vs testing accuracy and loss progression over epochs.
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/accuracy plot.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/accuracy_plot.png}
 	\caption{Accuracy Curve: Training vs Testing.}
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/loss plot.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/loss_plot.png}
 	\caption{Loss Curve: Training vs Testing.}
@@
 
-\subsection{Model Evaluation: Confusion Matrix}
+\subsection{Model Evaluation: Confusion Matrix and Evolution Metrics}
 
-Performance metrics are calculated to ensure a comprehensive evaluation of the model's performance across all classes. We use precision, recall, and F1-score to assess the model:
+To evaluate classification performance, we compute the confusion matrix and derive the following metrics:
 
 \[
-\text{Precision} = \frac{TP}{TP + FP}
-\]
-\[
-\text{Recall} = \frac{TP}{TP + FN}
-\]
-\[
-\text{F1-Score} = \frac{2 \times \text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}
+\text{Precision} = \frac{TP}{TP + FP}, \quad
+\text{Recall} = \frac{TP}{TP + FN}, \quad
+\text{F1-Score} = \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}
 \]
 
-These metrics ensure a balanced evaluation of model performance across all sleep stages.
+Additionally, we compute the evolution matrix \( E \) that captures the transition probabilities between predicted and true classes:
 
-\begin{figure}[h!]
+\[
+E(i, j) = \frac{C(i, j)}{\sum_{k} C(i, k)}
+\]
+
+Where \( C(i, j) \) is the confusion matrix count for true class \( i \) predicted as class \( j \).
+
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/confusion matrix samples.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/confusion_matrix_samples.png}
 	\caption{Confusion Matrix showing model performance across sleep stages.}
@@
 
-\subsection{Gradient Analysis: Training Progression}
+\subsection{Gradient Analysis Across Training}
 
-Understanding model training dynamics:
+Model training stability was assessed via gradient trends:
 
-\textbf{Early Training (Epochs 0-5):} High loss with accuracy starting to improve.  
-\textbf{Mid Training (Epochs 5-15):} Loss steadily decreases with stable gradient flow.  
-\textbf{Late Training (Epochs 15-20):} Accuracy plateaus with no severe overfitting.
+\begin{itemize}
+	\item \textbf{Epochs 0–5:} High loss, gradual accuracy increase.
+	\item \textbf{Epochs 5–15:} Stable gradient flow and decreasing loss.
+	\item \textbf{Epochs 15–20:} Accuracy plateaus, no overfitting.
+\end{itemize}
 
-\textbf{Conclusion:} The training process remains stable, with no vanishing or exploding gradients.
+\textbf{Conclusion:} Training remained stable with no vanishing/exploding gradients.
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/3d gradient.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/3d_gradient.png}
 	\caption{Gradient 3D Surface: Training vs Validation Metrics.}
@@
 
-\subsection{Performance Metrics: Precision, Recall, F1-Score}
+\subsection{Performance Metrics: Class-wise Scores}
 
-Evaluating model performance across all classes using key metrics:
+Key metrics computed per class to understand classification effectiveness:
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/precision_plot.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/precision_plot.png}
 	\caption{Precision Scores per Class.}
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/recall_plot.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/recall_plot.png}
 	\caption{Recall Scores per Class.}
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/f1_score_plot.png}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/f1_score_plot.png}
 	\caption{F1 Scores per Class.}
@@
 
-\subsection{Feature Importance Analysis with LIME}
+\subsection{Feature Importance Analysis using LIME}
 
-Understanding the contribution of different channels to model predictions, we used LIME (Local Interpretable Model-agnostic Explanations) to analyze feature importance.
-The analysis showed that the \textbf{EMG submental} and \textbf{EEG Pz-Oz} channels contribute the most to predictions, while the \textbf{EOG horizontal} channel has minimal importance, indicating lower relevance for classification.
+We used LIME to interpret feature contributions. The results highlight:
 
-This insight helps optimize feature selection and improve model efficiency.
+\begin{itemize}
+	\item \textbf{Most influential:} EMG submental, EEG Pz-Oz
+	\item \textbf{Least influential:} EOG horizontal
+\end{itemize}
 
-\begin{figure}[h!]
+This helps refine feature selection and optimize performance.
+
+\begin{figure}[H]
 	\centering
-    \includegraphics[width=0.8\textwidth]{img/feature importance chanels analysis.png}
-    \caption{Feature Importance Analysis for 4 Channels.}
+	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/feature_importance_chanels_analysis.png}
+	\caption{Feature Importance Analysis for 4 Channels using LIME.}
 \end{figure}

### 2025-05-05 Chapters/introduction.tex
changes

@@
-Sleep is an essential component of human health, contributing significantly to cognitive function, memory consolidation, and emotional regulation.
-Adequate and quality sleep ensures optimal brain performance and supports overall physical and mental well-being.
-Disruptions in sleep patterns can lead to various health issues, including impaired concentration, mood disorders, and chronic conditions such as cardiovascular diseases.
+Sleep is a fundamental aspect of human health that greatly contributes to cognitive function, memory consolidation, and emotional regulation.
+Proper and quality sleep guarantees proper brain function and overall physical and mental health.
+Sleep disturbances result in several diseases, such as impaired concentration, mood disorders, and chronic conditions like cardiovascular diseases.
 
-The classification of sleep stages is fundamental in understanding sleep architecture and diagnosing sleep disorders.
-Experts categorize sleep into five primary stages—N1, N2, N3, REM, and Wake—based on distinct patterns observed in brain activity.
-Each sleep stage is characterized by specific brainwave frequencies, which can be detected using electroencephalography (EEG) signals.
-Accurate classification of these stages is crucial for analyzing sleep quality and identifying abnormalities such as insomnia, sleep apnea, and narcolepsy.
+Sleep plays a critical function in sustaining physical and mental well-being.
+It is not a passive resting state but a sophisticated biological process crucial for memory consolidation, emotional homeostasis, metabolic health, and immune system function.
+Alterations in sleep quality or sleep patterns can reflect or lead to many health problems, such as insomnia, depression, cardiovascular disease, and neurodegenerative disorders.
+It is essential for clinicians and scientists to precisely assess sleep stages for the diagnosis of such conditions and the prescription of proper treatment plans.
 
-Table~\ref{tab:sleep_stages} provides an overview of the frequency ranges associated with each sleep stage.
+Staging of the sleep stages is the basis of understanding the architecture of sleep and the diagnosis of sleep disorders.
+Sleep is divided into five major stages—N1, N2, N3, REM, and Wake—based on unique patterns seen in brain function.
+Every stage of sleep is associated with particular frequencies of brainwaves, which can be picked up using the technique of electroencephalography (EEG) signals.
+The precise staging is important to study the quality of sleep and to detect abnormalities like insomnia, sleep apnea, and narcolepsy.
 
-\begin{table}[h]
+\subsection{The Role of EEG in Sleep Analysis}
+
+Electroencephalography (EEG) is among the major methods of tracking brain activity while asleep.
+EEG captures electrical activity created by neurons as they fire within the brain, recorded with electrodes applied to the scalp.
+These electrical signals are critical in distinguishing various stages of sleep since each stage has different patterns of brain wave activity.
+Whereas other physiological signals are indirect, EEG is a direct view into the electrical activity of the brain and hence the basis for research and diagnosis into sleep.
+
+\subsection{Understanding the SleepEDF Dataset}
+
+The Sleep-EDF (Sleep European Data Format) corpus, popular among researchers, comprises polysomnography (PSG) recordings taken from healthy subjects and mildly disordered sleep patients.
+Recordings are taken non-invasively and usually overnight in a laboratory setting.
+In the course of a session, several sensors are applied to the subject's body to record a number of different physiological signals such as EEG (electroencephalogram), EOG (electrooculogram), and EMG (electromyogram).
+
+Specifically, the database contains two EEG channels, namely Fpz-Cz and Pz-Oz, which record frontal and parietal brain activity.
+Additionally, it contains EOG signals to record eye movements and EMG signals to record muscle activity, particularly around the chin region.
+The recordings are manually annotated by experienced sleep technicians using visual patterns and set guidelines to mark sleep stages.
+The last annotation is retained in a hypnogram — a time series plot that shows the changes between various stages of sleep throughout the night.
+
+\subsection{Sleep Stages and Signal Patterns}
+
+Human sleep is commonly divided into five stages: Wake (W), Non-Rapid Eye Movement (NREM) stages N1, N2, and N3, and Rapid Eye Movement (REM).
+Each stage is characterized by specific frequency patterns in EEG signals:
+
+\begin{table}[H]
 	\centering
-\caption{Frequency ranges of different sleep stages.}
-\begin{tabular}{ll}
+	\caption{Brainwave Types and Their Characteristics in Sleep Staging}
+	\resizebox{\textwidth}{!}{%
+		\begin{tabular}{@{}lll@{}}
 			\toprule
-\textbf{Sleep Stage} & \textbf{Frequency Range (Hz)} \\
-\midrule
-Wake (Beta waves)       & 12--30 \\
-N1 (Light Sleep)        & 4--8  \\
-N2 (Moderate Sleep)     & 4--6  \\
-N3 (Deep Sleep, Delta)  & 0.5--4  \\
-REM (Theta waves)       & 4--6  \\
+			\textbf{Wave Type} & \textbf{Frequency Range} & \textbf{Associated Sleep Stage / Behavior} \\ \midrule
+			Alpha waves        & 8–13 Hz                 & Relaxed wakefulness, especially with eyes closed \\
+			Beta waves         & 13–30 Hz                & Alertness and active thinking; decrease during sleep \\
+			Theta waves        & 4–8 Hz                  & Light sleep (Stages N1 and N2) \\
+			Delta waves        & 0.5–4 Hz                & Deep sleep (Stage N3); synchronized neuronal firing \\
+			Sleep spindles \& K-complexes & ~12–15 Hz (spindles)     & Characteristic of Stage N2; used for stage classification \\ 
 			\bottomrule
-\end{tabular}
-\label{tab:sleep_stages}
+		\end{tabular}%
+	}
+	\label{tab:brainwaves}
 \end{table}
 
-To classify sleep stages effectively, bio-signals from multiple physiological channels are analyzed.
-The primary sources of information for sleep stage classification are EEG (electroencephalogram), EMG (electromyogram), and EOG (electrooculogram) signals.
-These channels capture brain activity, muscle movement, and eye movemets, respectively, offering a comprehensive view of sleep patterns.
+These patterns are extracted from raw EEG signals through filtering and segmentation.
+Experts use these patterns, along with eye movement and muscle tone data, to classify each 30-second segment (epoch) into one of the five stages.
 
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=0.5\linewidth]{img/Sleep EDF Data Visulization.png}
-    \caption{Sleep EDF Data Visulization}
-    \label{fig:enter-label}
-\end{figure}
-
-Despite the availability of multiple channels in the SleepEDF dataset, not all contribute equally to classification performance.
-To reduce model complexity without compromising accuracy, we selectively utilize four channels from the nine standard channels provided in the dataset.
-The selected channels are:
-
-\begin{itemize}
-    \item \textbf{EEG Fpz-Cz}: Captures frontal-to-central brain activity.  
-    \item \textbf{EEG Pz-Oz}: Measures parietal-to-occipital brain activity.  
-    \item \textbf{EMG submental}: Records muscle tone from the chin area.  
-    \item \textbf{EOG horizontal}: Tracks horizontal eye movements.  
-\end{itemize}
-
-In this study, we adopt a graph-based modeling approach to capture spatial relationships between the selected channels.
-Additionally, we employ a Transformer encoder to effectively model temporal dependencies across sleep cycles.
-To facilitate a better understanding of the input data, we also present visualizations of bio-signals from all channels.
-
-This approach aims to leverage the strengths of both spatial and temporal modeling techniques, resulting in a robust and efficient sleep stage classification framework.
+\subsection{The Need for Automation}
+Manual scoring of sleep stages, while precise, is time-consuming and subject to inter-rater reliability.
+It takes hours of professional scoring for one night's worth of recording.
+Automated sleep stage scoring provides a quicker, more scalable, and more possibly consistent solution.
+Using machine learning and deep learning models, particularly those that can address intricate spatial and temporal patterns in physiological data, we can develop systems that emulate expert opinion and aid in clinical decision-making.
+This thesis suggests just such a system — \textit{SleepGCN-Transformer} — that unifies graph-based representation of EEG sensor relationships with temporal learning from transformers.
+Our strategy not only targets high classification performance but also contributes to the increasing body of research in interpretable AI in medicine, moving us closer to clinically integrated, explainable, and fully automated sleep diagnostics.

### 2025-05-05 Chapters/paper1.tex
changes

\section{Methodology}

We begin by importing the Sleep-EDF data provided in EDF (European Data Format). The raw EEG recordings are loaded using MNE's \texttt{read\_raw\_edf} function, and corresponding annotations are aligned using \texttt{read\_annotations}. We optionally exclude all non-EEG channels unless explicitly required, focusing on EEG signals for our analysis. The dataset includes two EEG channels: \texttt{Fpz-Cz} and \texttt{Pz-Oz}, which are the primary input sources in our work.

During loading, we crop the signal to reduce unnecessary wake-time data, retaining 30 minutes before the first sleep stage and 30 minutes after the last. This ensures that our training data remains focused around relevant sleep activity. The channel names are cleaned by stripping prefixes for standardization.

After loading and cropping, we segment the data into fixed-length 30-second epochs using the known sampling frequency of the recordings. Only epochs labeled with valid sleep stages — specifically stages W, N1, N2, N3, and REM — are retained. For each epoch, the raw data is extracted and structured into a format suitable for training.

Once all epochs are collected, we flatten the 3D EEG data (epochs $\times$ channels $\times$ time points) into 2D arrays (epochs $\times$ features). This is necessary for traditional machine learning classifiers. To address class imbalance in sleep stages, we apply SMOTE (Synthetic Minority Oversampling Technique) to generate balanced training samples.

The balanced dataset is then split into training and testing sets in an 80-20 ratio. We train a  Machine Learning   on the training data. After training, predictions are made on the test set, and the model is evaluated using standard metrics including classification report and confusion matrix. The confusion matrix is visualized using a heatmap to give a clear view of the stage-wise performance of the classifier.

\section{Dataset Information}

We have used the Sleep-EDF Dataset available at \url{https://www.physionet.org/content/sleep-edfx/1.0.0/}. For our experiments, we selected recordings from approximately 30 patients, resulting in 60 files — 30 EDF (European Data Format) recordings and their corresponding hypnogram annotation files. The EDF files include various bio-signals captured through multiple channels. The hypnogram files contain sleep stage annotations, specifying the start time and type of each stage (Wake, N1, N2, N3, N4, REM) with timestamps in the HH:MM:SS format.

The EDF recordings consist of a range of physiological signals, including EEG (\texttt{Fpz-Cz} and \texttt{Pz-Oz}), EOG, EMG (submental), rectal temperature, and oro-nasal respiration. Each channel provides continuous signal recordings at a specific sampling frequency (Hz), which are used as input features for our downstream machine learning pipeline. From these signals, we derive two essential inputs: spatial features, used to construct graph-based representations, and temporal features, which capture the sequence and timing dynamics necessary for modeling the sleep process.

\begin{figure}
	\centering
	\includegraphics[width=0.4\linewidth]{"img/paper 1/EEG_Filter_chanel"}
	\caption{Filter EEG Chanels}
	\label{fig:eegfilterchanel}
\end{figure}

\section{Preprocessing}

This study applied machine learning techniques to classify sleep stages using EEG signals. The preprocessing pipeline included several crucial steps: channel mapping, data cropping, signal filtering, epoch creation, label mapping, and class balancing using SMOTE. These steps ensured clean, well-structured input for feature extraction and model training. The extracted features spanned time-domain, frequency-domain, and time-frequency representations. Multiple machine learning models—Gradient Boosting, Random Forest, Support Vector Machine (SVC), and Bagging—were trained and evaluated using metrics such as accuracy, precision, recall, and F1-score with cross-validation to assess generalization. The trained models were finally used to classify unseen EEG data into five sleep stages: Wake, N1, N2, N3, and REM.

\paragraph{A. Dataset Acquisition}

EEG signals were recorded using non-invasive electrodes placed on the scalp of human volunteers, capturing brain electrical activity during overnight sleep studies. Additional demographic data such as age, gender, medical history, and sleep habits were collected to provide contextual understanding. The data collection took place in controlled conditions to minimize noise and interference. Manual inspection of the raw EEG was performed to reject or correct artifacts where necessary, ensuring the quality of the input data. The final recordings were saved in standard EDF format for compatibility with downstream analysis. The dataset used in this study was sourced from publicly available Sleep-EDF recordings, which include EEG data and associated hypnogram annotations reflecting sleep stages.

\paragraph{B. Data Preprocessing}

Preprocessing was applied to prepare EEG data for modeling. Channel mapping was done to identify relevant EEG channels, particularly \texttt{Fpz-Cz} and \texttt{Pz-Oz}, and irrelevant or noisy channels were excluded. Data cropping was applied to isolate the sleep period from wake periods based on annotation markers. The signals were then filtered to remove noise and isolate specific frequency bands important for sleep analysis.

Epochs of 30-second duration were created from continuous EEG signals to structure the data for machine learning. Sleep stage labels corresponding to these epochs were derived from hypnogram annotations. Given the naturally imbalanced class distribution in sleep stage data (e.g., fewer REM and N1 stages), the SMOTE algorithm was employed to oversample the minority classes, thereby enhancing model robustness. After preprocessing, the dataset was split into training and testing subsets. This enabled reliable training of models and unbiased performance evaluation. The output from preprocessing was then used in the feature extraction and modeling stages.

\section{Model Architecture and Learning Framework}

The core modeling pipeline begins with transforming the raw EEG data into structured epochs suitable for machine learning classification. Each EEG recording is segmented into 30-second epochs using the sampling frequency and annotated sleep stages. Only valid sleep stages (Wake, N1, N2, N3, and REM) are retained for downstream processing. After epoching, each segment is extracted using the \texttt{mne.Epochs} function, resulting in structured EEG data with consistent time windows. These segments are then reshaped into two-dimensional feature vectors where each row corresponds to an epoch and each column to a time-series sample across all EEG channels.

To address the class imbalance inherent in sleep stage data, the SMOTE (Synthetic Minority Over-sampling Technique) algorithm is applied. This balances the dataset by synthetically generating new examples in underrepresented classes. The balanced dataset is then split into training and testing sets in an 80:20 ratio.

\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{"img/paper 1/Architechture1"}
	\caption{Architecture of Machine Learning Classifiers for Sleep Stage
		Classificatio}
	\label{fig:architechture1}
\end{figure}

\subsection{Random Forest Classifier}

We implemented a Random Forest classifier to establish a strong baseline model for sleep stage classification. The raw EEG recordings were segmented into 30-second epochs, and features were extracted by flattening the multidimensional epoch data. To handle class imbalance, the SMOTE algorithm was applied prior to training. The model was initialized with \textbf{100 decision trees} (\texttt{n\_estimators=100}) and a fixed \textbf{random seed} (\texttt{random\_state=42}) to ensure reproducibility. This configuration allowed the model to generalize well while maintaining robust performance across varying sleep stages.

\subsection{Gradient Boosting Classifier with PCA}

To improve computational efficiency and potentially boost model accuracy, we introduced Principal Component Analysis (PCA), retaining \textbf{95\% of the explained variance} (\texttt{n\_components=0.95}). The reduced feature set was passed to a \textbf{Gradient Boosting classifier} configured with \textbf{30 estimators} (\texttt{n\_estimators=30}), a \textbf{maximum tree depth of 3} (\texttt{max\_depth=3}), and a \textbf{learning rate of 0.1}. Class balancing was again addressed with SMOTE. The same 80-20 train-test split was maintained with \texttt{random\_state=42}. This model leverages the sequential learning of weak classifiers to optimize classification performance over multiple iterations.

\subsection{Ensemble Learning (Voting Classifier)}

\begin{sloppypar}
	For further enhancement, we employed a soft voting ensemble that combines multiple classifiers, each with complementary strengths. The ensemble includes a \textbf{Gradient Boosting Classifier} (\texttt{n\_estimators=50}, \texttt{max\_depth=3}, \texttt{learning\_rate=0.1}), a \textbf{Random Forest Classifier} (\texttt{n\_estimators=50}, \texttt{max\_depth=3}), a \textbf{Support Vector Classifier (SVC)} with an RBF kernel and probability estimates enabled, and a \textbf{Bagging Classifier} (\texttt{n\_estimators=30}, \texttt{max\_samples=1.0}, \texttt{max\_features=1.0}). These classifiers were integrated using a \textbf{soft voting strategy} in the \texttt{VotingClassifier(voting='soft')} to average their probabilistic predictions.

	Feature reduction was again performed using PCA with 95\% variance retention. This ensemble approach leverages model diversity to achieve improved generalization and more stable predictions across sleep classes.
\end{sloppypar}

\section{Results and Evaluation}

This section presents the results and evaluations for the models trained, including Random Forest, Bagging, Ensemble Learning, and Gradient Boosting. The following figures summarize the performance metrics for each model.

\subsection{Random Forest}

For the Random Forest model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/random_forest_confusion_matrix.png}
	\caption{Random Forest Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/accuracy_curve.png}
	\caption{Random Forest Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/loss_curve.png}
	\caption{Random Forest Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/precision_per_class.png}
	\caption{Random Forest Precision Per Class}
\end{figure}

\subsection{Bagging Classifier}

For the Bagging Classifier model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/BG_confusion_matrix.png}
	\caption{Bagging Classifier Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/accuracy_curve.png}
	\caption{Bagging Classifier Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/loss_curve.png}
	\caption{Bagging Classifier Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/precision_per_class.png}
	\caption{Bagging Classifier Precision Per Class}
\end{figure}

\subsection{Ensemble Learning}

For the Ensemble Learning model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/Ensemble_confusion_matrix.png}
	\caption{Ensemble Learning Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/accuracy_curve.png}
	\caption{Ensemble Learning Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/loss_curve.png}
	\caption{Ensemble Learning Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/precision_per_class.png}
	\caption{Ensemble Learning Precision Per Class}
\end{figure}

\subsection{Gradient Boosting}

For the Gradient Boosting model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/confusion_matrix.png}
	\caption{Gradient Boosting Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/accuracy_curve.png}
	\caption{Gradient Boosting Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/loss_curve.png}
	\caption{Gradient Boosting Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/precision_per_class.png}
	\caption{Gradient Boosting Precision Per Class}
\end{figure}

### 2025-05-05 Chapters/paper3.tex
changes

--- 2025-05-05 Chapters/paper1.tex
@@
 \section{Methodology}
-
-We begin by importing the Sleep-EDF data provided in EDF (European Data Format).
-The raw EEG recordings are loaded using MNE's \texttt{read\_raw\_edf} function, and corresponding annotations are aligned using \texttt{read\_annotations}.
-We optionally exclude all non-EEG channels unless explicitly required, focusing on EEG signals for our analysis.
-The dataset includes two EEG channels: \texttt{Fpz-Cz} and \texttt{Pz-Oz}, which are the primary input sources in our work.
-
-During loading, we crop the signal to reduce unnecessary wake-time data, retaining 30 minutes before the first sleep stage and 30 minutes after the last.
-This ensures that our training data remains focused around relevant sleep activity.
-The channel names are cleaned by stripping prefixes for standardization.
-
-After loading and cropping, we segment the data into fixed-length 30-second epochs using the known sampling frequency of the recordings.
-Only epochs labeled with valid sleep stages — specifically stages W, N1, N2, N3, and REM — are retained.
-For each epoch, the raw data is extracted and structured into a format suitable for training.
-
-Once all epochs are collected, we flatten the 3D EEG data (epochs $\times$ channels $\times$ time points) into 2D arrays (epochs $\times$ features).
-This is necessary for traditional machine learning classifiers.
-To address class imbalance in sleep stages, we apply SMOTE (Synthetic Minority Oversampling Technique) to generate balanced training samples.
-
-The balanced dataset is then split into training and testing sets in an 80-20 ratio.
-We train a  Machine Learning   on the training data.
-After training, predictions are made on the test set, and the model is evaluated using standard metrics including classification report and confusion matrix.
-The confusion matrix is visualized using a heatmap to give a clear view of the stage-wise performance of the classifier.
 
@@
 
-We have used the Sleep-EDF Dataset available at \url{https://www.physionet.org/content/sleep-edfx/1.0.0/}.
-For our experiments, we selected recordings from approximately 30 patients, resulting in 60 files — 30 EDF (European Data Format) recordings and their corresponding hypnogram annotation files.
-The EDF files include various bio-signals captured through multiple channels.
-The hypnogram files contain sleep stage annotations, specifying the start time and type of each stage (Wake, N1, N2, N3, N4, REM) with timestamps in the HH:MM:SS format.
-
-The EDF recordings consist of a range of physiological signals, including EEG (\texttt{Fpz-Cz} and \texttt{Pz-Oz}), EOG, EMG (submental), rectal temperature, and oro-nasal respiration.
-Each channel provides continuous signal recordings at a specific sampling frequency (Hz), which are used as input features for our downstream machine learning pipeline.
-From these signals, we derive two essential inputs: spatial features, used to construct graph-based representations, and temporal features, which capture the sequence and timing dynamics necessary for modeling the sleep process.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.4\linewidth]{"img/paper 1/EEG_Filter_chanel"}
-	\caption{Filter EEG Chanels}
-	\label{fig:eegfilterchanel}
-\end{figure}
-
-\section{Preprocessing}
-
-This study applied machine learning techniques to classify sleep stages using EEG signals.
-The preprocessing pipeline included several crucial steps: channel mapping, data cropping, signal filtering, epoch creation, label mapping, and class balancing using SMOTE.
-These steps ensured clean, well-structured input for feature extraction and model training.
-The extracted features spanned time-domain, frequency-domain, and time-frequency representations.
-Multiple machine learning models—Gradient Boosting, Random Forest, Support Vector Machine (SVC), and Bagging—were trained and evaluated using metrics such as accuracy, precision, recall, and F1-score with cross-validation to assess generalization.
-The trained models were finally used to classify unseen EEG data into five sleep stages: Wake, N1, N2, N3, and REM.
-
-\paragraph{A. Dataset Acquisition}
-
-EEG signals were recorded using non-invasive electrodes placed on the scalp of human volunteers, capturing brain electrical activity during overnight sleep studies.
-Additional demographic data such as age, gender, medical history, and sleep habits were collected to provide contextual understanding.
-The data collection took place in controlled conditions to minimize noise and interference.
-Manual inspection of the raw EEG was performed to reject or correct artifacts where necessary, ensuring the quality of the input data.
-The final recordings were saved in standard EDF format for compatibility with downstream analysis.
-The dataset used in this study was sourced from publicly available Sleep-EDF recordings, which include EEG data and associated hypnogram annotations reflecting sleep stages.
-
-\paragraph{B. Data Preprocessing}
-
-Preprocessing was applied to prepare EEG data for modeling.
-Channel mapping was done to identify relevant EEG channels, particularly \texttt{Fpz-Cz} and \texttt{Pz-Oz}, and irrelevant or noisy channels were excluded.
-Data cropping was applied to isolate the sleep period from wake periods based on annotation markers.
-The signals were then filtered to remove noise and isolate specific frequency bands important for sleep analysis.
-
-Epochs of 30-second duration were created from continuous EEG signals to structure the data for machine learning.
-Sleep stage labels corresponding to these epochs were derived from hypnogram annotations.
-Given the naturally imbalanced class distribution in sleep stage data (e.g., fewer REM and N1 stages), the SMOTE algorithm was employed to oversample the minority classes, thereby enhancing model robustness.
-After preprocessing, the dataset was split into training and testing subsets.
-This enabled reliable training of models and unbiased performance evaluation.
-The output from preprocessing was then used in the feature extraction and modeling stages.
+\section{Preprocessing Techniques}
 
@@
 
-The core modeling pipeline begins with transforming the raw EEG data into structured epochs suitable for machine learning classification.
-Each EEG recording is segmented into 30-second epochs using the sampling frequency and annotated sleep stages.
-Only valid sleep stages (Wake, N1, N2, N3, and REM) are retained for downstream processing.
-After epoching, each segment is extracted using the \texttt{mne.Epochs} function, resulting in structured EEG data with consistent time windows.
-These segments are then reshaped into two-dimensional feature vectors where each row corresponds to an epoch and each column to a time-series sample across all EEG channels.
-
-To address the class imbalance inherent in sleep stage data, the SMOTE (Synthetic Minority Over-sampling Technique) algorithm is applied.
-This balances the dataset by synthetically generating new examples in underrepresented classes.
-The balanced dataset is then split into training and testing sets in an 80:20 ratio.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.7\linewidth]{"img/paper 1/Architechture1"}
-	\caption{Architecture of Machine Learning Classifiers for Sleep Stage
-		Classificatio}
-	\label{fig:architechture1}
-\end{figure}
-
-\subsection{Random Forest Classifier}
-
-We implemented a Random Forest classifier to establish a strong baseline model for sleep stage classification.
-The raw EEG recordings were segmented into 30-second epochs, and features were extracted by flattening the multidimensional epoch data.
-To handle class imbalance, the SMOTE algorithm was applied prior to training.
-The model was initialized with \textbf{100 decision trees} (\texttt{n\_estimators=100}) and a fixed \textbf{random seed} (\texttt{random\_state=42}) to ensure reproducibility.
-This configuration allowed the model to generalize well while maintaining robust performance across varying sleep stages.
-
-\subsection{Gradient Boosting Classifier with PCA}
-
-To improve computational efficiency and potentially boost model accuracy, we introduced Principal Component Analysis (PCA), retaining \textbf{95\% of the explained variance} (\texttt{n\_components=0.95}).
-The reduced feature set was passed to a \textbf{Gradient Boosting classifier} configured with \textbf{30 estimators} (\texttt{n\_estimators=30}), a \textbf{maximum tree depth of 3} (\texttt{max\_depth=3}), and a \textbf{learning rate of 0.1}.
-Class balancing was again addressed with SMOTE.
-The same 80-20 train-test split was maintained with \texttt{random\_state=42}.
-This model leverages the sequential learning of weak classifiers to optimize classification performance over multiple iterations.
-
-\subsection{Ensemble Learning (Voting Classifier)}
-
-\begin{sloppypar}
-	For further enhancement, we employed a soft voting ensemble that combines multiple classifiers, each with complementary strengths.
-The ensemble includes a \textbf{Gradient Boosting Classifier} (\texttt{n\_estimators=50}, \texttt{max\_depth=3}, \texttt{learning\_rate=0.1}), a \textbf{Random Forest Classifier} (\texttt{n\_estimators=50}, \texttt{max\_depth=3}), a \textbf{Support Vector Classifier (SVC)} with an RBF kernel and probability estimates enabled, and a \textbf{Bagging Classifier} (\texttt{n\_estimators=30}, \texttt{max\_samples=1.0}, \texttt{max\_features=1.0}).
-These classifiers were integrated using a \textbf{soft voting strategy} in the \texttt{VotingClassifier(voting='soft')} to average their probabilistic predictions.
-
-	Feature reduction was again performed using PCA with 95\% variance retention.
-This ensemble approach leverages model diversity to achieve improved generalization and more stable predictions across sleep classes.
-\end{sloppypar}
-
 \section{Results and Evaluation}
-
-This section presents the results and evaluations for the models trained, including Random Forest, Bagging, Ensemble Learning, and Gradient Boosting.
-The following figures summarize the performance metrics for each model.
-
-\subsection{Random Forest}
-
-For the Random Forest model, we present the following evaluation metrics:
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/random_forest_confusion_matrix.png}
-	\caption{Random Forest Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/accuracy_curve.png}
-	\caption{Random Forest Accuracy Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/loss_curve.png}
-	\caption{Random Forest Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/precision_per_class.png}
-	\caption{Random Forest Precision Per Class}
-\end{figure}
-
-\subsection{Bagging Classifier}
-
-For the Bagging Classifier model, we present the following evaluation metrics:
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/BG_confusion_matrix.png}
-	\caption{Bagging Classifier Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/accuracy_curve.png}
-	\caption{Bagging Classifier Accuracy Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/loss_curve.png}
-	\caption{Bagging Classifier Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/precision_per_class.png}
-	\caption{Bagging Classifier Precision Per Class}
-\end{figure}
-
-\subsection{Ensemble Learning}
-
-For the Ensemble Learning model, we present the following evaluation metrics:
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/Ensemble_confusion_matrix.png}
-	\caption{Ensemble Learning Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/accuracy_curve.png}
-	\caption{Ensemble Learning Accuracy Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/loss_curve.png}
-	\caption{Ensemble Learning Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/precision_per_class.png}
-	\caption{Ensemble Learning Precision Per Class}
-\end{figure}
-
-\subsection{Gradient Boosting}
-
-For the Gradient Boosting model, we present the following evaluation metrics:
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/confusion_matrix.png}
-	\caption{Gradient Boosting Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/accuracy_curve.png}
-	\caption{Gradient Boosting Accuracy Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/loss_curve.png}
-	\caption{Gradient Boosting Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/precision_per_class.png}
-	\caption{Gradient Boosting Precision Per Class}
-\end{figure}

### 2025-05-05 Title/abstract.tex
changes

@@
 \end{center}
-This project presents \textit{SleepGCN-Transformer}, a hybrid model combining Graph Convolutional Networks (GCN) and Transformer encoders for sleep stage classification.
-Using the SleepEDF dataset, it incorporates four physiological signals: EEG (Fpz-Cz, Pz-Oz), EMG (submental), and EOG (horizontal).
-Preprocessing includes 30-second epochs, band-pass filtering (0.3--30 Hz), and graph-based EEG channel representation.
 
-The GCN module captures spatial relationships across EEG channels, while the Transformer encoder models temporal dependencies from graph-level embeddings.
-Focal Loss addresses class imbalances, and a CosineAnnealingLR scheduler optimizes learning rate decay.
-Training with the AdamW optimizer for 20 epochs achieves 93.12 \% training and 93.04\% validation accuracy.
+Correct and effective sleep stage classification is crucial in the diagnosis of sleep disorders, conventionally done through manual interpretation of polysomnography (PSG) signals.
+Automated sleep stage classification is investigated in this thesis by a series of increasingly sophisticated models, from traditional machine learning to sophisticated deep learning architectures.
+Initial trials on the SleepEDF dataset compared models like Random Forest, XGBoost, and ensemble models, which obtained maximum accuracy of 85.35\%, and set the foundation for automated systems to perform consumer-level sleep monitoring.
+Future work introduced deep learning models like BiLSTM and RNNs, and BiLSTM reached 81.13\% accuracy, which proved to have better ability to model temporal sequences.
 
-Model performance, assessed using precision, recall, and F1-score, demonstrates high efficacy.
-LIME-based feature importance analysis highlights EMG and EEG Pz-Oz channels as key contributors.
-This hybrid model exhibits state-of-the-art performance, with future directions focused on enhancing explainability and clinical integration.
+Based on these building blocks, we introduce \textit{SleepGCN-Transformer}, a new hybrid model that combines Graph Convolutional Networks (GCNs) to learn spatial relations between multi-channel EEG, EMG, and EOG signals, and Transformer encoders to learn temporal patterns.
+The model applies Focal Loss to handle class imbalance and a CosineAnnealingLR scheduler for dynamic learning rate adaptation.
+Trained with the AdamW optimizer on preprocessed 30-second epochs, SleepGCN-Transformer is 93.12\% training and 93.04\% validation accurate.
+Feature attribution by LIME indicates EMG and EEG Pz-Oz to be the most significant channels, demonstrating the model's interpretability.

### 2025-05-05 Title/title.tex
changes

@@
 \begin{center}
-    {{\Large \textbf{SleepGCN-Transformer: A Hybrid Graph Convolutional and
-Transformer Network for Sleep Stage Classification
-}}}\\
+	{{\Large \textbf{SleepGCN-Transformer: A Hybrid Graph Convolutional and\\
+				Transformer Network for Sleep Stage Classification}}}\\
 	\vspace{0.6cm}
-    {\large A Master's Report}\\
+
+	{\large A Master's Thesis}\\
 	{\large Submitted in partial fulfillment of the requirement for the award of the degree of}\\
 	\vspace{0.8cm}
+
 	{\large Master of Technology in Artificial Intelligence}\\
 	{\large by}\\
+
 	\textbf{{\Large Tanmay Rathod}}\\
@@
 	\vspace{0.7cm}
+
 	{\large Under the guidance of}\\
-    {\large  \textbf{Dr. Santosh Kumar Satapathy}  {\large }}\\
+	{\large \textbf{Dr. Santosh Kumar Satapathy}}\\
+
 	{\large Department of Information and Communication Technology}\\
 	\vspace{0.6cm}
+
 	\vfill
 	{\centering \includegraphics[width=0.21\textwidth]{img/pdpu_logo.png}}\\
+
 	{\large School of Technology}\\
@@
 	{\large Gandhinagar – 382426. Gujarat - India}\\
-    {\large February, 2025}
+	{\large May, 2025}
 \end{center}

### 2025-05-05 Chapters/paper1.tex
changes

@@
 	\centering
-	\includegraphics[width=0.4\linewidth]{"img/paper 1/EEG_Filter_chanel"}
+	\includegraphics[width=0.4\linewidth]{"./img/paper_1/EEG_Filter_chanel.png"}
 	\caption{Filter EEG Chanels}
@@
 	\centering
-	\includegraphics[width=0.7\linewidth]{"img/paper 1/Architechture1"}
+	\includegraphics[width=0.7\linewidth]{"img/paper_1/Architechture1"}
 	\caption{Architecture of Machine Learning Classifiers for Sleep Stage
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/random_forest_confusion_matrix.png}
+	\includegraphics[width=\textwidth]{img/paper_1/random_forest_confusion_matrix.png}
 	\caption{Random Forest Confusion Matrix}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/accuracy_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1/accuracy_curve.png}
 	\caption{Random Forest Accuracy Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/loss_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1/loss_curve.png}
 	\caption{Random Forest Loss Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/precision_per_class.png}
+	\includegraphics[width=\textwidth]{img/paper_1/precision_per_class.png}
 	\caption{Random Forest Precision Per Class}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/BG_confusion_matrix.png}
+	\includegraphics[width=\textwidth]{img/paper_1/BG_confusion_matrix.png}
 	\caption{Bagging Classifier Confusion Matrix}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/accuracy_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1/accuracy_curve.png}
 	\caption{Bagging Classifier Accuracy Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/loss_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1/loss_curve.png}
 	\caption{Bagging Classifier Loss Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/precision_per_class.png}
+	\includegraphics[width=\textwidth]{img/paper_1/precision_per_class.png}
 	\caption{Bagging Classifier Precision Per Class}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/Ensemble_confusion_matrix.png}
+	\includegraphics[width=\textwidth]{img/paper_1//Ensemble_confusion_matrix.png}
 	\caption{Ensemble Learning Confusion Matrix}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/accuracy_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1//accuracy_curve.png}
 	\caption{Ensemble Learning Accuracy Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/loss_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1//loss_curve.png}
 	\caption{Ensemble Learning Loss Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/precision_per_class.png}
+	\includegraphics[width=\textwidth]{img/paper_1//precision_per_class.png}
 	\caption{Ensemble Learning Precision Per Class}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/confusion_matrix.png}
+	\includegraphics[width=\textwidth]{img/paper_1//confusion_matrix.png}
 	\caption{Gradient Boosting Confusion Matrix}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/accuracy_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1//accuracy_curve.png}
 	\caption{Gradient Boosting Accuracy Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/loss_curve.png}
+	\includegraphics[width=\textwidth]{img/paper_1//loss_curve.png}
 	\caption{Gradient Boosting Loss Curve}
@@
 	\centering
-	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/precision_per_class.png}
+	\includegraphics[width=\textwidth]{img/paper_1//precision_per_class.png}
 	\caption{Gradient Boosting Precision Per Class}

### 2025-05-05 Chapters/paper1.tex
changes

@@
 	\centering
-	\includegraphics[width=0.7\linewidth]{"img/paper_1/Architechture1"}
+	\includegraphics[width=0.9\linewidth]{"img/paper_1/Architechture1"}
 	\caption{Architecture of Machine Learning Classifiers for Sleep Stage
@@
 \end{figure}
+
+\section{K-Fold Cross-Validation}
+
+K-Fold Cross Validation  5-fold cross-validation approach was employed to evaluate the model's performance.
+The 5-fold cross-validation procedure divides the dataset into 5 equal subsets (or folds).
+In each iteration, the model is trained on 80\% of the data (four folds) and tested on the remaining 20\% (the fifth fold).
+This process is repeated 5 times, with each fold serving as the test set once, ensuring that each data point is tested exactly once.
+
+This approach provides a more reliable estimate of model performance by reducing the potential bias that can arise from using a single train-test split.
+The final performance of the model is obtained by averaging the results from each fold, offering a comprehensive evaluation of its ability to generalize to unseen data.
+
+The advantage of K-fold cross-validation lies in its ability to provide a more robust and generalized estimate of model performance by using different subsets for training and testing in each iteration.
+This process reduces the risk of overfitting and ensures that the model's evaluation is based on multiple different test sets.
 
@@
 	\centering
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
 		\includegraphics[width=\textwidth]{img/paper_1/random_forest_confusion_matrix.png}
 		\caption{Random Forest Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Random Forest Accuracy Curve}
-\end{figure}
+	\end{minipage}
 
-\begin{figure}[H]
+	\vspace{1em}
+
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Random Forest Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Random Forest Precision Per Class}
+	\end{minipage}
 \end{figure}
@@
 	\centering
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
 		\includegraphics[width=\textwidth]{img/paper_1/BG_confusion_matrix.png}
 		\caption{Bagging Classifier Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Bagging Classifier Accuracy Curve}
-\end{figure}
+	\end{minipage}
 
-\begin{figure}[H]
+	\vspace{1em}
+
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Bagging Classifier Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Bagging Classifier Precision Per Class}
+	\end{minipage}
 \end{figure}
@@
 	\centering
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
 		\includegraphics[width=\textwidth]{img/paper_1//Ensemble_confusion_matrix.png}
 		\caption{Ensemble Learning Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Ensemble Learning Accuracy Curve}
-\end{figure}
+	\end{minipage}
 
-\begin{figure}[H]
+	\vspace{1em}
+
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Ensemble Learning Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Ensemble Learning Precision Per Class}
+	\end{minipage}
 \end{figure}
@@
 	\centering
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
 		\includegraphics[width=\textwidth]{img/paper_1//confusion_matrix.png}
 		\caption{Gradient Boosting Confusion Matrix}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Gradient Boosting Accuracy Curve}
-\end{figure}
+	\end{minipage}
 
-\begin{figure}[H]
+	\vspace{1em}
+
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Gradient Boosting Loss Curve}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
 		\centering
@@
 		\caption{Gradient Boosting Precision Per Class}
+	\end{minipage}
 \end{figure}

### 2025-05-05 Chapters/paper1.tex
changes

@@
 
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.6\linewidth]{"img/paper_1/k-FOLD.PNG"}
+	\caption{Architecture of Machine Learning Classifiers for Sleep Stage
+		Classificatio}
+	\label{fig:architechture1}
+\end{figure}
+
 This approach provides a more reliable estimate of model performance by reducing the potential bias that can arise from using a single train-test split.
@@
 This process reduces the risk of overfitting and ensures that the model's evaluation is based on multiple different test sets.
+
+\subsection{Random Forest with K-Fold Cross-Validation Pipeline}
+
+The model pipeline was created using \texttt{make\_pipeline} from \texttt{sklearn.pipeline}, which integrates two main components: a function transformer and a random forest classifier.
+The function transformer is applied to extract the EEG power bands from the data, which are then used as features for training the model.
+The \texttt{RandomForestClassifier} was chosen for its ability to perform well with high-dimensional data, and it was set with 100 estimators to ensure robust performance.
+The random forest classifier was trained and tested on each fold using the 5-fold cross-validation setup, providing an unbiased evaluation of the model’s performance.
+
+\subsection{Ensemble Classifier with K-Fold Cross-Validation Pipeline}
+
+The pipeline was constructed using \texttt{make\_pipeline} from \texttt{sklearn.pipeline}, which integrates a function transformer for extracting EEG power bands as features and an ensemble classifier for prediction.
+The function transformer applies the \texttt{eeg\_power\_band} function to preprocess the data, and the \texttt{VotingClassifier} combines the predictions from a \texttt{RandomForestClassifier} and  The ensemble classifier was trained and tested on each fold using the 5-fold cross-validation setup, offering a more robust evaluation by leveraging the strengths of both classifiers.
+
+\subsection{Gradient Boosting with K-Fold Cross-Validation Pipeline}
+In this section, we employed Gradient Boosting with 5-fold cross-validation to classify the EEG data.
+Gradient Boosting is an ensemble learning technique that builds models sequentially, where each model attempts to correct the errors made by the previous ones.
+This method can improve predictive performance by focusing on difficult-to-predict instances.
+The dataset was split into five subsets, and each subset was used as the test set once, while the remaining four were used for training.
+The model's generalization ability was thus thoroughly evaluated.
+
+A pipeline was constructed using \texttt{make\_pipeline} from \texttt{sklearn.pipeline}.
+It incorporates a function transformer that preprocesses the EEG data by extracting power band features through the \texttt{eeg\_power\_band} function.
+The processed data is then passed to a \texttt{GradientBoostingClassifier}, which is used for prediction.
+The \texttt{KFold} cross-validation technique with 5 splits ensures that the model is trained and validated on different portions of the dataset, offering a reliable estimate of its performance.
 
@@
 
-\section{Results and Evaluation}
-
-This section presents the results and evaluations for the models trained, including Random Forest, Bagging, Ensemble Learning, and Gradient Boosting.
-The following figures summarize the performance metrics for each model.
+\section{Results and Evaluation With Machine Learning }
 
@@
 \end{figure}
+
+\section{Results and Evaluation with K-fold Cross Validation }
+
+\begin{figure}[H]
+	\centering
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/Ensemble Learning_percentage_based.png}
+		\caption{Ensemble Learning Confusion Matrix }
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/Random Forest_percentage_based.png}
+		\caption{Random Forest Confusion Matrix }
+	\end{minipage}
+
+	\vspace{1em}
+
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/GB Classifier_percentage_based.png}
+		\caption{GB Classifier Confusion Matrix }
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/SGD_percentage_based.png}
+		\caption{SGD Confusion Matrix }
+	\end{minipage}
+\end{figure}
+
+\section{Comparison Sections}
+
+\begin{figure}[H]
+	\centering
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/accuracy_comparison.png}
+		\caption{Accuracy Comparison}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/bar_chart_improved.png}
+		\caption{Improved Bar Chart}
+	\end{minipage}
+
+	\vspace{1em}
+
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/box_plot_comparison.png}
+		\caption{Box Plot Comparison}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{img/paper_1/violin_plot_improved.png}
+		\caption{Improved Violin Plot}
+	\end{minipage}
+\end{figure}

### 2025-05-05 Chapters/paper2.tex
changes

\section{Methodology}

We employed the publicly available \textbf{Sleep-EDF} dataset for automated sleep stage classification. This dataset consists of two primary file types: Polysomnographic (PSG) recordings in European Data Format (EDF) and corresponding Hypnogram annotation files. The EDF files contain multi-channel biosignals, including EEG, while the Hypnogram files provide ground-truth sleep stage labels aligned with specific time intervals.

The dataset was partitioned into \textbf{training}, \textbf{validation}, and \textbf{testing} subsets using a \textbf{90:5:5} split. Given the inherent class imbalance in sleep stage distribution, we employed the \textbf{SMOTE} (Synthetic Minority Over-sampling Technique) to synthetically balance the dataset, improving model robustness across minority classes.

\subsection{Modeling Approach}

We implemented and evaluated several machine learning and deep learning architectures for the classification task. One of the core models used is the \textbf{Deep Neural Network (DNN)}, designed as a fully connected feedforward architecture suitable for non-sequential input.

\subsubsection{Deep Neural Network (DNN)}

The DNN consists of an input layer followed by three fully connected (\texttt{Dense}) layers comprising \textbf{128}, \textbf{64}, and \textbf{32} neurons respectively. Each layer employs the \textbf{ReLU} activation function to introduce non-linearity. To mitigate overfitting, \textbf{Dropout layers} with a dropout rate of \textbf{0.2} were interleaved between each dense layer. The final output layer uses a \textbf{Softmax} activation to produce class probabilities for the five sleep stages.

\section{Dataset Information}

 ensure efficient implementation and reproducibility, this research adheres to a well-structured computational framework, including consistent system configuration, version control, and the integration of essential libraries for data processing and deep learning tasks. The experiments were conducted on both Windows and Linux operating systems using an \textbf{Intel Core i5 11th Gen} processor (with \textit{Iris Xe Graphics} and a clock speed of \textbf{3.9 GHz}). The programming environment was built with \textbf{Python version 3.13.1}, managed via the \texttt{pip} package installer.

The study utilized the \textbf{Sleep-EDF} dataset, which contains EEG recordings and corresponding hypnograms for sleep stage classification. Prior to model training, the dataset was preprocessed into uniform, fixed-length segments. The dataset is publicly available at: \url{https://www.physionet.org/content/sleep-edf/1.0.0/}.

For project collaboration and version control, \textbf{GitHub} was employed, with the full implementation accessible at: \url{https://github.com/tanmay007thor/ContraWR}.

Several key libraries were used throughout the development pipeline:
\begin{itemize}
	\item \textbf{NumPy (2.2.2)} – Numerical computations
	\item \textbf{scikit-learn (1.6.1)} – Machine learning utilities
	\item \textbf{imbalanced-learn (0.13.0)} – Handling class imbalance (SMOTE)
	\item \textbf{MNE (1.9.0)} – EEG signal processing
	\item \textbf{TensorFlow/Keras (2.16.1)} – Deep learning model implementation
	\item \textbf{Matplotlib (3.10.0)} and \textbf{Seaborn (0.13.2)} – Data visualization
\end{itemize}

This setup ensures smooth execution of machine learning and deep learning models, including BiLSTM and other architectures, for automating sleep stage classification.

\section{Preprocessing Techniques}

To ensure consistency in the input data, all recordings were resampled to a \textbf{100 Hz} sampling frequency. Each signal was segmented into \textbf{30-second epochs}, resulting in \textbf{3000 time steps per sample}. These samples were stored in \texttt{PKL} (Pickle) format, each approximately \textbf{48 KB} in size, enabling efficient storage and fast access, particularly beneficial for edge device deployment.

Each sample consisted of input-output pairs: the inputs ($X$) included two EEG channels—\textbf{FPZ-CZ} and \textbf{PZ-OZ}—while the labels ($Y$) represented the corresponding sleep stages. The stages were categorized into five classes: \textbf{Wake, N1, N2, N3,} and \textbf{REM}. To streamline processing, raw labels were mapped to a standardized class format.

\section{Model Architecture and Learning Framework}

\subsubsection{Deep Neural Network (DNN)}

A Deep Neural Network (DNN) is a type of multilayer feedforward architecture composed of successive fully connected layers. In this model, the input layer is followed by three dense layers containing 128, 64, and 32 neurons respectively. Each dense layer uses the Rectified Linear Unit (ReLU) activation function to introduce non-linearity.

To improve generalization and prevent overfitting, dropout layers with a dropout rate of 0.2 are inserted between the dense layers. The final output layer employs a Softmax activation function to map the learned features to the five target sleep stages: Wake, N1, N2, N3, and REM.

\begin{table}[H]
	\centering
	\caption{Architecture of the Deep Neural Network}
	\label{tab:nn_architecture}
	\begin{tabular}{lcc}
		\hline
		\textbf{Layer Type} & \textbf{Units} & \textbf{Activation Function} \\
		\hline
		Input Layer         & \texttt{X\_train}          & - \\
		Dense Layer         & 128                        & ReLU \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		Dense Layer         & 64                         & ReLU \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		Dense Layer         & 32                         & ReLU \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		Output Layer        & Number of classes (e.g., 5) & Softmax \\
		\hline
	\end{tabular}
\end{table}

While DNNs are effective in extracting complex feature representations from EEG signals, they are inherently limited in modeling temporal dependencies due to their feedforward nature. This limits their standalone utility for time-series classification tasks like sleep staging.

\subsubsection{Forward Propagation}

Forward propagation computes the activations at each layer using the following equations:

\begin{align}
	z^{(l)} &= W^{(l)} a^{(l-1)} + b^{(l)} \\
	a^{(l)} &= \sigma(z^{(l)})
\end{align}

\noindent where:
\begin{itemize}
	\item \( z^{(l)} \) is the pre-activation (linear combination of inputs) at layer \( l \),
	\item \( a^{(l)} \) is the activation output at layer \( l \),
	\item \( W^{(l)} \) and \( b^{(l)} \) denote the weights and biases of layer \( l \),
	\item \( \sigma \) is the activation function (ReLU or Softmax depending on the layer).
\end{itemize}

\subsubsection{Backward Propagation}

During training, gradients of the loss function with respect to weights and biases are calculated using backpropagation. The process involves the following steps:

\begin{align}
	\delta^{(L)} &= \frac{\partial L}{\partial a^{(L)}} \odot \sigma'(z^{(L)}) \\
	\delta^{(l)} &= \left(W^{(l+1)}\right)^T \delta^{(l+1)} \odot \sigma'(z^{(l)}) \\
	\frac{\partial L}{\partial W^{(l)}} &= \delta^{(l)} \left(a^{(l-1)}\right)^T \\
	\frac{\partial L}{\partial b^{(l)}} &= \sum \delta^{(l)}
\end{align}

\noindent where:
\begin{itemize}
	\item \( \delta^{(l)} \) is the error signal at layer \( l \),
	\item \( \sigma'(z^{(l)}) \) is the derivative of the activation function,
	\item \( \frac{\partial L}{\partial W^{(l)}} \) and \( \frac{\partial L}{\partial b^{(l)}} \) are gradients used for weight and bias updates.
\end{itemize}

\subsubsection{Activation Functions}

The Softmax function is used in the output layer to produce class probabilities:

\begin{equation}
	\sigma(z_i) = \frac{e^{z_i}}{\sum_{j=1}^{N} e^{z_j}}
\end{equation}

\noindent where:
\begin{itemize}
	\item \( z_i \) is the input to the \( i^{th} \) output neuron,
	\item \( N \) is the number of output classes,
	\item The denominator ensures the outputs form a probability distribution summing to 1.
\end{itemize}

For hidden layers, the ReLU activation function is applied:

\begin{equation}
	f(x) = \max(0, x)
\end{equation}

\noindent where:
\begin{itemize}
	\item \( x \) is the input to the neuron,
	\item The function outputs \( x \) if \( x > 0 \), and 0 otherwise.
\end{itemize}

\subsubsection{Recurrent Neural Network (RNN)}

Recurrent Neural Networks (RNNs) are specifically designed for modeling sequential data by incorporating feedback connections, allowing them to retain information across time steps. This architecture is well-suited for EEG signal analysis due to its ability to capture temporal dependencies.

In the proposed model, two RNN layers are employed: the first with 128 units and the second with 64 units, both using the Tanh activation function. These layers process sequential inputs while maintaining internal memory of past computations.

\begin{table}[!h]
	\centering
	\caption{Recurrent Neural Network Architecture for Sleep Stage Classification}
	\label{tab:rnn_architecture}
	\begin{tabular}{lcc}
		\hline
		\textbf{Layer Type} & \textbf{Units} & \textbf{Activation Function} \\
		\hline
		Input Layer         & \texttt{X\_train}          & - \\
		RNN Layer 1         & 128                        & Tanh \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		RNN Layer 2         & 64                         & Tanh \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		Dense Layer         & 64                         & ReLU \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		Dense Layer         & 32                         & ReLU \\
		Dropout Layer       & -                          & 0.2 (Dropout Rate) \\
		Output Layer        & Number of classes (e.g., 5) & Softmax \\
		\hline
	\end{tabular}
\end{table}

Dropout layers with a 0.2 rate are introduced after each RNN and dense layer to mitigate overfitting. The dense layers, activated using ReLU, help to refine temporal features extracted by the RNN layers. The final classification is performed by a Softmax-activated output layer corresponding to the target sleep stages.

Although RNNs are effective for learning short-term dependencies in sequential EEG data, they are prone to issues such as vanishing gradients, which can hinder the learning of long-range patterns.

The operation of an RNN cell at each time step \( t \) is mathematically represented as:

\begin{equation}
	h_t = \tanh(W_h h_{t-1} + W_x x_t + b_h)
\end{equation}

\noindent where:
\begin{itemize}
	\item \( h_t \) is the hidden state at time step \( t \),
	\item \( x_t \) is the input at time step \( t \),
	\item \( W_h \) and \( W_x \) are the recurrent and input weight matrices, respectively,
	\item \( b_h \) is the bias vector,
	\item \( \tanh \) is the hyperbolic tangent activation function.
\end{itemize}

\subsubsection{Long Short-Term Memory (LSTM)}

Long Short-Term Memory (LSTM) networks extend traditional RNNs by incorporating memory cells and gating mechanisms, enabling them to retain long-term dependencies in sequential data. This makes them particularly effective for time-series applications like EEG-based sleep stage classification.

The proposed LSTM architecture includes two stacked LSTM layers with 128 and 64 units, respectively, both using the Tanh activation function to handle nonlinear temporal patterns in the data. To prevent overfitting, dropout layers with a rate of 0.2 are placed after each LSTM and dense layer.

\begin{table}[H]
	\centering
	\caption{LSTM Neural Network Architecture for Sleep Stage Classification}
	\label{tab:lstm_architecture}
	\begin{tabular}{lcc}
		\hline
		\textbf{Layer Type} & \textbf{Units} & \textbf{Activation Function} \\
		\hline
		Input Layer         & \texttt{X\_train}           & - \\
		LSTM Layer 1        & 128                         & Tanh \\
		Dropout Layer       & -                           & 0.2 (Dropout Rate) \\
		LSTM Layer 2        & 64                          & Tanh \\
		Dropout Layer       & -                           & 0.2 (Dropout Rate) \\
		Dense Layer         & 64                          & ReLU \\
		Dropout Layer       & -                           & 0.2 (Dropout Rate) \\
		Dense Layer         & 32                          & ReLU \\
		Dropout Layer       & -                           & 0.2 (Dropout Rate) \\
		Output Layer        & Number of classes (e.g., 5) & Softmax \\
		\hline
	\end{tabular}
\end{table}

The dense layers following the LSTM blocks (with 64 and 32 units) refine the extracted temporal features using ReLU activations. The final output layer uses the Softmax function to perform multi-class classification of sleep stages. LSTM models are particularly advantageous in capturing long-range dependencies and complex temporal dynamics, thus often outperforming simple RNNs for EEG sequence modeling.

LSTM cells operate through a series of gating mechanisms that regulate the flow of information. The key equations governing an LSTM cell at time step \( t \) are:

\begin{align}
	f_t &= \sigma(W_f x_t + U_f h_{t-1} + b_f) \quad \text{(Forget Gate)} \\
	i_t &= \sigma(W_i x_t + U_i h_{t-1} + b_i) \quad \text{(Input Gate)} \\
	\tilde{c}_t &= \tanh(W_c x_t + U_c h_{t-1} + b_c) \quad \text{(Candidate Cell State)} \\
	c_t &= f_t \odot c_{t-1} + i_t \odot \tilde{c}_t \quad \text{(Cell State Update)} \\
	o_t &= \sigma(W_o x_t + U_o h_{t-1} + b_o) \quad \text{(Output Gate)} \\
	h_t &= o_t \odot \tanh(c_t) \quad \text{(Hidden State Update)}
\end{align}

\noindent where:
\begin{itemize}
	\item \( f_t, i_t, o_t \) represent the forget, input, and output gates, respectively.
	\item \( \tilde{c}_t \) is the candidate cell state.
	\item \( c_t \) is the updated memory cell state.
	\item \( h_t \) is the hidden state at time \( t \).
	\item \( \sigma \) denotes the sigmoid activation function.
	\item \( \odot \) represents element-wise multiplication.
	\item \( W, U, b \) are the learned weight matrices and biases.
\end{itemize}

\subsubsection{Bidirectional Long Short-Term Memory (Bi-LSTM)}

The Bidirectional Long Short-Term Memory (Bi-LSTM) network enhances the LSTM model by processing input sequences in both forward and backward directions, enabling the model to learn context from both past and future data. This bidirectional approach improves the model's ability to capture dependencies in temporal data, making it highly effective for tasks such as EEG-based sleep stage classification.

The architecture consists of two bidirectional LSTM layers with 128 and 64 units, respectively, both utilizing the Tanh activation function. Dropout layers (with a rate of 0.2) are applied after each LSTM layer to prevent overfitting. Following the bidirectional LSTM layers, dense layers with 64 and 32 units, using ReLU activation, are added to refine the learned features before the final classification layer, which uses the Softmax activation to classify the sleep stages.

\begin{table}[!h]
	\centering
	\caption{Bidirectional LSTM Neural Network Architecture for Sleep Stage Classification}
	\label{tab:bi_lstm_architecture}
	\begin{tabular}{lcc}
		\hline
		\textbf{Layer Type} & \textbf{Units} & \textbf{Activation Function} \\
		\hline
		Input Layer         & \texttt{X\_train}            & - \\
		Bi-LSTM Layer 1     & 128                          & Tanh \\
		Dropout Layer       & -                            & 0.2 (Dropout Rate) \\
		Bi-LSTM Layer 2     & 64                           & Tanh \\
		Dropout Layer       & -                            & 0.2 (Dropout Rate) \\
		Dense Layer         & 64                           & ReLU \\
		Dropout Layer       & -                            & 0.2 (Dropout Rate) \\
		Dense Layer         & 32                           & ReLU \\
		Dropout Layer       & -                            & 0.2 (Dropout Rate) \\
		Output Layer        & Number of classes (e.g., 5)  & Softmax \\
		\hline
	\end{tabular}
\end{table}

Bi-LSTM networks are particularly well-suited for EEG-based sleep stage classification, as they allow the model to capture both past and future dependencies in the EEG signal, leading to improved accuracy in detecting sleep patterns.

The Bi-LSTM model processes input sequences in both forward and backward directions, maintaining two hidden states for each timestep. The key equations describing the forward and backward passes are:

\begin{align}
	\overrightarrow{h}_t &= \text{LSTM}(x_t, \overrightarrow{h}_{t-1}) \\
	\overleftarrow{h}_t &= \text{LSTM}(x_t, \overleftarrow{h}_{t+1}) \\
	h_t &= [\overrightarrow{h}_t; \overleftarrow{h}_t]
\end{align}

where:
\begin{itemize}
	\item \( \overrightarrow{h}_t \) is the forward LSTM hidden state at time step \( t \).
	\item \( \overleftarrow{h}_t \) is the backward LSTM hidden state at time step \( t \).
	\item \( h_t \) is the concatenated hidden state, combining both forward and backward states, used for final predictions.
\end{itemize}

\section{Results and Evaluation}

This section presents the performance evaluation of different deep learning models, using accuracy plots and confusion matrices for visual comparison.

\subsection{LSTM Model}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/lstm accuracy and loss plot.png}
	\caption{LSTM Accuracy and Loss Plot}
	\label{fig:lstm_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/LSTM_cm.png}
	\caption{LSTM Confusion Matrix}
	\label{fig:lstm_cm}
\end{figure}

\subsection{RNN Model}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/Rnn Accuracy Plot.png}
	\caption{RNN Accuracy Plot}
	\label{fig:rnn_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/RNN_cm.png}
	\caption{RNN Confusion Matrix}
	\label{fig:rnn_cm}
\end{figure}

\subsection{BiLSTM Model}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/BILSTM Accuaracy Plot.png}
	\caption{BiLSTM Accuracy Plot}
	\label{fig:bilstm_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/BiLSTM_cm.png}
	\caption{BiLSTM Confusion Matrix}
	\label{fig:bilstm_cm}
\end{figure}

\subsection{Neural Network (NN) Model}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{img/paper_2/deeplearning.png}
	\caption{Neural Network Accuracy Comparison}
	\label{fig:nn_accuracy}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.6\textwidth]{img/paper_2/NN_cm.png}
	\caption{Neural Network Confusion Matrix}
	\label{fig:nn_cm}
\end{figure}

\section{Comparison of Results}

In this section, we compare the performance of various models, including Neural Networks, RNN, LSTM, and Bi-LSTM, based on several evaluation metrics such as accuracy, precision, recall, and F1 score.

\subsection{Comparison Plot}
The following plot presents a comparison of the accuracy scores for different models.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{\img/paper_2/Comparision Plot.png}
	\caption{Accuracy Comparison Plot}
	\label{fig:accuracy_comparison}
\end{figure}

\subsection{Neural Network Comparison}
The performance of the Neural Network model is compared in the plot below, showing its accuracy.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{\img/paper_2/neural network comparison .png}
	\caption{Neural Network Comparison}
	\label{fig:nn_comparison}
\end{figure}

\subsection{Bi-LSTM Performance}
The Bi-LSTM model's accuracy is evaluated and compared with other models.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{\img/paper_2/BILSTM Accuaracy Plot.png}
	\caption{Bi-LSTM Accuracy Plot}
	\label{fig:bilstm_accuracy}
\end{figure}

\subsection{LSTM Performance}
The following figure shows the accuracy of the LSTM model, highlighting its effectiveness in comparison to other models.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{\img/paper_2/lstm accuracy and loss plot.png}
	\caption{LSTM Accuracy and Loss Plot}
	\label{fig:lstm_accuracy}
\end{figure}

\subsection{RNN Performance}
Lastly, the performance of the RNN model is shown below.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{\img/paper_2/Rnn Accuracy Plot.png}
	\caption{RNN Accuracy Plot}
	\label{fig:rnn_accuracy}
\end{figure}

### 2025-05-05 Chapters/paper2.tex
changes of paper 2

@@
 Given the inherent class imbalance in sleep stage distribution, we employed the \textbf{SMOTE} (Synthetic Minority Over-sampling Technique) to synthetically balance the dataset, improving model robustness across minority classes.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.70\textwidth]{img/paper_2/sleep event.PNG}
+	\caption{Event Mapping Plot}
+	\label{fig:lstm_accuracy}
+\end{figure}
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{\img/paper_2/Comparision Plot.png}
-	\caption{Accuracy Comparison Plot}
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/Comparision Plot.png}
+		\caption{Accuracy Comparison Across All Models}
 		\label{fig:accuracy_comparison}
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/BILSTM.png}
+		\caption{Bi-LSTM Model Performance}
+		\label{fig:bilstm_performance}
+	\end{minipage}
+
+	\vspace{0.5cm}
+
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/LSTM.png}
+		\caption{LSTM Model Performance}
+		\label{fig:lstm_performance}
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/RNN.png}
+		\caption{RNN Model Performance}
+		\label{fig:rnn_performance}
+	\end{minipage}
 \end{figure}
-
-\subsection{Neural Network Comparison}
-The performance of the Neural Network model is compared in the plot below, showing its accuracy.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{\img/paper_2/neural network comparison .png}
-	\caption{Neural Network Comparison}
-	\label{fig:nn_comparison}
-\end{figure}
-
-\subsection{Bi-LSTM Performance}
-The Bi-LSTM model's accuracy is evaluated and compared with other models.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{\img/paper_2/BILSTM Accuaracy Plot.png}
-	\caption{Bi-LSTM Accuracy Plot}
-	\label{fig:bilstm_accuracy}
-\end{figure}
-
-\subsection{LSTM Performance}
-The following figure shows the accuracy of the LSTM model, highlighting its effectiveness in comparison to other models.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{\img/paper_2/lstm accuracy and loss plot.png}
-	\caption{LSTM Accuracy and Loss Plot}
-	\label{fig:lstm_accuracy}
-\end{figure}
-
-\subsection{RNN Performance}
-Lastly, the performance of the RNN model is shown below.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\textwidth]{\img/paper_2/Rnn Accuracy Plot.png}
-	\caption{RNN Accuracy Plot}
-	\label{fig:rnn_accuracy}
-\end{figure}

### 2025-05-05 Chapters/paper3.tex
changes of paper 3

@@
 
-\section{Dataset Information}
+The SleepGCN-Transformer is a hybrid deep learning framework designed for accurate sleep stage classification by combining the strengths of Graph Convolutional Networks (GCNs) and Transformer architectures.
+This model is trained on the Sleep-EDF dataset, utilizing four critical physiological signal channels: EEG Fpz-Cz, EEG Pz-Oz, EOG (horizontal), and submental EMG.
+
+In this architecture, GCNs are employed to capture spatial relationships between the multimodal input signals, while the Transformer encoder is tasked with modeling the temporal dependencies across time sequences.
+This dual approach ensures that both spatial and sequential dynamics are effectively learned.
+
+Experimental results confirm the effectiveness of the model, achieving an accuracy of 93.12\% on the training set and 93.04\% on the validation set.
+To address class imbalance—particularly the underrepresentation of certain sleep stages—the model incorporates Focal Loss, which emphasizes harder-to-classify samples during training.
+
+Further analysis of feature importance revealed that the EMG and EEG Pz-Oz channels play a significant role in predicting sleep stages.
+This methodological framework not only enhances classification performance but also supports the integration of explainable AI techniques, contributing to the development of interpretable tools for clinical sleep analysis and decision-making.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=.9\linewidth]{"img/paper_3/Graph Convolution Neural Network -SleepGCN_Transformer Architechture"}
+	\caption{Proposed Architecture of the sleep GCN- Transformer Architecture}
+	\label{fig:graph-convolution-neural-network--sleepgcntransformer-architechture}
+\end{figure}
+
+\section{Dataset and Implementation Environment}
+
+Our experiments were conducted using the publicly available Sleep-EDF dataset, which is widely utilized for sleep stage classification research.
+The dataset can be accessed at \url{https://physionet.org/content/sleep-edfx/1.0.0/}, and it includes EEG recordings labeled with corresponding sleep stages, enabling effective supervised learning.
+
+The implementation was carried out in Python 3.13.1 using the PyTorch deep learning framework.
+Development was performed in Visual Studio Code (VS Code) on a machine powered by the Apple M1 chip, utilizing both CPU and Metal acceleration capabilities for training efficiency.
+
+For version control and collaborative development, we used GitHub to manage our codebase. The repository containing our source code is publicly available at \url{https://github.com/tanmay007thor/GCN}.
+
+The source code of this research is licensed under the Apache License 2.0, which allows for open use and distribution. The license details are available at \url{http://www.apache.org/licenses/}.
+
+To support data processing, model training, and evaluation, we relied on several key Python libraries.
+The primary libraries and their versions used in our project include: NumPy 2.2.2, Pandas 2.2.3, Scikit-learn 1.6.1, SciPy 1.15.1, Matplotlib 3.10.0, Seaborn 0.13.2, PyTorch 2.6.0, Torch Geometric 2.6.1, NetworkX 3.4.2, and MNE 1.9.0.
+This software environment ensured a reliable, efficient, and reproducible workflow throughout our research.
 
 \section{Preprocessing Techniques}
+\begin{figure}[H]
+	\centering
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_3/Signals}
+		\caption{Raw signal channels from the Sleep-EDF dataset}
+		\label{fig:raw_signals}
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_3/filtered chanels}
+		\caption{Preprocessed and filtered signal channels}
+		\label{fig:filtered_signals}
+	\end{minipage}
+\end{figure}
+
+\subsection{Data Information}
+
+The Sleep-EDF dataset comprises multichannel physiological signals, including EEG, EOG, EMG, nasal airflow, and temperature, each accompanied by event markers.
+These signals are recorded over 60-second intervals, with frequency content varying between 60 and 120 Hz.
+The dynamic nature of these signals introduces significant variability, making it challenging to identify consistent patterns for classification.
+Such rapid fluctuations can increase the likelihood of misclassification.
+Furthermore, the dataset presents several challenges such as high-frequency noise, class imbalance (notably a dominance of the Wake stage), and a substantial number of time steps—approximately 100 time steps per second—adding complexity to the task of modeling sleep stages.
+
+\subsection{Preprocessing Pipeline}
+
+To address these challenges, we developed a comprehensive preprocessing pipeline focused on selecting the most informative channels and reducing irrelevant data.
+Initially, we discarded several channels that were found to contribute minimally to sleep stage classification.
+The retained signals included the EEG channels (Fpz-Cz and Pz-Oz), the horizontal EOG channel (to track eye movement), and the submental EMG channel (for monitoring chin muscle activity).
+
+Each sleep stage in the dataset was mapped to a simplified class label to consolidate similar stages and reduce classification ambiguity.
+Table~\ref{tab:sleep_stage_mapping} presents the sleep stage-to-class mapping:
+
+\begin{table}[H]
+	\centering
+	\caption{Sleep Stage Mapping}
+	\label{tab:sleep_stage_mapping}
+	\begin{tabular}{llc}
+		\hline
+		\textbf{Sleep Stage} & \textbf{Label} & \textbf{Mapped Class} \\
+		\hline
+		Wake                & 0 & 0 \\
+		Stage 1             & 1 & 1 \\
+		Stage 2             & 2 & 2 \\
+		Stage 3             & 3 & 3 \\
+		Stage 4             & 4 & 3 \\
+		REM                 & 5 & 4 \\
+		\hline
+	\end{tabular}
+\end{table}
+
+The preprocessing begins with reading the EDF recordings and assigning sleep labels based on the mapped classes.
+Subsequently, we applied a bandpass filter to isolate relevant frequency ranges.
+Specifically, the signal was filtered to retain only the 0.3–30 Hz band using the `firwin` filter design, effectively removing noise and irrelevant frequencies outside this band.
+Though not a standard approach, this customized filtering was found to be better suited to our specific dataset and classification task.
+
+We then segmented the signals into fixed-length epochs of 30 seconds.
+The final preprocessed data was shaped into a three-dimensional format: $(X, 4, 3000)$, where $X$ is the number of epochs (samples), $4$ is the number of selected channels, and $3000$ corresponds to the number of time steps per epoch.
+
+\subsection{Filtering Method}
+
+The filtering was implemented using a windowed time-domain design via the `firwin` method.
+A Hamming window was employed, configured with a 0.0194 passband ripple and 53 dB stopband attenuation.
+The filter characteristics were as follows: lower passband edge at 0.30 Hz, lower transition bandwidth of 0.30 Hz (with a $-6$ dB cutoff at 0.15 Hz), upper passband edge at 30.00 Hz, and an upper transition bandwidth of 7.50 Hz (with a $-6$ dB cutoff at 33.75 Hz).
+These settings were carefully chosen to preserve the signal components relevant to sleep staging while minimizing noise.
 
@@
 
+To extract spatial features from the multi-channel data, we construct a graph using the four selected channels, where each channel acts as a node and the edges represent their interrelationships.
+These edges are weighted based on prior literature and empirical observations, assigning higher importance to EEG channels due to their strong influence on sleep stage detection.
+In contrast, connections involving EMG and EOG are assigned lower weights, as EOG shows weaker correlations.
+Although assigning self-loop weights is possible, we chose not to include them in this implementation.
+
+The final dataset comprises 11,076 samples, structured as \texttt{Data(x=[3000, 4], edge\_index=[2, 12], edge\_attr=[12], y=[1])}.
+Here, \texttt{x} represents the time-series data with 3,000 time steps across 4 channels, \texttt{edge\_index} defines the graph connectivity, \texttt{edge\_attr} contains the edge weights, and \texttt{y} denotes the sleep stage label.
+This dataset is created using the \texttt{SleepEEGDataset} class and is designed to effectively model spatial dependencies across EEG, EMG, and EOG signals, thereby enhancing sleep stage classification performance.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\linewidth]{"img/paper_3/Graph Weightage"}
+	\caption{Graph Dataset Creation}
+	\label{fig:graph-weightage}
+\end{figure}
+
+\subsection{GCN Architecture}
+
+To capture spatial dependencies across the selected EEG, EMG, and EOG channels, we incorporate a Graph Convolutional Network (GCN) into our model.
+Each sample is structured as a graph with four nodes—corresponding to the four selected channels—and each node contains a time series of 3,000 time steps.
+The dataset comprises 11,076 such graph-based samples.
+
+The GCN consists of five identical graph convolution layers.
+The initial input has a shape of (4, 3000), which is projected to (4, 256) after the first graph convolution.
+Each layer includes batch normalization (256 units), followed by a ReLU activation function and a dropout of 0.1.
+After the final GCN layer, we apply global mean pooling to obtain a graph-level representation.
+
+The propagation rule used in each GCN layer is given by:
+
+\begin{equation}
+	H^{(l+1)} = \sigma\left(\tilde{D}^{-1/2} \tilde{A} \tilde{D}^{-1/2} H^{(l)} W^{(l)}\right)
+\end{equation}
+
+\begin{itemize}
+	\item $H^{(l)} \in \mathbb{R}^{N \times D}$ is the feature matrix at layer $l$, where $N$ is the number of nodes and $D$ is the feature dimension.
+	\item $\tilde{A} = A + I$ is the adjacency matrix with self-loops.
+	\item $\tilde{D}$ is the degree matrix corresponding to $\tilde{A}$.
+	\item $W^{(l)}$ is the learnable weight matrix for layer $l$.
+	\item $\sigma$ denotes a non-linear activation function, typically ReLU.
+\end{itemize}
+
+The simplified form for one GCN layer operation is:
+
+\begin{equation}
+	X' = \sigma\left(\tilde{D}^{-1/2} \tilde{A} \tilde{D}^{-1/2} X W\right)
+\end{equation}
+
+To aggregate node features into a graph-level embedding, we use global mean pooling:
+
+\begin{equation}
+	x = \frac{1}{|V|} \sum_{v \in V} h_v
+\end{equation}
+
+\begin{itemize}
+	\item $V$ is the set of nodes in the graph.
+	\item $h_v$ is the feature vector of node $v$.
+	\item $x$ is the pooled feature vector representing the entire graph.
+\end{itemize}
+
+We apply standard activation and normalization functions during training:
+
+\begin{equation}
+	\text{ReLU}(x) = \max(0, x)
+\end{equation}
+
+\begin{equation}
+	\text{softmax}(x_i) = \frac{e^{x_i}}{\sum_j e^{x_j}}
+\end{equation}
+
+\begin{equation}
+	x'' = \frac{x' - \mu}{\sigma}
+\end{equation}
+
+\begin{itemize}
+	\item $\mu$ is the mean of $x'$.
+	\item $\sigma$ is the standard deviation of $x'$.
+	\item $x''$ is the normalized feature representation.
+\end{itemize}
+
+\subsection{Transformer Architecture}
+
+To capture temporal dependencies in the input signal, we utilize a Transformer encoder comprising two stacked layers.
+Initially, the input is projected into a continuous vector space using an embedding layer.
+Since Transformers lack inherent mechanisms for handling sequential order, positional encodings are added to preserve time-step information.
+
+Following embedding and encoding, we use multi-head self-attention to allow the model to focus on different aspects of the signal simultaneously.
+The attention outputs are concatenated and passed through a feedforward network with dropout regularization.
+Finally, a fully connected classifier outputs logits for four-class sleep stage classification.
+
+The operations within the Transformer encoder are summarized using the following equations:
+\subsection{Transformer: Key Components and Equations}
+
+\begin{itemize}
+	\item \textbf{Input Embedding:} Maps input tokens into a continuous vector space.
+	\begin{equation}
+		X_{\text{embed}} = X W_{\text{embed}}
+	\end{equation}
+
+	\item \textbf{Positional Encoding:} Adds sequence order information using sine and cosine functions.
+	\begin{equation}
+		PE(\text{pos}, 2i) = \sin\left(\frac{\text{pos}}{10000^{2i/d}}\right)
+	\end{equation}
+	\begin{equation}
+		PE(\text{pos}, 2i+1) = \cos\left(\frac{\text{pos}}{10000^{2i/d}}\right)
+	\end{equation}
+
+	\item \textbf{Query, Key, and Value Computation:} Core of attention mechanism, derived from input.
+	\begin{equation}
+		Q = XW_q,\quad K = XW_k,\quad V = XW_v
+	\end{equation}
+
+	\item \textbf{Scaled Dot-Product Attention:} Computes weighted output of values using attention scores.
+	\begin{equation}
+		H_1 = \text{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right) V
+	\end{equation}
+
+	\item \textbf{Multi-Head Attention:} Combines multiple attention outputs for diverse feature learning.
+	\begin{equation}
+		H = \text{Concat}(H_1, H_2, ..., H_h) W_0
+	\end{equation}
+\end{itemize}
+
+\subsection{Training Pipeline}
+
+The training process for the SleepGCN-Transformer model is designed to handle class imbalance, prevent overfitting, and improve generalization.
+Below, we outline the training methodology using key equations:
+
+\textbf{Label Smoothing:} To reduce overconfidence in predictions and improve generalization, we apply label smoothing:
+\begin{equation}
+	y_{\text{smooth}} = y(1 - \epsilon) + \frac{\epsilon}{C}
+\end{equation}
+
+\textbf{Combined Loss Function:} We incorporate both Focal Loss and smoothed labels:
+\begin{equation}
+	L = \alpha(1 - p)^\gamma (-y_{\text{smooth}} \log p)
+\end{equation}
+
+\textbf{Focal Loss (for class imbalance):}
+\begin{equation}
+	FL(p_t) = -\alpha(1 - p_t)^\gamma \log(p_t)
+\end{equation}
+
+\textbf{Class Weight Computation:} Each class receives a weight based on its frequency:
+\begin{equation}
+	w_c = \frac{N}{\sqrt{K + 1}}
+\end{equation}
+
+\textbf{AdamW Optimizer:} We update parameters using the AdamW optimization rule:
+\begin{equation}
+	\theta_{t+1} = \theta_t - \eta (\nabla\phi(\theta_t) + \lambda\theta_t)
+\end{equation}
+
+\textbf{AdamW with Decoupled Weight Decay:}
+\begin{equation}
+	\theta_t' = \theta_t - \eta \nabla\phi(\theta_t)
+\end{equation}
+\begin{equation}
+	\theta_{t+1} = \theta_t' - \eta \lambda \theta_t
+\end{equation}
+
+\textbf{Learning Rate Scheduling:} Cosine Annealing is used to adjust learning rates dynamically:
+\begin{equation}
+	\eta_t = \eta_{\min} + \frac{1}{2} (\eta_{\max} - \eta_{\min}) \left(1 + \cos\left(\frac{\pi T_{\text{cur}}}{T_{\text{max}}}\right)\right)
+\end{equation}
+
+\textbf{Pipeline Summary:}
+\begin{itemize}
+	\item Load preprocessed EEG dataset for training and validation.
+	\item Compute class weights.
+	\item Apply label smoothing and Focal Loss.
+	\item Optimize with AdamW and update learning rate via cosine scheduler.
+	\item Monitor loss and accuracy metrics per epoch.
+\end{itemize}
+
 \section{Results and Evaluation}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/loss plot}
+		\caption{Loss plot}
+		\label{fig:loss-plot}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/accuracy plot}
+		\caption{Accuracy plot}
+		\label{fig:accuracy-plot}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/3D acc loss plot}
+		\caption{3D Accuracy vs. Loss}
+		\label{fig:3d-acc-loss-plot}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/3d gradient}
+		\caption{3D Gradient visualization}
+		\label{fig:3d-gradient}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/sample distribution plot pdf}
+		\caption{Sample distribution}
+		\label{fig:sample-distribution-plot-pdf}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/auc vs roc curve}
+		\caption{AUC vs. ROC Curve}
+		\label{fig:auc-vs-roc-curve}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/confusion matrix percentatge wise}
+		\caption{Confusion matrix (percentage)}
+		\label{fig:confusion-matrix-percentatge-wise}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/confusion matrix samples}
+		\caption{Confusion matrix (samples)}
+		\label{fig:confusion-matrix-samples}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.31\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/f1_score_plot}
+		\caption{F1 Score}
+		\label{fig:f1scoreplot}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.31\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/precision_plot}
+		\caption{Precision}
+		\label{fig:precisionplot}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.31\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/recall_plot}
+		\caption{Recall}
+		\label{fig:recallplot}
+	\end{minipage}
+\end{figure}
+
+\subsection{XAI Visualization with Heatmap and Plots}
+
+\begin{figure}[H]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{"img/paper_3/stage n1"}
+		\caption{Stage N1}
+		\label{fig:stage-n1}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{"img/paper_3/stage n2"}
+		\caption{Stage N2}
+		\label{fig:stage-n2}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{"img/paper_3/stage n3"}
+		\caption{Stage N3}
+		\label{fig:stage-n3}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{"img/paper_3/stage rem"}
+		\caption{Stage REM}
+		\label{fig:stage-rem}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{"img/paper_3/stage wake"}
+		\caption{Stage Wake}
+		\label{fig:stage-wake}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/XAI4}
+		\caption{XAI Visualization}
+		\label{fig:xai4}
+	\end{minipage}
+\end{figure}
+
+\subsection{Time Complexity Analysis}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/transformer_encoder_on²}
+		\caption{Transformer Encoder Complexity}
+		\label{fig:transformerencoderon2}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/data_loading_complexity_with_legend_updated}
+		\caption{Data Loading Complexity}
+		\label{fig:dataloadingcomplexitywithlegendupdated}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/evaluation_oe_val·…}
+		\caption{Evaluation Complexity}
+		\label{fig:evaluationoeval}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/fully_connected_layer_o1}
+		\caption{Fully Connected Layer Complexity}
+		\label{fig:fullyconnectedlayero1}
+	\end{minipage}
+\end{figure}
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/gcn_layers_on}
+		\caption{GCN Layers Complexity}
+		\label{fig:gcnlayerson}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/training_loop_ot·…}
+		\caption{Training Loop Complexity}
+		\label{fig:trainingloopot}
+	\end{minipage}
+\end{figure}

### 2025-05-05 Chapters/reference.tex
changes of ref

\begin{thebibliography}{99}

	\bibitem{han2024classification}
	H.~Han et~al., ``Classification and automatic scoring of arousal intensity during sleep stages using machine learning,'' \emph{Scientific Reports}, vol.~14, no.~1, Dec. 2024, doi: \href{https://doi.org/10.1038/s41598-023-50653-9}{10.1038/s41598-023-50653-9}.

	\bibitem{tsinalis2016automatic}
	O.~Tsinalis et~al., ``Automatic Sleep Stage Scoring Using Time-Frequency Analysis and Stacked Sparse Autoencoders,'' \emph{Annals of Biomedical Engineering}, vol.~44, no.~5, May 2016, doi: \href{https://doi.org/10.1007/s10439-015-1444-y}{10.1007/s10439-015-1444-y}.

	\bibitem{krakovska2011automatic}
	A.~Krakovská and K.~Mezeiová, ``Automatic sleep scoring: A search for an optimal combination of measures,'' \emph{Artificial Intelligence in Medicine}, vol.~53, no.~1, Sep. 2011, doi: \href{https://doi.org/10.1016/j.artmed.2011.06.004}{10.1016/j.artmed.2011.06.004}.

	\bibitem{biswal2018expert}
	S.~Biswal et~al., ``Expert-level sleep scoring with deep neural networks,'' \emph{Journal of the American Medical Informatics Association}, vol.~25, no.~12, Dec. 2018, doi: \href{https://doi.org/10.1093/jamia/ocy131}{10.1093/jamia/ocy131}.

	\bibitem{ronzhina2012sleep}
	M.~Ronzhina et~al., ``Sleep scoring using artificial neural networks,'' \emph{Sleep Medicine Reviews}, vol.~16, no.~3, Jun. 2012, doi: \href{https://doi.org/10.1016/j.smrv.2011.06.003}{10.1016/j.smrv.2011.06.003}.

	\bibitem{alsolai2022systematic}
	H.~Alsolai et~al., ``A Systematic Review of Literature on Automated Sleep Scoring,'' \emph{IEEE Access}, vol.~10, pp.~79419--79443, 2022, doi: \href{https://doi.org/10.1109/ACCESS.2022.3194145}{10.1109/ACCESS.2022.3194145}.

	\bibitem{fiorillo2019automated}
	L.~Fiorillo et~al., ``Automated sleep scoring: A review of the latest approaches,'' \emph{Sleep Medicine Reviews}, vol.~48, Dec. 2019, doi: \href{https://doi.org/10.1016/j.smrv.2019.07.007}{10.1016/j.smrv.2019.07.007}.

	\bibitem{faust2019review}
	O.~Faust et~al., ``A review of automated sleep stage scoring based on physiological signals for the new millennia,'' \emph{Computer Methods and Programs in Biomedicine}, vol.~176, Jul. 2019, doi: \href{https://doi.org/10.1016/j.cmpb.2019.04.032}{10.1016/j.cmpb.2019.04.032}.

	\bibitem{chriskos2021review}
	P.~Chriskos et~al., ``A review on current trends in automatic sleep staging through bio-signal recordings and future challenges,'' \emph{Sleep Medicine Reviews}, vol.~55, Feb. 2021, doi: \href{https://doi.org/10.1016/j.smrv.2020.101377}{10.1016/j.smrv.2020.101377}.

	\bibitem{phan2022sleeptransformer}
	H.~Phan, K.~B. Mikkelsen, O.~Y. Chen, P.~Koch, A.~Mertins, and M.~De~Vos, ``SleepTransformer: Automatic Sleep Staging with Interpretability and Uncertainty Quantification,'' \emph{IEEE Transactions on Biomedical Engineering}, 2022.

	\bibitem{han2024classification}
	H.~Han et~al., ``Classification and automatic scoring of arousal intensity during sleep stages using machine learning,'' \emph{Scientific Reports}, vol.~14, no.~1, Dec. 2024.

	\bibitem{tsinalis2016automatic}
	O.~Tsinalis et~al., ``Automatic Sleep Stage Scoring Using Time-Frequency Analysis and Stacked Sparse Autoencoders,'' \emph{Annals of Biomedical Engineering}, vol.~44, no.~5, May 2016.

	\bibitem{krakovska2011automatic}
	A.~Krakovská and K.~Mezeiová, ``Automatic sleep scoring: A search for an optimal combination of measures,'' \emph{Artificial Intelligence in Medicine}, vol.~53, no.~1, Sep. 2011.

	\bibitem{biswal2018expert}
	S.~Biswal et~al., ``Expert-level sleep scoring with deep neural networks,'' \emph{JAMIA}, vol.~25, no.~12, Dec. 2018.

	\bibitem{ronzhina2012sleep}
	M.~Ronzhina et~al., ``Sleep scoring using artificial neural networks,'' \emph{Sleep Medicine Reviews}, vol.~16, no.~3, Jun. 2012.

	\bibitem{alsolai2022systematic}
	H.~Alsolai et~al., ``A Systematic Review of Literature on Automated Sleep Scoring,'' \emph{IEEE Access}, vol.~10, 2022.

	\bibitem{fiorillo2019automated}
	L.~Fiorillo et~al., ``Automated sleep scoring: A review of the latest approaches,'' \emph{Sleep Medicine Reviews}, vol.~48, Dec. 2019.

	\bibitem{faust2019review}
	O.~Faust et~al., ``A review of automated sleep stage scoring based on physiological signals for the new millennia,'' \emph{Computer Methods and Programs in Biomedicine}, vol.~176, Jul. 2019.

	\bibitem{chriskos2021review}
	P.~Chriskos et~al., ``A review on current trends in automatic sleep staging through bio-signal recordings and future challenges,'' \emph{Sleep Medicine Reviews}, vol.~55, Feb. 2021.

	\bibitem{phan2022sleeptransformer}
	H.~Phan et~al., ``SleepTransformer: Automatic Sleep Staging with Interpretability and Uncertainty Quantification,'' \emph{IEEE Transactions on Biomedical Engineering}, 2022.

	\bibitem{dai2023multichannelsleepnet}
	Y.~Dai et~al., ``MultiChannelSleepNet: A Transformer-Based Model for Automatic Sleep Stage Classification With PSG,'' \emph{IEEE J. Biomed. Health Inform.}, 2023.

	\bibitem{guo2024flexsleeptransformer}
	Y.~Guo, M.~Nowakowski, and W.~Dai, ``FlexSleepTransformer: A Transformer-Based Sleep Staging Model with Flexible Input Channel Configurations,'' \emph{Scientific Reports}, 2024.

	\bibitem{yao2022cnntransformer}
	Z.~Yao and X.~Liu, ``A CNN-Transformer Deep Learning Model for Real-time Sleep Stage Classification in an Energy-Constrained Wireless Device,'' \emph{medRxiv}, 2022.

	\bibitem{supratak2017deepsleepnet}
	A.~Supratak et~al., ``DeepSleepNet: A Model for Automatic Sleep Stage Scoring Based on Raw Single-Channel EEG,'' \emph{IEEE Trans. Neural Syst. Rehabil. Eng.}, vol.~25, no.~11, pp.~1998--2008, 2017.

	\bibitem{eldele2021attnsleep}
	E.~Eldele et~al., ``AttnSleep: An Attention-Based Deep Learning Approach for Sleep Stage Classification With Single-Channel EEG,'' \emph{IEEE Trans. Neural Syst. Rehabil. Eng.}, 2021.

	\bibitem{ji2022jkstgcn}
	X.~Ji, Y.~Li, and P.~Wen, ``Jumping Knowledge-Based Spatial-Temporal Graph Convolutional Networks for Automatic Sleep Stage Classification,'' \emph{IEEE Trans. Neural Syst. Rehabil. Eng.}, 2022.

	\bibitem{mousavi2019sleepeegnet}
	S.~Mousavi, F.~Afghah, and U.~R. Acharya, ``SleepEEGNet: Automated Sleep Stage Scoring With Sequence-to-Sequence Deep Learning Approach,'' \emph{PLOS ONE}, 2019.

	\bibitem{phan2019seqsleepnet}
	H.~Phan et~al., ``SeqSleepNet: End-to-End Hierarchical Recurrent Neural Network for Sequence-to-Sequence Automatic Sleep Staging,'' \emph{IEEE Trans. Neural Syst. Rehabil. Eng.}, 2019.

	\bibitem{zhang2024cnntransformerconvlstm}
	Zhang et~al., ``A CNN-Transformer-ConvLSTM-CRF Hybrid Network for Sleep Stage Classification,'' \emph{IEEE Sensors Journal}, 2024.

	\bibitem{liu2024review}
	Liu et~al., ``Automatic Sleep Stage Classification Using Deep Learning: Signals, Data Representation, and Neural Networks,'' \emph{Artificial Intelligence Review}, 2024.

	\bibitem{ctdal2022}
	CT-DAL, ``Convolutional Transformer with Domain Adversarial Learning for Multi-Channel Sleep Stage Classification,'' \emph{Available on Semantic Scholar}, 2022.

	\bibitem{pei2022hybrid}
	W.~Pei, Y.~Li, S.~Siuly, and P.~Wen, ``A Hybrid Deep Learning Scheme for Multi-Channel Sleep Stage Classification,'' \emph{Computers, Materials \& Continua}, 2022.

	\bibitem{mostafaei2024crossmodality}
	S.~H. Mostafaei, J.~Tanha, and A.~Sharafkhaneh, ``A Novel Deep Learning Model Based on Transformer and Cross-Modality Attention for Classification of Sleep Stages,'' \emph{Journal of Biomedical Informatics}, 2024.

	\bibitem{jeong2024explainablevit}
	Jeong et~al., ``Explainable Vision Transformer for Automatic Visual Sleep Staging,'' \emph{npj Digital Medicine}, 2024.

	\bibitem{pei2024mfcc}
	W.~Pei, Y.~Li et~al., ``An Automatic Method Using MFCC Features for Sleep Stage Classification,'' \emph{Biomedical Signal Processing and Control}, 2024.

	\bibitem{sors2018cnn}
	A.~Sors et~al., ``A Convolutional Neural Network for Sleep Stage Scoring from Raw Single-Channel EEG,'' \emph{Biomedical Signal Processing and Control}, 2018.

	\bibitem{seo2020iitnet}
	H.~Seo et~al., ``Intra- and Inter-Epoch Temporal Context Network (IITNet) Using Sub-Epoch Features for Automatic Sleep Scoring on Raw Single-Channel EEG,'' \emph{Biomedical Signal Processing and Control}, 2020.

	\bibitem{biswal2017sleepnet}
	S.~Biswal et~al., ``SLEEPNET: Automated Sleep Staging System via Deep Learning,'' \emph{Journal of the American Medical Informatics Association}, 2017.

	\bibitem{phan2018jointcnn}
	H.~Phan et~al., ``Joint Classification and Prediction CNN Framework for Automatic Sleep Stage Classification,'' \emph{IEEE Transactions on Biomedical Engineering}, 2018.

\end{thebibliography}

### 2025-05-06 Chapters/paper3.tex
Updated paper3 and compiled files

@@
 
+\begin{center}
+	\begin{minipage}[t]{0.48\textwidth}
+		\begin{align}
+			\text{Accuracy} &= \frac{TP + TN}{TP + TN + FP + FN} \\
+			\text{Precision} &= \frac{TP}{TP + FP} \\
+			\text{Recall} &= \frac{TP}{TP + FN} \\
+			\text{F1\text{-}Score} &= 2 \cdot \frac{\text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}} \\
+			\text{Support} &= TP + FN \\
+			\text{Specificity} &= \frac{TN}{TN + FP}
+		\end{align}
+	\end{minipage}
+	\hfill	
+	\begin{minipage}[t]{0.48\textwidth}
+		\begin{align}
+			\text{Macro Precision} &= \frac{1}{C} \sum_{i=1}^{C} \frac{TP_i}{TP_i + FP_i} \\
+			\text{Weighted Precision} &= \sum_{i=1}^{C} \frac{n_i}{N} \cdot \frac{TP_i}{TP_i + FP_i} \\
+			\mathcal{L}_{CE} &= -\sum_{i=1}^{C} y_i \log(\hat{y}_i) \\
+			\mathcal{L}_{FL} &= -\alpha_t (1 - p_t)^\gamma \log(p_t)
+		\end{align}
+	\end{minipage}
+\end{center}
 \begin{figure}[htbp]
@@
 
-\begin{figure}[htbp]
+\begin{figure}[H]
 	\centering
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/transformer_encoder_on²}
-		\caption{Transformer Encoder Complexity}
-		\label{fig:transformerencoderon2}
+	\begin{minipage}[b]{0.40\linewidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_3/preprocessing_on³}
+		\caption{Preprocessing (\(O(MN^3 + KEN + MN)\))}
+		\label{fig:preprocessingon3}
 	\end{minipage}
 	\hfill
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/data_loading_complexity_with_legend_updated}
-		\caption{Data Loading Complexity}
-		\label{fig:dataloadingcomplexitywithlegendupdated}
+	\begin{minipage}[b]{0.40\linewidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_3/training_loop_ot·…}
+		\caption{Training Loop (\(O(T(LN + 2N^2D/10^4 + DC))\))}
+		\label{fig:trainingloopot}
 	\end{minipage}
@@
 
-\begin{figure}[htbp]
+\noindent Preprocessing: \(O(MN^3 + KEN + MN)\) \\
+Training Loop: \(O(T(LN + 2N^2D/10^4 + DC))\)
+
+\begin{figure}[H]
 	\centering
-	\begin{minipage}[b]{0.45\linewidth}
+	\begin{minipage}[b]{0.40\linewidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_3/gcn_layers_on}
+		\caption{GCN Layers (\(O(LN)\))}
+		\label{fig:gcnlayerson}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.40\linewidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_3/transformer_encoder_on²}
+		\caption{Transformer Encoder (\(O(N^2D)\))}
+		\label{fig:transformerencoderon2}
+	\end{minipage}
+\end{figure}
+
+\noindent GCN Layers: \(O(LN)\) \\
+Transformer Encoder: \(O(N^2D)\)
+
+\begin{figure}[H]
+	\centering
+	\begin{minipage}[b]{0.40\linewidth}
+		\centering
 		\includegraphics[width=\linewidth]{img/paper_3/evaluation_oe_val·…}
-		\caption{Evaluation Complexity}
+		\caption{Evaluation Loop (\(O(E_{\text{val}}(LN + 2N^2D/10^4 + DC))\))}
 		\label{fig:evaluationoeval}
@@
 	\hfill
-	\begin{minipage}[b]{0.45\linewidth}
+	\begin{minipage}[b]{0.40\linewidth}
+		\centering
 		\includegraphics[width=\linewidth]{img/paper_3/fully_connected_layer_o1}
-		\caption{Fully Connected Layer Complexity}
+		\caption{Fully Connected Layer (\(O(DC)\))}
 		\label{fig:fullyconnectedlayero1}
@@
 
-\begin{figure}[htbp]
-	\centering
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/gcn_layers_on}
-		\caption{GCN Layers Complexity}
-		\label{fig:gcnlayerson}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/training_loop_ot·…}
-		\caption{Training Loop Complexity}
-		\label{fig:trainingloopot}
-	\end{minipage}
-\end{figure}
+\noindent Evaluation Loop: \(O(E_{\text{val}}(LN + 2N^2D/10^4 + DC))\) \\
+Fully Connected Layer: \(O(DC)\)

### 2025-05-06 Chapters/paper3.tex
changes

@@
 \end{center}
+
+The model demonstrates high performance in classifying the Wake and N3 sleep stages, achieving correct classification rates of 97.14\% and 97.94\%, respectively.
+In contrast, the N2 stage is accurately classified 88.3\% of the time, though it shows some confusion with neighboring N1 and REM stages.
+Misclassifications are more prominent for the N1 and REM stages, with the REM stage frequently misidentified as N1, exhibiting a 17.97\% confusion rate.
+These results indicate strong general performance by the model, though enhancements may be beneficial for improving distinction between the N1 and REM stages.
+
+\begin{figure}[htbp]
+	\centering
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/confusion matrix percentatge wise}
+		\caption{Confusion matrix (percentage)}
+		\label{fig:confusion-matrix-percentatge-wise}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[b]{0.45\linewidth}
+		\includegraphics[width=\linewidth]{img/paper_3/confusion matrix samples}
+		\caption{Confusion matrix (samples)}
+		\label{fig:confusion-matrix-samples}
+	\end{minipage}
+\end{figure}
+
 \begin{figure}[htbp]
@@
 \end{figure}
+
+The class weight distribution used during training is as follows: 0: 1.1923, 1: 5.8152, 2: 2.4719, 3: 4.8223, 4: 4.0506.
+These values reflect adjustments to counter class imbalance and ensure fairer learning across all sleep stages.
+
+The progression of model performance across selected epochs is summarized below:
+\begin{itemize}
+	\item Epoch 1: Train Loss = 0.1915, Train Accuracy = 82.91\%; Validation Loss = 0.2110, Validation Accuracy = 83.84\%
+	\item Epoch 2: Train Loss = 0.1695, Train Accuracy = 87.74\%; Validation Loss = 0.2168, Validation Accuracy = 69.46\%
+	\item Epoch 3: Train Loss = 0.1669, Train Accuracy = 88.15\%; Validation Loss = 0.1803, Validation Accuracy = 88.39\%
+	\item Epoch 4: Train Loss = 0.1659, Train Accuracy = 88.38\%; Validation Loss = 0.1534, Validation Accuracy = 91.83\%
+	\item Epoch 5: Train Loss = 0.1621, Train Accuracy = 89.45\%; Validation Loss = 0.1615, Validation Accuracy = 89.11\%
+	\item Epoch 10: Train Loss = 0.1533, Train Accuracy = 91.03\%; Validation Loss = 0.1471, Validation Accuracy = 91.16\%
+	\item Epoch 15: Train Loss = 0.1458, Train Accuracy = 92.59\%; Validation Loss = 0.1484, Validation Accuracy = 90.98\%
+	\item Epoch 20: Train Loss = 0.1413, Train Accuracy = 93.12\%; Validation Loss = 0.1390, Validation Accuracy = 93.04\%
+\end{itemize}
 
@@
 
+The sample distribution plot reveals a nearly normal distribution of data across the various sleep stages, confirming a balanced dataset.
+For the test phase, approximately 1,050 samples were utilized, with 210 samples assigned to each class, ensuring equitable and unbiased performance evaluation.
+
 \begin{figure}[htbp]
@@
 		\label{fig:auc-vs-roc-curve}
-	\end{minipage}
-\end{figure}
-
-\begin{figure}[htbp]
-	\centering
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/confusion matrix percentatge wise}
-		\caption{Confusion matrix (percentage)}
-		\label{fig:confusion-matrix-percentatge-wise}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/confusion matrix samples}
-		\caption{Confusion matrix (samples)}
-		\label{fig:confusion-matrix-samples}
 	\end{minipage}

### 2025-05-06 Chapters/paper1.tex
changes

@@
 	\centering
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
@@
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
@@
 	\end{minipage}
-
+\end{figure}
 	\vspace{1em}
-
-	\begin{minipage}[b]{0.45\textwidth}
+\begin{figure}[H]
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
@@
 \end{figure}
+
+The Random Forest classifier demonstrates solid performance with an accuracy of 0.764, precision of 0.777, and sensitivity of 0.770.
+Its F1-Score of 0.766 reflects a balanced performance between precision and recall.
 
@@
 
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
@@
 \end{figure}
+
+The Gradient Boosting classifier achieves moderate performance with an accuracy of 0.702, precision of 0.687, and sensitivity of 0.689.
+Its F1-Score of 0.687 indicates a fairly balanced trade-off between precision and recall.
 
@@
 	\centering
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//Ensemble_confusion_matrix.png}
+		\includegraphics[width=\textwidth]{img/paper_1/Ensemble_confusion_matrix.png}
 		\caption{Ensemble Learning Confusion Matrix}
@@
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//accuracy_curve.png}
+		\includegraphics[width=\textwidth]{img/paper_1/accuracy_curve.png}
 		\caption{Ensemble Learning Accuracy Curve}
@@
 
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//loss_curve.png}
+		\includegraphics[width=\textwidth]{img/paper_1/loss_curve.png}
 		\caption{Ensemble Learning Loss Curve}
@@
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//precision_per_class.png}
+		\includegraphics[width=\textwidth]{img/paper_1/precision_per_class.png}
 		\caption{Ensemble Learning Precision Per Class}
@@
 \end{figure}
+
+The Ensemble Learning model exhibits strong performance with an accuracy of 0.831, precision of 0.830, and sensitivity of 0.820.
+Its F1-Score of 0.819 underscores its effectiveness in maintaining a balance between precision and recall.
 
@@
 	\centering
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//confusion_matrix.png}
+		\includegraphics[width=\textwidth]{img/paper_1/confusion_matrix.png}
 		\caption{Gradient Boosting Confusion Matrix}
@@
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//accuracy_curve.png}
+		\includegraphics[width=\textwidth]{img/paper_1/accuracy_curve.png}
 		\caption{Gradient Boosting Accuracy Curve}
@@
 
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//loss_curve.png}
+		\includegraphics[width=\textwidth]{img/paper_1/loss_curve.png}
 		\caption{Gradient Boosting Loss Curve}
@@
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
-		\includegraphics[width=\textwidth]{img/paper_1//precision_per_class.png}
+		\includegraphics[width=\textwidth]{img/paper_1/precision_per_class.png}
 		\caption{Gradient Boosting Precision Per Class}
@@
 
-\section{Results and Evaluation with K-fold Cross Validation }
+The Gradient Boosting Classifier shows competent performance with an accuracy of 0.753, precision of 0.737, and sensitivity of 0.746.
+Its F1-Score of 0.734 suggests a solid equilibrium between precision and recall.
+
+\section{Results and Evaluation with K-fold Cross Validation}
+\begin{table}[!h]
+	\centering
+	\caption{Classifier Performance Summary}
+	\begin{tabular}{lcccc}
+		\toprule
+		\textbf{Classifier} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} \\
+		\midrule
+		Ensemble Learning & 0.95 & 0.86 & 0.79 & 0.82 \\
+		Random Forest     & 0.88 & 0.85 & 0.81 & 0.82 \\
+		Gradient Boosting & 0.87 & 0.83 & 0.81 & 0.82 \\
+		SGD Classifier    & 0.80 & 0.73 & 0.70 & 0.70 \\
+		\bottomrule
+	\end{tabular}
+\end{table}
 
@@
 	\centering
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
 		\includegraphics[width=\textwidth]{img/paper_1/Ensemble Learning_percentage_based.png}
-		\caption{Ensemble Learning Confusion Matrix }
+		\caption{Ensemble Learning Percentage-Based Confusion Matrix}
 	\end{minipage}
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
 		\includegraphics[width=\textwidth]{img/paper_1/Random Forest_percentage_based.png}
-		\caption{Random Forest Confusion Matrix }
+		\caption{Random Forest Percentage-Based Confusion Matrix}
 	\end{minipage}
+\end{figure}
 
-	\vspace{1em}
+\textbf{Analysis:} Ensemble Learning outperforms all other models with the highest average accuracy (95\%) and strong macro F1-score (0.82), particularly excelling in the Wake stage with near-perfect precision and recall.
+Random Forest and Gradient Boosting both deliver balanced performance with 88\% and 87\% accuracy respectively, showing robust results in N2 and REM stages.
+SGD Classifier, while computationally efficient, lags behind in performance, especially in accurately detecting the N1 stage, reflecting a much lower F1-score (0.70).
 
-	\begin{minipage}[b]{0.45\textwidth}
+\begin{figure}[H]
+	\centering
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
 		\includegraphics[width=\textwidth]{img/paper_1/GB Classifier_percentage_based.png}
-		\caption{GB Classifier Confusion Matrix }
+		\caption{Gradient Boosting Percentage-Based Confusion Matrix}
 	\end{minipage}
 	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{minipage}[t]{0.48\textwidth}
 		\centering
 		\includegraphics[width=\textwidth]{img/paper_1/SGD_percentage_based.png}
-		\caption{SGD Confusion Matrix }
+		\caption{SGD Percentage-Based Confusion Matrix}
 	\end{minipage}
@@
 
+The comparison tables highlight that Gradient Boosting and Ensemble Learning consistently outperform other classifiers across overall and per-stage metrics, particularly in precision and F1-score.
+
 \begin{figure}[H]
 	\centering
-	\begin{minipage}[b]{0.45\textwidth}
+	\includegraphics[width=0.99\linewidth]{img/paper_1/bar plot}
+	\caption{Comparison of Classifier Performance Based on Precision, Recall, and F1-Score}
+	\label{fig:bar-plot}
+\end{figure}
+
+ Notably, REM and Wake (W) stages achieve the highest precision and recall, while N1 and N2 stages remain more challenging for all models.
+\begin{table}[h!]
 	\centering
-		\includegraphics[width=\textwidth]{img/paper_1/accuracy_comparison.png}
-		\caption{Accuracy Comparison}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
+	\begin{tabular}{lcccc}
+		\toprule
+		\textbf{Classifier} & \textbf{Accuracy} & \textbf{Precision} & \textbf{Sensitivity} & \textbf{F1-Score} \\
+		\midrule
+		Random Forest       & 0.764             & 0.777              & 0.770                & 0.766             \\
+		Gradient Boosting   & 0.702             & 0.687              & 0.689                & 0.687             \\
+		Ensemble Learning   & \textbf{0.831}    & \textbf{0.830}     & \textbf{0.820}       & \textbf{0.819}    \\
+		Bagging Classifier  & 0.753             & 0.737              & 0.746                & 0.734             \\
+		\bottomrule
+	\end{tabular}
+	\caption{ Performance Metrics of Classifiers}
+\end{table}
+
+\begin{table}[htbp]
 	\centering
-		\includegraphics[width=\textwidth]{img/paper_1/bar_chart_improved.png}
-		\caption{Improved Bar Chart}
-	\end{minipage}
 
-	\vspace{1em}
+	\begin{tabular}{lcccc}
+		\toprule
+		\textbf{Stage} & \textbf{Metric} & \textbf{Random Forest} & \textbf{Ensemble (Gradient)} & \textbf{Bagging} \\
+		\midrule
+		\textbf{W}     & Precision       & 0.86                   & \textbf{0.97}                & 0.95             \\
+		& Recall          & \textbf{0.91}          & 0.86                         & \textbf{0.97}    \\
+		\textbf{N1}    & Precision       & 0.50                   & \textbf{0.57}                & 0.57             \\
+		& Recall          & 0.59                   & \textbf{0.79}                & \textbf{0.79}    \\
+		\textbf{N2}    & Precision       & 0.48                   & \textbf{0.68}                & 0.56             \\
+		& Recall          & 0.44                   & \textbf{0.56}                & 0.41             \\
+		\textbf{REM}   & Precision       & 0.93                   & \textbf{0.98}                & 0.87             \\
+		& Recall          & 0.88                   & \textbf{0.97}                & 0.93             \\
+		\bottomrule
+	\end{tabular}
 
-	\begin{minipage}[b]{0.45\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{img/paper_1/box_plot_comparison.png}
-		\caption{Box Plot Comparison}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{img/paper_1/violin_plot_improved.png}
-		\caption{Improved Violin Plot}
-	\end{minipage}
-\end{figure}
+\end{table}

### 2025-05-06 Chapters/paper2.tex
changes

@@
 	\centering
-	\includegraphics[width=0.70\textwidth]{img/paper_2/sleep event.PNG}
+	\includegraphics[width=0.60\textwidth]{img/paper_2/sleep event.PNG}
 	\caption{Event Mapping Plot}
@@
 	i_t &= \sigma(W_i x_t + U_i h_{t-1} + b_i) \quad \text{(Input Gate)} \\
-	\tilde{c}_t &= \tanh(W_c x_t + U_c h_{t-1} + b_c) \quad \text{(Candidate Cell State)} \\
+	\tilde{c}t &= \tanh(W_c x_t + U_c h{t-1} + b_c) \quad \text{(Candidate Cell State)} \\
 	c_t &= f_t \odot c_{t-1} + i_t \odot \tilde{c}_t \quad \text{(Cell State Update)} \\
@@
 \begin{align}
-	\overrightarrow{h}_t &= \text{LSTM}(x_t, \overrightarrow{h}_{t-1}) \\
-	\overleftarrow{h}_t &= \text{LSTM}(x_t, \overleftarrow{h}_{t+1}) \\
+	\overrightarrow{h}t &= \text{LSTM}(x_t, \overrightarrow{h}{t-1}) \\
+	\overleftarrow{h}t &= \text{LSTM}(x_t, \overleftarrow{h}{t+1}) \\
 	h_t &= [\overrightarrow{h}_t; \overleftarrow{h}_t]
@@
 
-This section presents the performance evaluation of different deep learning models, using accuracy plots and confusion matrices for visual comparison.
+\subsection{Neural Network (NN) Model}
+
+\begin{figure}[H]
+	\centering
+	\begin{minipage}{0.66\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/deeplearning.png}
+		\caption{Neural Network Accuracy and Loss Comparison}
+		\label{fig:nn_accuracy}
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.33\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/NN_cm.png}
+		\caption{Neural Network Confusion Matrix}
+		\label{fig:nn_cm}
+	\end{minipage}
+\end{figure}
 
@@
 	\centering
-	\includegraphics[width=0.85\textwidth]{img/paper_2/lstm accuracy and loss plot.png}
+	\begin{minipage}{0.66\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/lstm accuracy and loss plot.png}
 		\caption{LSTM Accuracy and Loss Plot}
 		\label{fig:lstm_accuracy}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.32\textwidth}
 		\centering
-	\includegraphics[width=0.6\textwidth]{img/paper_2/LSTM_cm.png}
+		\includegraphics[width=\linewidth]{img/paper_2/LSTM_cm.png}
 		\caption{LSTM Confusion Matrix}
 		\label{fig:lstm_cm}
+	\end{minipage}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=0.85\textwidth]{img/paper_2/Rnn Accuracy Plot.png}
+	\begin{minipage}{0.66\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/Rnn Accuracy Plot.png}
 		\caption{RNN Accuracy Plot}
 		\label{fig:rnn_accuracy}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.32\textwidth}
 		\centering
-	\includegraphics[width=0.6\textwidth]{img/paper_2/RNN_cm.png}
+		\includegraphics[width=\linewidth]{img/paper_2/RNN_cm.png}
 		\caption{RNN Confusion Matrix}
 		\label{fig:rnn_cm}
+	\end{minipage}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=0.85\textwidth]{img/paper_2/BILSTM Accuaracy Plot.png}
+	\begin{minipage}{0.66\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{img/paper_2/BILSTM Accuaracy Plot.png}
 		\caption{BiLSTM Accuracy Plot}
 		\label{fig:bilstm_accuracy}
-\end{figure}
-
-\begin{figure}[H]
+	\end{minipage}
+	\hfill
+	\begin{minipage}{0.32\textwidth}
 		\centering
-	\includegraphics[width=0.6\textwidth]{img/paper_2/BiLSTM_cm.png}
+		\includegraphics[width=\linewidth]{img/paper_2/BiLSTM_cm.png}
 		\caption{BiLSTM Confusion Matrix}
 		\label{fig:bilstm_cm}
-\end{figure}
-
-\subsection{Neural Network (NN) Model}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.85\textwidth]{img/paper_2/deeplearning.png}
-	\caption{Neural Network Accuracy Comparison}
-	\label{fig:nn_accuracy}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.6\textwidth]{img/paper_2/NN_cm.png}
-	\caption{Neural Network Confusion Matrix}
-	\label{fig:nn_cm}
+	\end{minipage}
 \end{figure}
@@
 
-In this section, we compare the performance of various models, including Neural Networks, RNN, LSTM, and Bi-LSTM, based on several evaluation metrics such as accuracy, precision, recall, and F1 score.
+This section presents a comparative analysis of multiple machine learning models applied to the sleep stage classification problem.
+The evaluation metrics considered include accuracy, weighted and macro averages of precision, recall, and F1-score.
+
+\subsection{Testing  Models Overview}
+
+\begin{table}[H]
+	\centering
+	\caption{Performance Comparison of Models}
+	\label{tab:metrics_summary}
+	\begin{tabular}{lcccccc}
+		\hline
+		\textbf{Model} & \textbf{Accuracy} & \textbf{Wgt. F1} & \textbf{Macro F1} & \textbf{Macro Prec.} & \textbf{Macro Rec.} \\
+		\hline
+		Random Forest          & \textbf{0.843} & \textbf{0.823} & 0.641 & \textbf{0.901} & 0.558 \\
+		XGBoost                & \textbf{0.854} & \textbf{0.843} & 0.644 & \textbf{0.776} & 0.599 \\
+		Neural Network         & \textbf{0.777} & \textbf{0.750} & 0.487 & 0.567 & 0.452 \\
+		LSTM                   & \textbf{0.800} & \textbf{0.785} & 0.461 & 0.499 & 0.436 \\
+		BiLSTM                 & \textbf{0.811} & \textbf{0.803} & 0.472 & 0.494 & 0.457 \\
+		RNN                    & 0.706 & 0.690 & 0.332 & 0.381 & 0.345 \\
+		AdaBoost               & 0.312 & 0.353 & 0.273 & 0.323 & 0.399 \\
+		KNN (with PCA)         & 0.194 & 0.189 & 0.272 & 0.296 & 0.359 \\
+		\hline
+	\end{tabular}
+\end{table}
+
+\subsection{Discussion}
+
+Among all the models tested, the \textbf{XGBoost classifier} achieved the highest overall accuracy (0.854) and weighted F1-score (0.843), narrowly outperforming the Random Forest model.
+While Random Forest achieved a slightly higher macro precision (0.901 vs.
+0.776), its recall across minority classes such as N1 and REM was relatively lower.
+
+The deep learning-based models (LSTM and BiLSTM) followed closely behind the tree-based models, with BiLSTM slightly outperforming LSTM (macro F1 of 0.472 vs.
+0.461).
+The \textbf{Neural Network model}, although simpler, showed moderate performance (macro F1 of 0.487), better than AdaBoost and KNN.
+
+\textbf{K-Nearest Neighbors (KNN)} and \textbf{AdaBoost} performed poorly in this context.
+KNN failed to handle the class imbalance effectively, leading to very low overall accuracy (0.194).
+AdaBoost performed slightly better but still struggled to generalize, especially for majority class (Wake) classification.
+
+The \textbf{RNN model} had the lowest macro F1-score (0.332) among deep models, indicating potential vanishing gradient issues or insufficient context capture compared to LSTM-based architectures.
+
+\subsection{Conclusion}
+
+Tree-based ensemble methods (XGBoost, Random Forest) consistently outperformed others across all key metrics.
+Deep learning models, especially BiLSTM, offer competitive alternatives with relatively strong performance.
+Simpler models like KNN and AdaBoost underperform and are not suitable for this imbalanced multiclass classification task without further tuning or class balancing.
 

### 2025-05-06 Chapters/paper3.tex
changes

@@
 This model is trained on the Sleep-EDF dataset, utilizing four critical physiological signal channels: EEG Fpz-Cz, EEG Pz-Oz, EOG (horizontal), and submental EMG.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.8\linewidth]{"img/paper_3/hypnogram sleep stage transition"}
+	\caption{Hypnogram With Transition Heighlights }
+	\label{fig:hypnogram-sleep-stage-transition}
+\end{figure}
 
@@
 Subsequently, we applied a bandpass filter to isolate relevant frequency ranges.
-Specifically, the signal was filtered to retain only the 0.3–30 Hz band using the `firwin` filter design, effectively removing noise and irrelevant frequencies outside this band.
+Specifically, the signal was filtered to retain only the 0.3–30 Hz band using the firwin filter design, effectively removing noise and irrelevant frequencies outside this band.
 Though not a standard approach, this customized filtering was found to be better suited to our specific dataset and classification task.
@@
 
-The filtering was implemented using a windowed time-domain design via the `firwin` method.
+The filtering was implemented using a windowed time-domain design via the firwin method.
 A Hamming window was employed, configured with a 0.0194 passband ripple and 53 dB stopband attenuation.
@@
 Although assigning self-loop weights is possible, we chose not to include them in this implementation.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=.9\linewidth]{"img/paper_3/Graph Convolution Neural Network -SleepGCN_Transformer Architechture"}
+	\caption{Proposed Architecture of the sleep GCN- Transformer Architecture}
+	\label{fig:graph-convolution-neural-network--sleepgcntransformer-architechture}
+\end{figure}
 
@@
 
-The training process for the SleepGCN-Transformer model is designed to handle class imbalance, prevent overfitting, and improve generalization.
-Below, we outline the training methodology using key equations:
+\textbf{Fully Connected Layer:}  
+The final stage of the model includes a linear transformation layer that maps a 256-dimensional input vector to 5 output logits, corresponding to the five target sleep stages.
+This transformation enables direct classification output.
 
-\textbf{Label Smoothing:} To reduce overconfidence in predictions and improve generalization, we apply label smoothing:
-\begin{equation}
+\subsubsection*{Why Focal Loss Instead of Standard Cross-Entropy?}
+
+Standard cross-entropy loss treats all training samples equally, which can lead to a bias toward majority classes in imbalanced datasets.
+In such settings, the model tends to suppress the prediction of minority classes, resulting in poor generalization.
+Focal Loss addresses this by dynamically adjusting the contribution of each sample based on the confidence of its prediction.
+This mechanism reduces the influence of well-classified samples and increases the focus on harder, misclassified instances.
+
+Focal Loss introduces two key parameters: the focusing parameter $\gamma$, which controls the down-weighting of easy examples, and the class weighting factor $\alpha$, which helps balance class frequencies during training.
+These features make Focal Loss particularly effective for classification tasks involving significant class imbalance.
+
+\subsubsection*{Focal Loss Formulation}
+
+The Focal Loss is mathematically expressed as:
+
+\[
+FL(p_t) = -\alpha(1 - p_t)^{\gamma} \log(p_t)
+\]
+
+Here, $p_t$ denotes the predicted probability for the target class, $\alpha$ is the balancing factor to compensate for class imbalance, and $\gamma$ adjusts the focus on misclassified examples.
+
+To enhance numerical stability and improve generalization, label smoothing is applied:
+
+\[
 y_{\text{smooth}} = y(1 - \epsilon) + \frac{\epsilon}{C}
-\end{equation}
+\]
 
-\textbf{Combined Loss Function:} We incorporate both Focal Loss and smoothed labels:
-\begin{equation}
-	L = \alpha(1 - p)^\gamma (-y_{\text{smooth}} \log p)
-\end{equation}
+This prevents issues such as $\log(0)$ and ensures a more stable optimization process. The combined implementation of Focal Loss with label smoothing can be written in a PyTorch-like style as:
 
-\textbf{Focal Loss (for class imbalance):}
-\begin{equation}
-	FL(p_t) = -\alpha(1 - p_t)^\gamma \log(p_t)
-\end{equation}
+\[
+L = \alpha(1 - p)^{\gamma} \left( - y_{\text{smooth}} \log(p) \right)
+\]
 
-\textbf{Class Weight Computation:} Each class receives a weight based on its frequency:
-\begin{equation}
-	w_c = \frac{N}{\sqrt{K + 1}}
-\end{equation}
+\subsubsection*{Learning Rate Scheduling}
 
-\textbf{AdamW Optimizer:} We update parameters using the AdamW optimization rule:
-\begin{equation}
-	\theta_{t+1} = \theta_t - \eta (\nabla\phi(\theta_t) + \lambda\theta_t)
-\end{equation}
+The learning rate is a critical hyperparameter that governs the pace of model updates during training.
+A high learning rate may cause divergence, while a low learning rate can result in slow convergence or suboptimal local minima.
+Therefore, adaptive learning rate scheduling is essential for efficient training.
 
-\textbf{AdamW with Decoupled Weight Decay:}
-\begin{equation}
-	\theta_t' = \theta_t - \eta \nabla\phi(\theta_t)
-\end{equation}
-\begin{equation}
-	\theta_{t+1} = \theta_t' - \eta \lambda \theta_t
-\end{equation}
+CosineAnnealingLR is used as the scheduling strategy, as it provides a smooth and gradual reduction in the learning rate.
+It allows for large updates in the early stages of training and fine-tuning in later stages, thereby helping the model generalize better.
+The learning rate at each epoch $t$ is computed using the following cosine annealing formula:
 
-\textbf{Learning Rate Scheduling:} Cosine Annealing is used to adjust learning rates dynamically:
-\begin{equation}
-	\eta_t = \eta_{\min} + \frac{1}{2} (\eta_{\max} - \eta_{\min}) \left(1 + \cos\left(\frac{\pi T_{\text{cur}}}{T_{\text{max}}}\right)\right)
-\end{equation}
+\[
+\eta_t = \eta_{\min} + \frac{1}{2} (\eta_{\max} - \eta_{\min}) \left( 1 + \cos \left( \frac{\pi T_{\text{cur}}}{T_{\text{max}}} \right) \right)
+\]
 
-\textbf{Pipeline Summary:}
-\begin{itemize}
-	\item Load preprocessed EEG dataset for training and validation.
-	\item Compute class weights.
-	\item Apply label smoothing and Focal Loss.
-	\item Optimize with AdamW and update learning rate via cosine scheduler.
-	\item Monitor loss and accuracy metrics per epoch.
-\end{itemize}
+In this equation, $\eta_t$ is the learning rate at epoch $t$, $\eta_{\min}$ and $\eta_{\max}$ are the minimum and maximum learning rates respectively, $T_{\text{cur}}$ is the current epoch, and $T_{\text{max}}$ is the total number of epochs.
+
+\subsubsection*{Training Methodology Overview}
+
+The training pipeline is encapsulated within a dedicated class named \texttt{SleepTrainer}, which manages model training, validation, and optimization.
+The training process begins with the computation of class weights to address data imbalance.
+The model is then trained using mini-batch gradient descent, where each batch is used to compute the loss and update the model parameters accordingly.
+Performance is periodically evaluated on a separate validation set, and the learning rate is adjusted dynamically using the scheduler.
+
+\subsubsection*{Hyperparameters Used}
+
+The training setup uses a batch size of 32 and an initial learning rate of 0.0003.
+The optimizer of choice is AdamW, with a weight decay of $1 \times 10^{-4}$ to prevent overfitting.
+Training is conducted for 20 epochs, and the CosineAnnealingLR scheduler is applied to adjust the learning rate over time.
+This configuration ensures smooth convergence and robust generalization.
+
+\subsubsection*{Handling Class Imbalance}
+
+EEG sleep stage data is inherently imbalanced, with some stages appearing far more frequently than others.
+To address this, class weights are computed and incorporated into the loss function.
+The weighting formula is given by:
+
+\[
+w_c = \left( \frac{\text{Total Samples}}{\text{Class Count} + 1} \right)^{0.5}
+\]
+
+This formulation ensures that less frequent classes receive proportionally higher weights, encouraging the model to give them more attention during training.
+These computed weights are then integrated directly into the Focal Loss function to balance the influence of each class.
+
+\subsubsection*{Testing Data Distribution Analysis}
+
+Balanced testing data is crucial for an unbiased evaluation of model performance.
+An imbalanced test set can lead to skewed metrics that do not accurately reflect the model’s generalization capabilities.
+In this pipeline, the testing data is curated to ensure an even distribution across all sleep stages.
 
@@
 			\text{Weighted Precision} &= \sum_{i=1}^{C} \frac{n_i}{N} \cdot \frac{TP_i}{TP_i + FP_i} \\
-			\mathcal{L}_{CE} &= -\sum_{i=1}^{C} y_i \log(\hat{y}_i) \\
+			\mathcal{L}{CE} &= -\sum{i=1}^{C} y_i \log(\hat{y}_i) \\
 			\mathcal{L}_{FL} &= -\alpha_t (1 - p_t)^\gamma \log(p_t)
@@
 The progression of model performance across selected epochs is summarized below:
-\begin{itemize}
-	\item Epoch 1: Train Loss = 0.1915, Train Accuracy = 82.91\%; Validation Loss = 0.2110, Validation Accuracy = 83.84\%
-	\item Epoch 2: Train Loss = 0.1695, Train Accuracy = 87.74\%; Validation Loss = 0.2168, Validation Accuracy = 69.46\%
-	\item Epoch 3: Train Loss = 0.1669, Train Accuracy = 88.15\%; Validation Loss = 0.1803, Validation Accuracy = 88.39\%
-	\item Epoch 4: Train Loss = 0.1659, Train Accuracy = 88.38\%; Validation Loss = 0.1534, Validation Accuracy = 91.83\%
-	\item Epoch 5: Train Loss = 0.1621, Train Accuracy = 89.45\%; Validation Loss = 0.1615, Validation Accuracy = 89.11\%
-	\item Epoch 10: Train Loss = 0.1533, Train Accuracy = 91.03\%; Validation Loss = 0.1471, Validation Accuracy = 91.16\%
-	\item Epoch 15: Train Loss = 0.1458, Train Accuracy = 92.59\%; Validation Loss = 0.1484, Validation Accuracy = 90.98\%
-	\item Epoch 20: Train Loss = 0.1413, Train Accuracy = 93.12\%; Validation Loss = 0.1390, Validation Accuracy = 93.04\%
-\end{itemize}
 
@@
 		\label{fig:3d-acc-loss-plot}
-	\end{minipage}
-	\hfill
-	\begin{minipage}[b]{0.45\linewidth}
-		\includegraphics[width=\linewidth]{img/paper_3/3d gradient}
-		\caption{3D Gradient visualization}
-		\label{fig:3d-gradient}
 	\end{minipage}
@@
 
+This section presents the distribution of sleep stage data, which approximates a normal distribution.
+The dataset is well-balanced, with an equal number of samples across all sleep stages.
+For testing purposes, approximately 1,050 samples were selected, ensuring that each sleep stage class contains around 210 samples.
+This balanced distribution supports a fair and unbiased evaluation of the model's performance.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=0.7\linewidth]{"img/paper_3/shap important"}
+	\caption{Bar plot illustrating EEG channel importance based on SHAP values. Higher SHAP values indicate greater influence on the model's sleep stage predictions.}
+
+	\label{fig:shap-important}
+\end{figure}
+
+\subsubsection*{Interpretable Sleep Stage Classification: Hypnogram Transitions and Feature Importance using LIME and SHAP}
+
 \begin{figure}[H]
@@
 \end{figure}
+
+To improve the interpretability of the sleep stage classification model, we employ Explainable AI (XAI) techniques.
+Initially, we analyze the hypnogram transition plot, which illustrates the temporal sequence of sleep stages.
+This visualization helps assess whether the model's predictions align with the natural and expected progression of sleep cycles.
 
@@
 
+Subsequently, we utilize LIME (Local Interpretable Model-Agnostic Explanations) and SHAP (SHapley Additive exPlanations) to investigate feature importance.
+These tools provide both global and local explanations of the model's behavior.
+Specifically, they identify the EEG channels that significantly influence the model's decisions, offering insight into which features are most critical in sleep stage classification.
+
 \begin{figure}[htbp]
@@
 \end{figure}
+
+By leveraging these interpretability techniques, we ensure that the model's predictions are not only accurate but also aligned with known physiological sleep dynamics, thereby enhancing trust and transparency in the model's outcomes.
 

### 2025-05-08 Chapters/paper1.tex
changes

@@
 The output from preprocessing was then used in the feature extraction and modeling stages.
+
+\begin{algorithm}
+	\textbf{Input:} raw\_fname, annot\_fname, load\_eeg\_only, crop\_wake\_mins\\
+	\textbf{Output:} raw\\
+
+	mapping $\gets$ \{'EOG horizontal': 'eog', \dots\}\\
+	exclude $\gets$ mapping.keys() \textbf{if} load\_eeg\_only \textbf{then do nothing}\\
+	raw $\gets$ read\_raw\_edf(raw\_fname, exclude)\\
+	annots $\gets$ read\_annotations(annot\_fname)\\
+	raw.set\_annotations(annots)\\
+	\textbf{if} not load\_eeg\_only \textbf{then}\\
+	\hspace{1cm} raw.set\_channel\_types(mapping)\\
+	\textbf{if} crop\_wake\_mins > 0 \textbf{then}\\
+	\hspace{1cm} mask $\gets$ [x[-1] in \{'1','2','3','4','R'\} for x in annots.description]\\
+	\hspace{1cm} inds $\gets$ where(mask)\\
+	\hspace{1cm} tmin $\gets$ annots[inds[0]].onset - crop\_wake\_mins * 60\\
+	\hspace{1cm} tmax $\gets$ annots[inds[-1]].onset + crop\_wake\_mins * 60\\
+	\hspace{1cm} raw.crop(tmin, tmax)\\
+	rename EEG channel names\\
+	subj\_nb $\gets$ int(fname[3:5]), rec\_nb $\gets$ int(fname[5])\\
+	raw.info.subject\_info $\gets$ \{id: subj\_nb, rec\_id: rec\_nb\}\\
+	\textbf{Return} raw\\
+	\caption{load\_sleep\_physionet\_raw}
+\end{algorithm}
+
+\begin{algorithm}
+	\textbf{Input:} fnames, l\_freq, h\_freq\\
+	\textbf{Output:} raws\\
+
+	raws $\gets$ [load\_sleep\_physionet\_raw(f[0], f[1]) for f in fnames]\\
+	\textbf{for each} raw \textbf{in} raws \textbf{do}\\
+	\hspace{1cm} raw.load\_data()\\
+	\hspace{1cm} raw.filter(l\_freq, h\_freq)\\
+	\textbf{Return} raws\\
+	\caption{Load and Filter All Files}
+\end{algorithm}
+
+\begin{algorithm}
+	\textbf{Input:} raw\_data\_list, epoch\_length = 30\\
+	\textbf{Output:} epochs\_list, labels\\
+
+	epochs\_list $\gets$ [\,]\\
+	labels $\gets$ [\,]\\
+	\textbf{for each} raw \textbf{in} raw\_data\_list \textbf{do}\\
+	\hspace{1cm} sfreq $\gets$ int(raw.info['sfreq'])\\
+	\hspace{1cm} (events, event\_ids) $\gets$ events\_from\_annotations(raw)\\
+	\hspace{1cm} sleep\_ids $\gets$ [1,2,3,4,5]\\
+	\hspace{1cm} filtered\_events $\gets$ [event for event in events if event[2] in sleep\_ids]\\
+	\hspace{1cm} filtered\_event\_ids $\gets$ \{k:v for k,v in event\_ids.items() if v in sleep\_ids\}\\
+	\hspace{1cm} filtered\_events $\gets$ np.array(filtered\_events, dtype=int)\\
+	\hspace{1cm} epochs $\gets$ Epochs(raw, filtered\_events, filtered\_event\_ids, 0, epoch\_length - 1/sfreq)\\
+	\hspace{1cm} epochs\_list.append(epochs)\\
+	\hspace{1cm} labels.extend([event[2] for event in filtered\_events])\\
+	\textbf{Return} epochs\_list, labels\\
+	\caption{Epoch Creation}
+\end{algorithm}
+
+\begin{algorithm}
+	\textbf{Input:} epochs\_list, labels\\
+	\textbf{Output:} model performance metrics\\
+
+	(epochs\_list, labels) $\gets$ create\_epochs(raws)\\
+	X $\gets$ concatenate([epochs.get\_data() for epochs in epochs\_list], axis=0)\\
+	(n\_epochs, n\_channels, n\_times) $\gets$ X.shape\\
+	X $\gets$ reshape(X, (n\_epochs, n\_channels * n\_times))\\
+	y $\gets$ np.array(labels)\\
+	(X\_resampled, y\_resampled) $\gets$ SMOTE().fit\_resample(X, y)\\
+	(X\_train, X\_test, y\_train, y\_test) $\gets$ train\_test\_split(X\_resampled, y\_resampled, 0.2)\\
+	model $\gets$ initialize\_ML\_model()\\
+	model.fit(X\_train, y\_train)\\
+	y\_pred $\gets$ model.predict(X\_test)\\
+	report $\gets$ classification\_report(y\_test, y\_pred)\\
+	cm $\gets$ confusion\_matrix(y\_test, y\_pred)\\
+	display(report)\\
+	display\_confusion\_matrix(cm)\\
+	\caption{ML Training and Evaluation Pipeline}
+\end{algorithm}
+
+	\begin{algorithm}
+		\textbf{Input:} list of (signal\_file, annotation\_file) pairs \\
+		\textbf{Output:} average accuracy, classification report, confusion matrix
+
+		\vspace{0.3cm}
+		\texttt{epochs\_list = []} \\
+		\texttt{for file\_pair in fnames:} \\
+		\hspace*{0.5cm}\texttt{raw = read\_raw\_edf(file\_pair[0])} \\
+		\hspace*{0.5cm}\texttt{annot = read\_annotations(file\_pair[1])} \\
+		\hspace*{0.5cm}\texttt{raw.set\_annotations(annot)} \\
+		\hspace*{0.5cm}\texttt{annot.crop(annot[1].onset - 1800, annot[-2].onset + 1800)} \\
+		\hspace*{0.5cm}\texttt{events = events\_from\_annotations(raw)} \\
+		\hspace*{0.5cm}\texttt{epochs = Epochs(raw, events)} \\
+		\hspace*{0.5cm}\texttt{epochs\_list.append(epochs)} \\
+
+		\vspace{0.3cm}
+		\texttt{all\_epochs = concatenate\_epochs(epochs\_list)} \\
+		\texttt{y = all\_epochs.events[:, 2]} \\
+
+		\vspace{0.3cm}
+		\textbf{Define feature extraction:} \\
+		\texttt{def eeg\_power\_band(epochs):} \\
+		\hspace*{0.5cm}\texttt{psds = compute\_psd(epochs)} \\
+		\hspace*{0.5cm}\texttt{psds = normalize(psds)} \\
+		\hspace*{0.5cm}\texttt{features = []} \\
+		\hspace*{0.5cm}\texttt{for band in freq\_bands:} \\
+		\hspace*{1cm}\texttt{features.append(mean\_psd\_in\_band)} \\
+		\hspace*{0.5cm}\texttt{return concatenate(features)} \\
+
+		\vspace{0.3cm}
+		\textbf{Train ML pipeline with KFold:} \\
+		\texttt{pipeline = [eeg\_power\_band, Classifier()]} \\
+		\texttt{kf = KFold(n\_splits=5)} \\
+		\texttt{results = \{\}} \\
+		\texttt{for train\_idx, test\_idx in kf.split(all\_epochs):} \\
+		\hspace*{0.5cm}\texttt{X\_train = all\_epochs[train\_idx]} \\
+		\hspace*{0.5cm}\texttt{X\_test = all\_epochs[test\_idx]} \\
+		\hspace*{0.5cm}\texttt{y\_train = y[train\_idx]} \\
+		\hspace*{0.5cm}\texttt{y\_test = y[test\_idx]} \\
+		\hspace*{0.5cm}\texttt{pipeline.fit(X\_train, y\_train)} \\
+		\hspace*{0.5cm}\texttt{y\_pred = pipeline.predict(X\_test)} \\
+		\hspace*{0.5cm}\texttt{store accuracy, y\_true, y\_pred} \\
+
+		\caption{General ML Pipeline with K-Fold Cross-Validation}
+	\end{algorithm}
+
+ \RestyleAlgo{ruled}
+ \SetKwComment{Comment}{/* }{ */}
+ \DontPrintSemicolon
+
+ \begin{algorithm*}[!t]
+ 	\caption{Main Pipeline for Sleep Stage Classification}
+ 	\label{alg:main}
+
+ 	\KwData{Raw EEG and annotation file paths}
+ 	\KwResult{Trained SleepGCN\_Transformer model}
+
+ 	fnames $\gets$ \texttt{[('SC4001E0-PSG.edf', 'SC4001EC-Hypnogram.edf')]} \;
+ 	$(X_{\text{all}}, Y_{\text{all}}) \gets$ \texttt{preprocess\_data(fnames)} \;
+ 	$X_{\text{tensor}} \gets \texttt{torch.tensor}(X_{\text{all}}, \texttt{float32})$ \;
+ 	$Y_{\text{tensor}} \gets \texttt{torch.tensor}(Y_{\text{all}}, \texttt{long})$ \;
+
+ 	dataset $\gets$ \texttt{SleepEEGDataset}($X_{\text{tensor}}, Y_{\text{tensor}}$) \;
+ 	$train\_size \gets 0.8 \times \texttt{len}(dataset)$ \;
+ 	$val\_size \gets \texttt{len}(dataset) - train\_size$ \;
+
+ 	$(train\_dataset, val\_dataset) \gets$ \texttt{random\_split}(dataset, [train\_size, val\_size]) \;
+
+ 	model $\gets$ \texttt{SleepGCN\_Transformer}(\;
+ 	\Indp
+ 	input\_dim=4, hidden\_dim=256, output\_dim=5,\;
+ 	num\_layers=5, nhead=4, dropout=0.1)\;
+ 	\Indm
+
+ 	trainer $\gets$ \texttt{SleepTrainer}(\;
+ 	\Indp
+ 	model, train\_dataset, val\_dataset,\;
+ 	batch\_size=32, lr=0.0003,\;
+ 	num\_epochs=20, weight\_decay=1e-4)\;
+ 	\Indm
+
+ 	\texttt{trainer.train()} \;
+
+ \end{algorithm*}
+
+ \RestyleAlgo{ruled}
+
+ %% This is needed if you want to add comments in
+ %% your algorithm with \Comment
+ \SetKwComment{Comment}{/* }{ */}
+
+ \begin{algorithm*}[!t]
+ 	\caption{SleepEEGDataset Class Definition}
+ 	\label{alg:sleepdataset}
+
+ 	\KwData{Input features $X$, labels $Y$}
+ 	\KwResult{A graph-based dataset with node features and edge structure}
+
+ 	\textbf{class} SleepEEGDataset(Dataset): \\
+ 	\Indp
+ 	\textbf{def} \_\_init\_\_(self, X, Y): \\
+ 	\Indp
+ 	self.X = X \\
+ 	self.Y = Y \\
+ 	A = np.array([ \\
+
+ 	]) \\
+ 	edge\_index = torch.tensor(np.array(np.nonzero(A)), dtype=torch.long) \\
+ 	edge\_weights = torch.tensor(A[A.nonzero()], dtype=torch.float32) \\
+ 	self.edge\_index = edge\_index \\
+ 	self.edge\_weights = edge\_weights \\
+ 	\Indm
+
+ 	\BlankLine
+
+ 	\textbf{def} len(self): \\
+ 	\Indp
+ 	\textbf{return} len(self.X) \\
+ 	\Indm
+
+ 	\BlankLine
+
+ 	\textbf{def} get(self, idx): \\
+ 	\Indp
+
+ 	label = torch.tensor([self.Y[idx]], dtype=torch.long) \\
+ 	\textbf{return} Data(x=node\_features, edge\_index=self.edge\_index, \\
+ 	\hspace{6em}edge\_attr=self.edge\_weights, y=label) \\
+ 	\Indm
+ \end{algorithm*}
+
+ \RestyleAlgo{ruled}
+ \SetKwComment{Comment}{/* }{ */}
+
+ \begin{algorithm*}[!t]
+ 	\caption{Preprocess Sleep EEG Data}\label{alg:preprocess_data}
+ 	\KwData{List of file pairs: $(psg\_file, ann\_file)$}
+ 	\KwResult{Preprocessed feature matrix $X$ and labels $Y$}
+ 	$X\_all \gets []$\;
+ 	$Y\_all \gets []$\;
+ 	\ForEach{$(psg\_file, ann\_file)$ in fnames}{ 
+ 		$raw \gets \text{mne.io.read\_raw\_edf}(psg\_file, preload=True)$\;
+ 		$annotations \gets \text{mne.read\_annotations}(ann\_file)$\;
+ 		$raw.set\_annotations(annotations)$\;
+ 		$raw.pick\_channels(channels)$\;
+ 		$raw.filter(0.3, 30., fir\_design='firwin')$\;
+ 		$data \gets raw.get\_data()$\;
+ 		$sfreq \gets raw.info['sfreq']$\;
+
+ 		$epochs\_list \gets []$\;
+ 		$labels\_list \gets []$\;
+
+ 		\ForEach{$annot$ in annotations}{
+ 			$stage \gets annot['description']$\;
+ 			\If{$stage \in stage\_mapping$}{
+ 				$onset\_sample \gets \text{int}(annot['onset'] \times sfreq)$\;
+ 				$duration\_samples \gets \text{int}(annot['duration'] \times sfreq)$\;
+ 				$epoch\_length\_samples \gets \text{int}(30 \times sfreq)$\;
+
+ 				\For{$i \gets 0$ \KwTo $duration\_samples$ with step $epoch\_length\_samples$}{
+ 					$start \gets onset\_sample + i$\;
+ 					$stop \gets start + epoch\_length\_samples$\;
+ 					\If{$stop > data.shape[1]$}{
+ 						\textbf{break}\;
+ 					}
+ 					$epochs\_list.append(data[:, start:stop])$\;
+ 					$labels\_list.append(stage\_mapping[stage])$\;
+ 				}
+ 			}
+ 		}
+
+ 		$X \gets \text{np.array}(epochs\_list)$\;
+ 		$Y \gets \text{np.array}(labels\_list)$\;
+ 		$X\_all.append(X)$\;
+ 		$Y\_all.append(Y)$\;
+ 	}
+ 	$X\_all \gets \texttt{np.concatenate}(X\_all, axis=0)$\;
+ 	$Y\_all \gets \texttt{np.concatenate}(Y\_all, axis=0)$\;
+ 	\KwRet{$X\_all, Y\_all$}
+ \end{algorithm*}
+
+ \RestyleAlgo{ruled}
+ \SetKwComment{Comment}{/* }{ */}
+
+ \begin{algorithm*}[!t]
+ 	\caption{Training Pipeline for Sleep EEG Model}\label{alg:training_pipeline}
+ 	\KwData{Training and Validation datasets, Model, Learning rate, Epochs, Batch size, Weight decay}
+ 	\KwResult{Trained model with performance metrics on training and validation data}
+ 	\SetKwFunction{FocalLoss}{FocalLoss}
+ 	\SetKwFunction{SleepTrainer}{SleepTrainer}
+ 	\SetKwFunction{Train}{train}
+ 	\SetKwFunction{GetClassWeights}{get\_class\_weights}
+ 	\SetKwFunction{ComputeAccuracy}{compute\_accuracy}
+
+ 	\BlankLine
+
+ 	\SetKwProg{Fn}{Function}{:}{}
+ 	\Fn{\FocalLoss{logits, targets}}{
+ 		probs $\gets$ \texttt{F.softmax(logits, dim=-1)}\;
+ 		targets\_one\_hot $\gets$ \texttt{F.one\_hot(targets, num\_classes=logits.shape[1]).float()}\;
+ 		targets\_one\_hot $\gets$ targets\_one\_hot * (1 - \texttt{label\_smoothing}) + \texttt{label\_smoothing} / logits.shape[1]\;
+ 		ce\_loss $\gets$ -targets\_one\_hot * \texttt{torch.log(probs + 1e-9)}\;
+ 		focal\_loss $\gets$ \texttt{alpha * (1 - probs) ** gamma * ce\_loss}\;
+ 		\KwRet \texttt{focal\_loss.mean()}
+ 	}
+
+ 	\Fn{\SleepTrainer{model, train\_dataset, val\_dataset, batch\_size, lr, num\_epochs, weight\_decay}}{
+ 		class\_weights $\gets$ \texttt{self.get\_class\_weights(train\_dataset).to(device)}\;
+ 		loss\_fn $\gets$ \texttt{FocalLoss(alpha=class\_weights, gamma=3, label\_smoothing=0.1)}\;
+ 		optimizer $\gets$ \texttt{optim.AdamW(model.parameters(), lr=lr, weight\_decay=weight\_decay)}\;
+ 		scheduler $\gets$ \texttt{optim.lr\_scheduler.CosineAnnealingLR(optimizer, T\_max=num\_epochs)}\;
+ 		train\_loader $\gets$ \texttt{DataLoader(train\_dataset, batch\_size=batch\_size, shuffle=True)}\;
+ 		val\_loader $\gets$ \texttt{DataLoader(val\_dataset, batch\_size=batch\_size, shuffle=False)}\;
+
+ 		\For{$epoch \gets 1$ \KwTo num\_epochs}{
+ 			model.train()\;
+ 			total\_train\_loss, total\_train\_accuracy $\gets$ 0, 0\;
+
+ 			\ForEach{$batch$ in train\_loader}{
+ 				batch $\gets$ batch.to(device)\;
+ 				optimizer.zero\_grad()\;
+ 				output $\gets$ model(batch)\;
+ 				target $\gets$ batch.y\;
+ 				loss $\gets$ loss\_fn(output, target)\;
+ 				loss.backward()\;
+ 				optimizer.step()\;
+ 				total\_train\_loss += loss.item()\;
+ 				total\_train\_accuracy += compute\_accuracy(output, target)\;
+ 			}
+
+ 			scheduler.step()\;
+
+ 			\textbf{Validation Phase:}\;
+ 			model.eval()\;
+ 			total\_val\_loss, total\_val\_accuracy $\gets$ 0, 0\;
+ 			\ForEach{$batch$ in val\_loader}{
+ 				batch $\gets$ batch.to(device)\;
+ 				output $\gets$ model(batch)\;
+ 				target $\gets$ batch.y\;
+ 				loss $\gets$ loss\_fn(output, target)\;
+ 				total\_val\_loss += loss.item()\;
+ 				total\_val\_accuracy += compute\_accuracy(output, target)\;
+ 			}
+
+ 			avg\_train\_loss $\gets$ total\_train\_loss / len(train\_loader)\;
+ 			avg\_train\_accuracy $\gets$ total\_train\_accuracy / len(train\_loader)\;
+ 			avg\_val\_loss $\gets$ total\_val\_loss / len(val\_loader)\;
+ 			avg\_val\_accuracy $\gets$ total\_val\_accuracy / len(val\_loader)\;
+
+ 			print(f"Epoch {epoch}/{num\_epochs}: Train Loss: {avg\_train\_loss:.4f}, Train Acc: {avg\_train\_accuracy:.4f} | Val Loss: {avg\_val\_loss:.4f}, Val Acc: {avg\_val\_accuracy:.4f}")
+ 		}
+ 	}
+ 	\KwRet{Trained Model}
+ \end{algorithm*}
+
+ \RestyleAlgo{ruled}
+ \SetKwComment{Comment}{/* }{ */}
+
+ \begin{algorithm*}[!t]
+ 	\caption{SleepGCN\_Transformer Model Declaration}\label{alg:sleep_gcn_transformer}
+ 	\KwData{Input features of shape (batch size, input\_dim), graph structure with edges (edge\_index), batch information (batch)}
+ 	\KwResult{Predicted outputs for each input, of shape (batch size, output\_dim)}
+ 	\SetKwFunction{SleepGCNTransformer}{SleepGCN\_Transformer}
+ 	\SetKwFunction{Forward}{forward}
+
+ 	\BlankLine
+
+ 	\SetKwProg{Fn}{Function}{:}{}
+ 	\Fn{\SleepGCNTransformer{input\_dim, hidden\_dim, output\_dim, num\_layers, nhead, dropout}}{
+ 		\texttt{gcn\_layers} $\gets$ \texttt{nn.ModuleList()}\;
+ 		\texttt{batch\_norms} $\gets$ \texttt{nn.ModuleList()}\;
+ 		\texttt{gcn\_layers.append(GCNConv(input\_dim, hidden\_dim))}\;
+ 		\texttt{batch\_norms.append(BatchNorm(hidden\_dim))}\;
+ 		\For{$i \gets 1$ \KwTo num\_layers - 1}{
+ 			\texttt{gcn\_layers.append(GCNConv(hidden\_dim, hidden\_dim))}\;
+ 			\texttt{batch\_norms.append(BatchNorm(hidden\_dim))}\;
+ 		}
+ 		\texttt{encoder\_layer} $\gets$ \texttt{TransformerEncoderLayer(d\_model=hidden\_dim, nhead=nhead, dropout=dropout, batch\_first=True)}\;
+ 		\texttt{transformer} $\gets$ \texttt{TransformerEncoder(encoder\_layer, num\_layers=2)}\;
+ 		\texttt{fc} $\gets$ \texttt{nn.Linear(hidden\_dim, output\_dim)}\;
+ 		\texttt{activation} $\gets$ \texttt{ReLU()}\;
+ 		\texttt{dropout} $\gets$ \texttt{Dropout(dropout)}\;
+ 	}
+
+ 	\BlankLine
+
+ 	\Fn{\Forward{data}}{
+ 		\texttt{X, edge\_index, batch} $\gets$ \texttt{data.x, data.edge\_index, data.batch}\;
+ 		\For{$i \gets 0$ \KwTo len(gcn\_layers) - 1}{
+ 			\texttt{X} $\gets$ \texttt{activation(gcn\_layers[i](X, edge\_index))}\;
+ 			\texttt{X} $\gets$ \texttt{batch\_norms[i](X)}\;
+ 			\texttt{X} $\gets$ \texttt{dropout(X)}\;
+ 		}
+ 		\texttt{X} $\gets$ \texttt{global\_mean\_pool(X, batch)}\;
+ 		\texttt{X} $\gets$ \texttt{X.unsqueeze(1)}\;
+ 		\texttt{X} $\gets$ \texttt{transformer(X)}\;
+ 		\texttt{X} $\gets$ \texttt{X.squeeze(1)}\;
+ 		\KwRet \texttt{fc(X)}\;
+ 	}
+ \end{algorithm*}
 

### 2025-05-08 Chapters/paper1.tex
changes

--- 2025-05-06 Chapters/paper1.tex
@@
 The output from preprocessing was then used in the feature extraction and modeling stages.
+
+\begin{algorithm}
+	\textbf{Input:} raw\_fname, annot\_fname, load\_eeg\_only, crop\_wake\_mins\\
+	\textbf{Output:} raw\\
+
+	mapping $\gets$ \{'EOG horizontal': 'eog', \dots\}\\
+	exclude $\gets$ mapping.keys() \textbf{if} load\_eeg\_only \textbf{then do nothing}\\
+	raw $\gets$ read\_raw\_edf(raw\_fname, exclude)\\
+	annots $\gets$ read\_annotations(annot\_fname)\\
+	raw.set\_annotations(annots)\\
+	\textbf{if} not load\_eeg\_only \textbf{then}\\
+	\hspace{1cm} raw.set\_channel\_types(mapping)\\
+	\textbf{if} crop\_wake\_mins > 0 \textbf{then}\\
+	\hspace{1cm} mask $\gets$ [x[-1] in \{'1','2','3','4','R'\} for x in annots.description]\\
+	\hspace{1cm} inds $\gets$ where(mask)\\
+	\hspace{1cm} tmin $\gets$ annots[inds[0]].onset - crop\_wake\_mins * 60\\
+	\hspace{1cm} tmax $\gets$ annots[inds[-1]].onset + crop\_wake\_mins * 60\\
+	\hspace{1cm} raw.crop(tmin, tmax)\\
+	rename EEG channel names\\
+	subj\_nb $\gets$ int(fname[3:5]), rec\_nb $\gets$ int(fname[5])\\
+	raw.info.subject\_info $\gets$ \{id: subj\_nb, rec\_id: rec\_nb\}\\
+	\textbf{Return} raw\\
+	\caption{load\_sleep\_physionet\_raw}
+\end{algorithm}
+
+\begin{algorithm}
+	\textbf{Input:} fnames, l\_freq, h\_freq\\
+	\textbf{Output:} raws\\
+
+	raws $\gets$ [load\_sleep\_physionet\_raw(f[0], f[1]) for f in fnames]\\
+	\textbf{for each} raw \textbf{in} raws \textbf{do}\\
+	\hspace{1cm} raw.load\_data()\\
+	\hspace{1cm} raw.filter(l\_freq, h\_freq)\\
+	\textbf{Return} raws\\
+	\caption{Load and Filter All Files}
+\end{algorithm}
+
+\begin{algorithm}
+	\textbf{Input:} raw\_data\_list, epoch\_length = 30\\
+	\textbf{Output:} epochs\_list, labels\\
+
+	epochs\_list $\gets$ [\,]\\
+	labels $\gets$ [\,]\\
+	\textbf{for each} raw \textbf{in} raw\_data\_list \textbf{do}\\
+	\hspace{1cm} sfreq $\gets$ int(raw.info['sfreq'])\\
+	\hspace{1cm} (events, event\_ids) $\gets$ events\_from\_annotations(raw)\\
+	\hspace{1cm} sleep\_ids $\gets$ [1,2,3,4,5]\\
+	\hspace{1cm} filtered\_events $\gets$ [event for event in events if event[2] in sleep\_ids]\\
+	\hspace{1cm} filtered\_event\_ids $\gets$ \{k:v for k,v in event\_ids.items() if v in sleep\_ids\}\\
+	\hspace{1cm} filtered\_events $\gets$ np.array(filtered\_events, dtype=int)\\
+	\hspace{1cm} epochs $\gets$ Epochs(raw, filtered\_events, filtered\_event\_ids, 0, epoch\_length - 1/sfreq)\\
+	\hspace{1cm} epochs\_list.append(epochs)\\
+	\hspace{1cm} labels.extend([event[2] for event in filtered\_events])\\
+	\textbf{Return} epochs\_list, labels\\
+	\caption{Epoch Creation}
+\end{algorithm}
+
+\begin{algorithm}
+	\textbf{Input:} epochs\_list, labels\\
+	\textbf{Output:} model performance metrics\\
+
+	(epochs\_list, labels) $\gets$ create\_epochs(raws)\\
+	X $\gets$ concatenate([epochs.get\_data() for epochs in epochs\_list], axis=0)\\
+	(n\_epochs, n\_channels, n\_times) $\gets$ X.shape\\
+	X $\gets$ reshape(X, (n\_epochs, n\_channels * n\_times))\\
+	y $\gets$ np.array(labels)\\
+	(X\_resampled, y\_resampled) $\gets$ SMOTE().fit\_resample(X, y)\\
+	(X\_train, X\_test, y\_train, y\_test) $\gets$ train\_test\_split(X\_resampled, y\_resampled, 0.2)\\
+	model $\gets$ initialize\_ML\_model()\\
+	model.fit(X\_train, y\_train)\\
+	y\_pred $\gets$ model.predict(X\_test)\\
+	report $\gets$ classification\_report(y\_test, y\_pred)\\
+	cm $\gets$ confusion\_matrix(y\_test, y\_pred)\\
+	display(report)\\
+	display\_confusion\_matrix(cm)\\
+	\caption{ML Training and Evaluation Pipeline}
+\end{algorithm}
+
+	\begin{algorithm}
+		\textbf{Input:} list of (signal\_file, annotation\_file) pairs \\
+		\textbf{Output:} average accuracy, classification report, confusion matrix
+
+		\vspace{0.3cm}
+		\texttt{epochs\_list = []} \\
+		\texttt{for file\_pair in fnames:} \\
+		\hspace*{0.5cm}\texttt{raw = read\_raw\_edf(file\_pair[0])} \\
+		\hspace*{0.5cm}\texttt{annot = read\_annotations(file\_pair[1])} \\
+		\hspace*{0.5cm}\texttt{raw.set\_annotations(annot)} \\
+		\hspace*{0.5cm}\texttt{annot.crop(annot[1].onset - 1800, annot[-2].onset + 1800)} \\
+		\hspace*{0.5cm}\texttt{events = events\_from\_annotations(raw)} \\
+		\hspace*{0.5cm}\texttt{epochs = Epochs(raw, events)} \\
+		\hspace*{0.5cm}\texttt{epochs\_list.append(epochs)} \\
+
+		\vspace{0.3cm}
+		\texttt{all\_epochs = concatenate\_epochs(epochs\_list)} \\
+		\texttt{y = all\_epochs.events[:, 2]} \\
+
+		\vspace{0.3cm}
+		\textbf{Define feature extraction:} \\
+		\texttt{def eeg\_power\_band(epochs):} \\
+		\hspace*{0.5cm}\texttt{psds = compute\_psd(epochs)} \\
+		\hspace*{0.5cm}\texttt{psds = normalize(psds)} \\
+		\hspace*{0.5cm}\texttt{features = []} \\
+		\hspace*{0.5cm}\texttt{for band in freq\_bands:} \\
+		\hspace*{1cm}\texttt{features.append(mean\_psd\_in\_band)} \\
+		\hspace*{0.5cm}\texttt{return concatenate(features)} \\
+
+		\vspace{0.3cm}
+		\textbf{Train ML pipeline with KFold:} \\
+		\texttt{pipeline = [eeg\_power\_band, Classifier()]} \\
+		\texttt{kf = KFold(n\_splits=5)} \\
+		\texttt{results = \{\}} \\
+		\texttt{for train\_idx, test\_idx in kf.split(all\_epochs):} \\
+		\hspace*{0.5cm}\texttt{X\_train = all\_epochs[train\_idx]} \\
+		\hspace*{0.5cm}\texttt{X\_test = all\_epochs[test\_idx]} \\
+		\hspace*{0.5cm}\texttt{y\_train = y[train\_idx]} \\
+		\hspace*{0.5cm}\texttt{y\_test = y[test\_idx]} \\
+		\hspace*{0.5cm}\texttt{pipeline.fit(X\_train, y\_train)} \\
+		\hspace*{0.5cm}\texttt{y\_pred = pipeline.predict(X\_test)} \\
+		\hspace*{0.5cm}\texttt{store accuracy, y\_true, y\_pred} \\
+
+		\caption{General ML Pipeline with K-Fold Cross-Validation}
+	\end{algorithm}
 

### 2025-05-08 Chapters/paper3.tex
changes

@@
 These settings were carefully chosen to preserve the signal components relevant to sleep staging while minimizing noise.
+
+\begin{algorithm}[H]
+	\caption{Main Pipeline for Sleep Stage Classification}
+	\label{alg:main}
+	\textbf{Input:} list of (PSG file, Hypnogram file) pairs\\
+	\textbf{Output:} trained SleepGCN\_Transformer model
+	\begin{enumerate}
+		\item fnames $\leftarrow$ [('SC4001E0-PSG.edf','SC4001EC-Hypnogram.edf')]\label{step:fnames}
+		\item $(X_{\mathrm{all}},Y_{\mathrm{all}}) \leftarrow \texttt{preprocess\_data}(fnames)$
+		\item $X_{\mathrm{tensor}} \leftarrow \texttt{to\_tensor}(X_{\mathrm{all}})$; \quad
+		$Y_{\mathrm{tensor}} \leftarrow \texttt{to\_tensor}(Y_{\mathrm{all}})$
+		\item dataset $\leftarrow$ \texttt{SleepEEGDataset}($X_{\mathrm{tensor}},Y_{\mathrm{tensor}}$)
+		\item train\_size $\leftarrow 0.8\times |\text{dataset}|$; \quad
+		val\_size $\leftarrow |\text{dataset}| - \text{train\_size}$
+		\item $(\text{train\_ds},\text{val\_ds}) \leftarrow \texttt{split}(dataset,\,[\text{train\_size},\text{val\_size}])$
+		\item model $\leftarrow \texttt{init\_model}(4,256,5,5,4,0.1)$
+		\item trainer $\leftarrow \texttt{init\_trainer}(model,\text{train\_ds},\text{val\_ds},32,20,1e-4)$
+		\item \texttt{trainer.train()}
+	\end{enumerate}
+\end{algorithm}
+
+\begin{algorithm}[H]
+	\caption{SleepEEGDataset Class Definition}
+	\label{alg:sleepdataset}
+	\textbf{Input:} feature matrix $X$, labels $Y$\\
+	\textbf{Output:} graph dataset with node features and edges
+	\begin{enumerate}
+		\item \textbf{class} SleepEEGDataset
+		\begin{enumerate}
+			\item \_\_init\_\_(self,$X,Y$):
+			\begin{enumerate}
+				\item self.X$\leftarrow X$; self.Y$\leftarrow Y$
+				\item $A\leftarrow\texttt{build\_adjacency}(X)$
+				\item self.edge\_index$\leftarrow \texttt{find\_edges}(A)$
+				\item self.edge\_weights$\leftarrow \texttt{edge\_weights}(A)$
+			\end{enumerate}
+			\item len(self): \textbf{return} $|\,$self.X$|$
+			\item get(self,idx):
+			\begin{enumerate}
+				\item node\_feat$\leftarrow$ self.X[idx]
+				\item label$\leftarrow$ self.Y[idx]
+				\item \textbf{return} (node\_feat,self.edge\_index,self.edge\_weights,label)
+			\end{enumerate}
+		\end{enumerate}
+	\end{enumerate}
+\end{algorithm}
+
+\begin{algorithm}[H]
+	\caption{Preprocess Sleep EEG Data}
+	\label{alg:preprocess_data}
+	\textbf{Input:} list of (psg\_file, ann\_file)\\
+	\textbf{Output:} feature matrix $X_{\mathrm{all}}$, labels $Y_{\mathrm{all}}$
+	\begin{enumerate}
+		\item $X_{\mathrm{all}}\leftarrow\emptyset$, $Y_{\mathrm{all}}\leftarrow\emptyset$
+		\item \textbf{for each} $(p,a)$ in fnames:
+		\begin{enumerate}
+			\item raw$\leftarrow$ load PSG from $p$
+			\item annots$\leftarrow$ load annotations from $a$
+			\item raw$\leftarrow$ filter and preprocess(raw)
+			\item $E\leftarrow$ extract epochs(raw,annots)
+			\item $L\leftarrow$ extract labels(annots)
+			\item $X_{\mathrm{all}}\leftarrow X_{\mathrm{all}}\cup E$
+			\item $Y_{\mathrm{all}}\leftarrow Y_{\mathrm{all}}\cup L$
+		\end{enumerate}
+		\item \textbf{return} $X_{\mathrm{all}},Y_{\mathrm{all}}$
+	\end{enumerate}
+\end{algorithm}
+
+\begin{algorithm}[H]
+	\caption{Training Pipeline for Sleep EEG Model}
+	\label{alg:training_pipeline}
+	\textbf{Input:} training set, validation set, model, epochs, batch size \\
+	\textbf{Output:} trained model, performance metrics
+	\begin{enumerate}
+		\item loss\_fn$\leftarrow$ define loss()
+		\item optimizer$\leftarrow$ init optimizer(model)
+		\item scheduler$\leftarrow$ init scheduler(optimizer,epochs)
+		\item \textbf{for} epoch = 1 to epochs:
+		\begin{enumerate}
+			\item model.train()
+			\item \textbf{for each} batch in train\_ds:
+			\begin{enumerate}
+				\item output$\leftarrow$model(batch)
+				\item loss$\leftarrow$loss\_fn(output, batch.y)
+				\item backpropagate and step optimizer
+			\end{enumerate}
+			\item scheduler.step()
+			\item model.eval()
+			\item \textbf{for each} batch in val\_ds:
+			\begin{enumerate}
+				\item output$\leftarrow$model(batch)
+				\item compute validation metrics
+			\end{enumerate}
+		\end{enumerate}
+		\item \textbf{return} trained model, metrics
+	\end{enumerate}
+\end{algorithm}
+
+\begin{algorithm}[H]
+	\caption{SleepGCN\_Transformer Model Declaration}
+	\label{alg:sleep_gcn_transformer}
+	\textbf{Input:} input\_dim, hidden\_dim, output\_dim, num\_layers, nhead, dropout \\
+	\textbf{Output:} network layers initialized
+	\begin{enumerate}
+		\item Initialize list of GCN layers and batch norms
+		\item \textbf{for} $i$ in $1\ldots$num\_layers:
+		\begin{enumerate}
+			\item append GCNConv layer
+			\item append BatchNorm layer
+		\end{enumerate}
+		\item Initialize Transformer encoder with nhead and dropout
+		\item Initialize final fully‐connected layer
+		\item \textbf{return} model components
+	\end{enumerate}
+\end{algorithm}
 

### 2025-05-08 Title/conclusion.tex
changes

@@
-Our proposed SleepGCN-Transformer model achieves 93.12\% training accuracy and
-93.04\% validation accuracy, demonstrating its effectiveness in sleep stage classification.
-The integration of Graph Convolution Networks (GCN) captures spatial dependencies
-across EEG, EOG, and EMG channels, while the Transformer extracts temporal patterns. The
-use of Focal Loss enhances class balancing, improving performance on underrepresented sleep
-stages. Feature importance analysis highlights EMG and EEG Pz-Oz as key predictors. This
-robust approach lays the foundation for future work in Explainable AI, enabling medical
-professionals to interpret AI-driven sleep diagnostics effectively.
+\section{Discussion}
+
+\subsection{Principal Contributions}
+This study proposes a novel graph-based representation for sleep EEG data, integrating a hybrid GCN-Transformer architecture to improve the classification of sleep stages.
+By effectively capturing both spatial and temporal dependencies, our model outperforms traditional deep learning approaches in terms of accuracy and robustness.
+The incorporation of graph structures, combined with attention mechanisms, facilitates a more nuanced understanding of sleep dynamics.
+
+\subsection{Distinctive Aspects of Our Approach}
+Conventional methods such as CNNs primarily emphasize spatial features, whereas RNNs address temporal dynamics but are limited in handling long-range dependencies.
+Our model bridges this gap by combining the spatial modeling capabilities of Graph Convolutional Networks (GCNs) with the sequence modeling strengths of Transformers.
+Moreover, the use of Focal Loss and Label Smoothing addresses class imbalance effectively.
+The explainability component of our framework, supported by XAI techniques, identifies key EEG channels that influence sleep stage transitions, enhancing interpretability and clinical relevance.
+
+\subsection{Performance Highlights}
+Several components contribute to the model’s success.
+The graph-based representation enables the retention of spatial relationships across EEG nodes, promoting more effective learning of brain activity patterns.
+The fusion of GCN and Transformer modules allows the network to model complex spatial-temporal interactions, yielding highly accurate sleep stage predictions.
+Techniques for mitigating class imbalance—such as Focal Loss and Label Smoothing—enhance reliability, especially in underrepresented stages.
+Additionally, explainability features provide actionable insights into the neural correlates of sleep, supporting broader medical and neuroscientific applications.
+
+\subsection{Limitations and Future Directions}
+Despite its strengths, the proposed method has notable limitations.
+The combined GCN-Transformer model incurs higher computational costs compared to lighter architectures like CNNs or LSTMs, which may restrict deployment in resource-constrained environments.
+The Sleep-EDF dataset, although widely used, includes a limited sample of subjects, potentially affecting generalizability.
+Furthermore, while acceleration via Apple M1 and Metal APIs improves efficiency, larger models would benefit from more powerful hardware such as GPUs or TPUs.
+Future work may explore more scalable architectures and validate the approach on larger, more diverse datasets.
+
+\section{Conclusion}
+
+The proposed SleepGCN-Transformer architecture demonstrates superior performance in sleep stage classification, achieving 93.12\% training accuracy and 93.04\% validation accuracy.
+By integrating GCNs to model spatial interdependencies and Transformers to extract temporal patterns, the model effectively captures the complexity of sleep physiology.
+The application of Focal Loss contributes to robust handling of class imbalance, improving classification across all stages.
+Feature importance analysis indicates that EMG and the EEG Pz-Oz channel play a critical role in prediction.
+This approach not only improves predictive performance but also enhances interpretability, laying a foundation for future work in explainable and clinically applicable sleep analysis tools.

### 2025-05-08 Chapters/discussion.tex
changes

--- 2025-05-08 Title/conclusion.tex
@@
-\section{Discussion}
-
 \subsection{Principal Contributions}
@@
 Furthermore, while acceleration via Apple M1 and Metal APIs improves efficiency, larger models would benefit from more powerful hardware such as GPUs or TPUs.
-Future work may explore more scalable architectures and validate the approach on larger, more diverse datasets.
-
-\section{Conclusion}
-
-The proposed SleepGCN-Transformer architecture demonstrates superior performance in sleep stage classification, achieving 93.12\% training accuracy and 93.04\% validation accuracy.
-By integrating GCNs to model spatial interdependencies and Transformers to extract temporal patterns, the model effectively captures the complexity of sleep physiology.
-The application of Focal Loss contributes to robust handling of class imbalance, improving classification across all stages.
-Feature importance analysis indicates that EMG and the EEG Pz-Oz channel play a critical role in prediction.
-This approach not only improves predictive performance but also enhances interpretability, laying a foundation for future work in explainable and clinically applicable sleep analysis tools.

### 2025-05-08 Chapters/resources.tex
changes

\begin{figure}[h!]
	\centering
	\begin{minipage}[b]{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper3}
		\caption{Automated Sleep Staging System with EEG Signal using Machine Learning Techniques}
		\label{fig:paper3}
	\end{minipage}\hfill
	\begin{minipage}[b]{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper1}
		\caption{Deep Neural Model for Automated Sleep Staging System using Single-Channel EEG Signal}
		\label{fig:paper1}
	\end{minipage}\hfill
	\begin{minipage}[b]{0.3\linewidth}
		\centering
		\includegraphics[width=\linewidth]{img/paper2}
		\caption{SleepGCN-Transformer: A Hybrid Graph-Convolutional and Transformer Network for Sleep Stage Classification}
		\label{fig:paper2}
	\end{minipage}
\end{figure}

\section{1. Dataset}
We conducted our experiments using the \textbf{Sleep-EDF dataset}, a publicly available dataset for sleep analysis research.\\
\texttt{https://physionet.org/content/sleep-edfx/1.0.0/}

\section{2. Development Environment}
Our implementation was carried out using the following configurations:
\begin{itemize}
	\item \textbf{Programming Language}: Python 3.13.1
	\item \textbf{Deep Learning Framework}: PyTorch
	\item \textbf{IDE}: VS Code
	\item \textbf{Hardware}: Apple M1 Chip (CPU and Metal for acceleration)
\end{itemize}

\section{3. Version Control}
To manage our codebase, we used GitHub for version control.\\
QR code available alongside the figures.

\section{4. License}
The source code of this research is registered under the \textbf{Apache License 2.0}.\\
\texttt{http://www.apache.org/licenses/}

\section{5. Key Python Libraries \& Versions}
\begin{itemize}
	\item \textbf{NumPy}: 2.2.2
	\item \textbf{Pandas}: 2.2.3
	\item \textbf{Scikit-learn}: 1.6.1
	\item \textbf{SciPy}: 1.15.1
	\item \textbf{Matplotlib}: 3.10.0
	\item \textbf{Seaborn}: 0.13.2
	\item \textbf{PyTorch}: 2.6.0
	\item \textbf{Torch Geometric}: 2.6.1
	\item \textbf{NetworkX}: 3.4.2
	\item \textbf{MNE}: 1.9.0
\end{itemize}
This setup ensures a robust and reproducible environment for the development and evaluation of our sleep stage classification model.

### 2025-05-08 Title/conclusion.tex
changes

--- 2025-05-05 Title/conclusion.tex
@@
-Our proposed SleepGCN-Transformer model achieves 93.12\% training accuracy and
-93.04\% validation accuracy, demonstrating its effectiveness in sleep stage classification.
-The integration of Graph Convolution Networks (GCN) captures spatial dependencies
-across EEG, EOG, and EMG channels, while the Transformer extracts temporal patterns. The
-use of Focal Loss enhances class balancing, improving performance on underrepresented sleep
-stages. Feature importance analysis highlights EMG and EEG Pz-Oz as key predictors. This
-robust approach lays the foundation for future work in Explainable AI, enabling medical
-professionals to interpret AI-driven sleep diagnostics effectively.
+The proposed SleepGCN-Transformer architecture demonstrates superior performance in sleep stage classification, achieving 93.12\% training accuracy and 93.04\% validation accuracy.
+By integrating GCNs to model spatial interdependencies and Transformers to extract temporal patterns, the model effectively captures the complexity of sleep physiology.
+The application of Focal Loss contributes to robust handling of class imbalance, improving classification across all stages.
+Feature importance analysis indicates that EMG and the EEG Pz-Oz channel play a critical role in prediction.
+This approach not only improves predictive performance but also enhances interpretability, laying a foundation for future work in explainable and clinically applicable sleep analysis tools.

### 2025-05-08 Chapters/resources.tex
changes

@@
 
-\section{1. Dataset}
+\section{Dataset}
 We conducted our experiments using the \textbf{Sleep-EDF dataset}, a publicly available dataset for sleep analysis research.\\
@@
 
-\section{2. Development Environment}
+\section{Development Environment}
 Our implementation was carried out using the following configurations:
@@
 	\item \textbf{Programming Language}: Python 3.13.1
-	\item \textbf{Deep Learning Framework}: PyTorch
-	\item \textbf{IDE}: VS Code
+	\item \textbf{Environment Manager}: pip (with virtual environment)
+	\item \textbf{Deep Learning Frameworks}: PyTorch, TensorFlow
+	\item \textbf{IDE}: VS Code, Jupyter Notebook (.ipynb)
 	\item \textbf{Hardware}: Apple M1 Chip (CPU and Metal for acceleration)
@@
 
-\section{3. Version Control}
+\section{Version Control (.GIT)}
 To manage our codebase, we used GitHub for version control.\\
@@
 
-\section{4. License}
+\section{License}
 The source code of this research is registered under the \textbf{Apache License 2.0}.\\
@@
 
-\section{5. Key Python Libraries \& Versions}
+\section{Key Python Libraries \& Versions}
 \begin{itemize}

### 2025-05-13 Chapters/resources.tex
changes

@@
 This setup ensures a robust and reproducible environment for the development and evaluation of our sleep stage classification model.
+
+\section{published and submitted work }
+
+\begin{itemize}
+		\item \textit{Automated Sleep Staging System Using EEG Signal Feature-Based Classification by Machine Learning Techniques}, Tanmay Rathod, Santosh Kumar Satapathy, Nibedita Das, Rajesh Kumar Mohapatra, Suren Kumar Sahu, Vaishvi R Shah, 2024 IEEE 8th International Conference on Information and Communication Technology (CICT).
+\url{https://ieeexplore.ieee.org/document/10899578}
+
+	\item \textit{Automated Sleep Staging System with EEG Signal using Machine Learning Techniques}, Santosh Kumar Satapathy, Tanmay Rathod, Nibedita Das, Rajesh Kumar Mohapatra, Suren Sahu, Jaynil Joshi, 2025 IEEE International Conference on Interdisciplinary Approaches in Technology and Management for Social Innovation (IATMSI).
+\url{https://ieeexplore.ieee.org/document/10985276}
+
+	  \item \textit{SleepGCN-Transformer: A Hybrid Graph-Convolutional Approach for EEG-based Sleep Staging}, Tanmay Rathod et al., submitted to \textit{Biomedical Signal Processing and Control}, May 2025 (under review).
+
+	\end{itemize}
+
+\section{Contact Information}
+
+For inquiries or collaboration, contact the author via the following:
+
+\begin{itemize}
+	\item \textbf{Email:} \href{mailto:tanmayrathod777@gmail.com}{tanmayrathod777@gmail.com}
+	\item \textbf{GitHub:} \url{https://github.com/tanmay007thor}
+	\item \textbf{IEEE Xplore Author Profile:} \url{https://ieeexplore.ieee.org/author/577823064896395}
+	\item \textbf{Portfolio Website:} \url{https://tanmay-dev.netlify.app/}
+	\item \textbf{LinkedIn:} \url{https://www.linkedin.com/in/tanmay-rathod-15a332230}
+	\item \textbf{Phone:} +91 90165 89777
+\end{itemize}

### 2025-05-13 Title/approval.tex
changes of panel

@@
 
-This thesis entitled \textbf{\enquote{SleepGCN-Transformer: A Hybrid Graph
-Convolutional and Transformer Network for Sleep
-Stage Classification}} by \textbf{Tanmay Rathod} is recommended for the degree of \textbf{M.Tech} in \textbf{Artificial Intelligence}.
-\vspace{0.8cm}
+This thesis entitled \textbf{\enquote{SleepGCN-Transformer: A Hybrid Graph Convolutional and Transformer Network for Sleep Stage Classification}} by \textbf{Tanmay Rathod} is recommended for the degree of \textbf{M.Tech} in \textbf{Artificial Intelligence}.
 
-\hspace{9cm}  
-\begin{flushright}   
-    \makebox[1.8in]{\hrulefill}\\
-    Guide : Dr Santosh Satapathy, \\
-    Assistant Professor \\
-    Dept. of ICT \\
+\vspace{1cm}
 
-\end{flushright}
-\vspace{0.5cm}
-
-\hspace{9cm}  
-\begin{flushright}   
-    \makebox[1.8in]{\hrulefill}\\
-    Dr Nitin Singh Rajput, \\
-    Assistant Professor \\
-    Dept. of ICT
-\end{flushright}
-\vspace{0.5cm}
-
-% \hspace{9cm}\\
-\begin{flushright}
-    \makebox[1.8in]{\hrulefill}\\
-    Dr. Paawan Sharma,\\
-    Head Of Department \\ 
-    Dept. of ICT
-\end{flushright}
+\begin{table}[h!]
+	\centering
+	\begin{tabular}{@{}m{0.45\textwidth}<{\centering} m{0.45\textwidth}<{\centering}@{}}
+		\textbf{Guide} & \textbf{Panel Members} \\[4em]
+		\rule{0.4\textwidth}{0.4pt} & \rule{0.4\textwidth}{0.4pt} \\[1em]
+		Dr. Santosh Satapathy, & Member 1, \\[1em]
+		Assistant Professor, & Designation, \\[1em]
+		Dept. of ICT & Dept. of ICT \\[2em]
+		\textbf{Head of Department} & \\ [4em]
+		\rule{0.4\textwidth}{0.4pt} & \rule{0.4\textwidth}{0.4pt} \\[1em]
+		Dr. Paawan Sharma & Member 2, \\[1em]
+		HoD, Dept. of ICT, & Designation,\\[1em]
+		School of Technology & Designation\\[1em]
+	\end{tabular}
+\end{table}
 

### 2025-05-19 Title/acknowledgement.tex
changes

@@
-%\begin{center}
-%\textbf{\large Acknowledgment}
-%\end{center} 
 \section*{\textcolor{internationalkleinblue}{\textbf{Acknowledgment}}}
- While my name appears as the sole contributor to the completion of this summer internship, it is important to recognize that the guidance and support of many individuals played a significant role in its success.
-This internship is the result of a collective effort, shaped by the contributions of numerous people, to whom I am deeply grateful.
 
- I extend my heartfelt appreciation to my internal guide, Dr. Santosh Kumar Satapathy
-Author, for his consistent guidance, encouragement, and insightful suggestions throughout this internship.
-His belief in my abilities provided the confidence and motivation necessary to complete this project.
-Dr.
-Santosh Kumar Satapathy
-Author’s support, along with his generous provision of time and access to resources, was instrumental in achieving the objectives of this internship.
+Although this thesis bears only my name as the author, its successful completion is the result of the support, guidance, and encouragement of several individuals to whom I am deeply grateful.
 
- I am also profoundly thankful for the unwavering support of my parents.
-Their continuous inspiration, moral backing, and blessings have been the cornerstone of my journey.
-Their understanding and encouragement have been invaluable, and I am truly fortunate to have had their steadfast support throughout this endeavor.
+First and foremost, I would like to express my sincere gratitude to my internal guide, Dr.
+Santosh Kumar Satapathy, from Pandit Deendayal Energy University (PDEU), for his unwavering support throughout the course of this project.
+His valuable insights, constructive feedback, and continuous encouragement were instrumental in shaping the direction of this research and overcoming various challenges.
+
+I also acknowledge the faculty members and staff of the M.Tech in Artificial Intelligence program at PDEU for providing a stimulating academic environment and the foundational knowledge that enabled me to undertake this work.
+
+Finally, I extend heartfelt thanks to my family and friends for their constant encouragement, patience, and moral support throughout this journey.
+
+This thesis would not have been possible without the collective contributions of these individuals, and I remain truly thankful to all.
+
 \begin{flushright}

### 2025-05-19 Title/approval.tex
changes

@@
 	\begin{tabular}{@{}m{0.45\textwidth}<{\centering} m{0.45\textwidth}<{\centering}@{}}
-		\textbf{Guide} & \textbf{Panel Members} \\[4em]
-		\rule{0.4\textwidth}{0.4pt} & \rule{0.4\textwidth}{0.4pt} \\[1em]
-		Dr. Santosh Satapathy, & Member 1, \\[1em]
-		Assistant Professor, & Designation, \\[1em]
-		Dept. of ICT & Dept. of ICT \\[2em]
-		\textbf{Head of Department} & \\ [4em]
-		\rule{0.4\textwidth}{0.4pt} & \rule{0.4\textwidth}{0.4pt} \\[1em]
-		Dr. Paawan Sharma & Member 2, \\[1em]
-		HoD, Dept. of ICT, & Designation,\\[1em]
-		School of Technology & Designation\\[1em]
+		\textbf{Expert 1} & \textbf{Expert 2} \\[2.5em]
+		\rule{0.35\textwidth}{0.4pt} & \rule{0.35\textwidth}{0.4pt} \\[0.5em]
+		Mr. Nilesh Ranpura & Dr. Hridoy Sankar Dutta \\[0.5em]
+		Director of Engineering, & Lecturer (Assistant Professor), \\[0.5em]
+		Arrow Inc. (E-Infochips) & Deakin University \\[3em]
+	\end{tabular}
+\end{table}
+
+\begin{table}[h!]
+	\centering
+	\begin{tabular}{@{}m{0.31\textwidth}<{\centering} m{0.31\textwidth}<{\centering} m{0.31\textwidth}<{\centering}@{}}
+		\textbf{Guide} & \textbf{Head of Department} & \textbf{Director} \\[2.5em]
+		\rule{0.25\textwidth}{0.4pt} & \rule{0.25\textwidth}{0.4pt} & \rule{0.25\textwidth}{0.4pt} \\[0.5em]
+		Dr. Santosh Satapathy & Dr. Paawan Sharma & Prof.(Dr.) Dhaval Pujara \\[0.5em]
+		Assistant Professor, & HoD, Dept. of ICT, & Director, \\[0.5em]
+		Dept. of ICT & School of Technology & School of Technology \\
 	\end{tabular}
